---
format: rostools-theme-revealjs
search: false
slide-number: true
smaller: true
---

## Evaluating a crystal ball - part 1

# {.center}

The future is now

## {.center}

::: {.notes}
Session 18: Evaluating classification models for cognitive impairment
prediction.

Key dataset: ad_validation_logistic with observed classes and predicted
probabilities.
:::

# The ABCD Framework {.center}

A useful way to remember prediction-model evaluation:

- **A** = Calibration-in-the-large
- **B** = Calibration slope
- **C** = C-statistic / Discrimination
- **D** = Decision curve / Clinical usefulness

::: {.notes}
This framework separates statistical performance from clinical usefulness.

Reference: https://pmc.ncbi.nlm.nih.gov/articles/PMC4155437/
:::

## Case Study: Cognitive Impairment Prediction

We evaluate a logistic regression model predicting cognitive impairment from:

- Age
- Sex (male)
- Tau protein
- Phosphorylated tau
- Amyloid beta 42
- APOE genotype

**Outcome**: Impaired vs Control (gold-standard neurophyschological test)

## The Prediction Model

The model was trained on a separate training set:

$$
  impairment_i \sim age + male + tau + p\_tau + Ab\_42 + genotype
$$

where predictors are measured in cerebrospinal fluid via lumbar puncture.

::: {.notes}
Tau and p_tau: biomarkers related to neurofibrillary tangles Ab_42: Amyloid beta
42 forms plaques in Alzheimer's disease Genotype: APOE allele pairs (E2E2, E2E3,
E3E3, E3E4, E2E4, E4E4)
:::

## Dataset Visualization {.r-stretch}

```{r}
#| label: data-plt1
#| fig-cap: "Tau vs p-tau by genotype and diagnosis"

library(dplyr)
library(ggplot2)

ad_data <- medisr::ad_data

ad_data$diagnosis <- factor(ad_data$diagnosis,
  levels = c("Impaired", "Control"),
  labels = c("impaired", "control"),
)

ad_data$Genotype <- factor(ad_data$Genotype,
  levels = c("E3E3", "E3E4", "E4E4", "E2E3", "E2E4", "E2E2"),
  labels = c("E2E2", "E2E3", "E3E3", "E3E4", "E2E4", "E4E4")
)

test <- ad_data[seq(1, nrow(ad_data), by = 2), ]

p1 <- ggplot(test, aes(tau, p_tau, colour = diagnosis)) +
  geom_point(
    aes(alpha = factor(male)),
    size = 2.4,
    position = position_jitter(width = 0.03, height = 0.03)
  ) +
  facet_wrap(~ Genotype) +
  scale_alpha_manual(
    values = c(`0` = 0.55, `1` = 0.9),
    name = "Male"
  ) +
  labs(
    x = "Tau",
    y = "p-tau",
    colour = "Diagnosis"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

p1
```

## Biomarker Distributions {.r-stretch}

```{r}
#| label: data-plt2
#| fig-cap: "Distribution of biomarkers by diagnosis"

library(tidyr)

biomarkers <- test |>
  pivot_longer(
    cols = c(tau, p_tau, Ab_42),
    names_to = "biomarker",
    values_to = "value"
  )

p2 <- ggplot(
  biomarkers,
  aes(diagnosis, value, fill = diagnosis)
) +
  geom_violin(
    alpha = 0.25,
    colour = NA,
    trim = FALSE
  ) +
  geom_boxplot(
    width = 0.18,
    outlier.shape = NA,
    alpha = 0.65
  ) +
  geom_jitter(
    width = 0.08,
    height = 0,
    size = 1.4,
    alpha = 0.55
  ) +
  facet_wrap(~ biomarker, scales = "free_y") +
  labs(
    x = NULL,
    y = NULL,
    fill = "Diagnosis"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

p2
```

# Evaluating Prediction Models

## What We Need

A prediction model can be evaluated from two things:

- What actually happened (observed outcome)
- What the model predicted (predicted probabilities)

Here we use predictions from a pre-trained logistic regression model.

::: {.notes}
We work with ad_validation_logistic which contains: - Class: the observed
outcome - .pred_impaired: predicted probability of impairment

No need to fit a model; we simply evaluate its predictions.
:::

## The Validation Data

```{r}
#| label: get-data
#| echo: false

ad_validation_logistic <- medisr::ad_validation_logistic
```

Class = observed outcome. `.pred_impaired` = predicted probability of
impairment.

```{r}
#| label: looking-at-data
#| echo: false

head(ad_validation_logistic)
```

::: {.notes}
The dataset contains 167 observations with: - Class: factor with levels
"control" and "impaired" - .pred_impaired: numeric predicted probabilities
:::

## Confusion Matrix

How many were correctly classified?

```{r}
#| label: conf-mat-prep
#| echo: false

library(probably)

predictions <- ad_validation_logistic |>
    mutate(
        .pred_class = probably::make_two_class_pred(
            .pred_impaired,
            levels(Class),
            threshold = 0.5 )
        )
```

```{r}
#| label: conf-mat
#| fig-cap: "Confusion matrix for threshold = 0.5"

library(yardstick)

conf_mat(
  predictions,
  truth = Class,
  estimate = .pred_class
)
```

::: {.notes}
Pattern observed: - Many more control participants (123) than cognitively
impaired (44) - Only 25/44 impaired were labeled correctly (56.8%) - 109/123
controls were labeled correctly (88.6%)

The model is much better at identifying controls than impaired participants.
:::

## Accuracy: How Often Correct? {.center}

> How often does the model get the class right?

$$
  Accuracy = \frac{TP + TN}{TP + TN + FP + FN}
$$

```{r}
#| label: accuracy

accuracy(predictions, Class, .pred_class)
```

**Accuracy = 0.802** (80.2% classified correctly)

## The Problem with Accuracy {.center}

Accuracy is easy to understand but has an important limitation.

::: {.fragment}
It does not tell us **which class** the model gets wrong.
:::

::: {.fragment}
It can be misleading when classes are **unbalanced**.
:::

::: {.fragment}
Example: If 95% are controls, a model that always predicts "control" has 95%
accuracy, despite failing to identify **any** impaired participants.
:::

## Specificity: Avoiding False Positives {.center}

> Among people who are **not** "impaired", how many does the model correctly
> identify?

Specificity measures how well the model identifies the **negative class**
(controls).

$$
  Sensitivity = \frac{TP}{TP + FN}
$$

```{r}
#| label: specificity

specificity(predictions, Class, .pred_class)
```

**Specificity = 0.886** (88.6% of controls correctly identified)

::: {.notes}
A highly specific model produces few false positives: it rarely labels a control
as impaired.

Particularly useful when false positives are costly (e.g., unnecessary
treatment, concern).
:::

## Sensitivity: Avoiding False Negatives {.center}

> Among people who **are** "impaired", how many does the model correctly
> identify?

Sensitivity measures how well the model identifies the **positive class**
(impaired).

$$
  Specificity = \frac{TN}{TN + FP}
$$

```{r}
#| label: sensitivity

sensitivity(predictions, Class, .pred_class)
```

**Sensitivity = 0.568** (56.8% of impaired correctly identified)

::: {.notes}
A highly sensitive model produces few false negatives: it rarely misses an
impaired participant.

Our model has high specificity (good at recognizing controls) but lower
sensitivity (misses substantial proportion of impaired).

This distinction is crucial for understanding model performance.
:::

## Sensitivity vs Specificity {.center .r-fit-text}

::: {.columns}
::::: {.column}
### Specificity = 88.6%

- Good at identifying controls
- Few false positives
- Conservative about calling impairment

::::::: {.column}
### Sensitivity = 56.8%

- Misses many impaired cases
- More false negatives
- Not aggressive enough
:::::::
:::::
:::

## Discussion Task {.center}

**Discuss with your neighbor:**

- Which diagnosis would you prefer a **lower** probability threshold? Why?
- Which diagnosis would you prefer a **higher** probability threshold? Why?

Consider consequences: - Cost of treatment - Side-effects - Effectiveness of
treatment

::: {.notes}
Sensitivity and specificity depend on the chosen probability threshold.

Lower threshold: more willing to classify as impaired → higher sensitivity,
lower specificity Higher threshold: more conservative → higher specificity,
lower sensitivity

The "right" threshold depends on the costs of false positives vs false
negatives.
:::

# Discrimination

## ROC and AUC

Instead of choosing one threshold, we can look at model discrimination across
**all** possible thresholds.

The ROC curve shows sensitivity against 1 − specificity.

```{r}
#| label: roc
#| fig-cap: "ROC curve for cognitive impairment prediction"

roc_curve(ad_validation_logistic, Class, .pred_impaired, event_level = "first" ) |>
    autoplot()
```

## Area Under the Curve

```{r}
#| label: auc

roc_auc(ad_validation_logistic, Class, .pred_impaired)
```

**AUC** = area under the ROC curve

::: {.notes}
AUC can be interpreted as the probability that the model assigns a higher
predicted probability to a randomly chosen "impaired" observation than to a
randomly chosen "control" observation.

AUC measures **discrimination** (how well the model distinguishes between
classes). It does NOT tell us whether the predicted probabilities themselves are
accurate.
:::

# Calibration

## Are the Probabilities Accurate? {.center}

Calibration asks a different question:

> When the model says 70%, does the event happen about 70% of the time?

A simple calibration plot compares predicted probability with the observed
proportion.

## Calibration Plot {.r-stretch}

```{r}
#| label: calibration
#| fig-cap: "Calibration plot with breaks"

library(probably)

cal_plot_breaks(ad_validation_logistic, Class, .pred_impaired)
```

A well-calibrated model follows the diagonal line.

## Smooth Calibration Curve {.r-stretch}

```{r}
#| label: cal-smooth
#| fig-cap: "Smooth calibration curve"

cal_plot_logistic(ad_validation_logistic, Class, .pred_impaired, smooth = FALSE)
```

::: {.notes}
Calibration is about the **accuracy of the probabilities**, not just the
rankings.

A model can have excellent discrimination (high AUC) and poor calibration
(probabilities don't match observed frequencies).

The calibration model can be written as a logistic regression of the outcome on
the logit of the predicted probability.
:::

# ABCD Framework in Detail

## A: Calibration-in-the-large

Are predictions systematically too high or too low?

## B: Calibration Slope

Are predictions too extreme or not extreme enough?

## C: C-statistic

How well does the model discriminate between outcomes? (AUC)

## D: Decision Curve

Does using the model improve clinical decisions?

```{r}
#| label: cal-model
#| echo: false

calibration_model <- ad_validation_logistic |>
    mutate(
        outcome = Class == "impaired",
        logit_p = log(.pred_impaired / (1 - .pred_impaired))
    ) |>
        glm(outcome ~ logit_p, data = _, family = binomial)

broom::tidy(calibration_model)
```

::: {.notes}
The intercept is related to calibration-in-the-large (A) and the coefficient of
logit_p is the calibration slope (B).

Ideally: calibration-in-the-large ≈ 0 and calibration slope ≈ 1.

This is called the C-statistic of the AUC.
:::

# Prediction vs Effect Estimation

## Different Questions {.center}

::: {.columns}
::::: {.column}
### Prediction

> How accurately can we predict what will happen?

- Focus: accuracy of predictions
- Tools: accuracy, sensitivity, specificity, AUC, calibration
:::::
::::: {.column}
### Effect Estimation

> How does changing a variable relate to what happens?

- Focus: causal associations
- Tools: regression coefficients, odds ratios
:::::
:::

::: {.notes}
Important distinction:

A model can be excellent for prediction without giving an unbiased estimate of a
causal effect.

The dataset ad_validation_logistic contains only predictions, not the original
predictors, so we cannot recover the original regression coefficients.

What we can estimate is the association between the model's predicted risk and
the observed outcome (the calibration slope is one such measure).
:::

## Key Idea

> No single metric evaluates a prediction model completely.

A useful model should: - Discriminate well - Produce sensible probabilities -
Lead to better decisions

# Summary: Evaluation Toolkit {.smaller}

  | Question                                             | Measure         |
  | ---------------------------------------------------- | --------------- |
  | Does the model classify correctly at this threshold? | Accuracy        |
  | Does it avoid false positives?                       | Specificity     |
  | Does it avoid false negatives?                       | Sensitivity     |
  | Can it distinguish the two outcomes?                 | ROC / AUC       |
  | Are the probabilities accurate?                      | Calibration     |
  | Does it improve clinical decisions?                  | Decision curve  |
  | What is the effect of a predictor?                   | Effect estimate |

## {.center}

::: {.callout-important appearance="simple" icon=false}
The key idea: **No single metric tells the whole story.**

A useful model discriminates well, produces accurate probabilities, and improves
decisions.
:::
