# Data wrangling using R

<!-- TODO: Add an introduction here -->

## Learning objectives

The learning objectives for this session are:

{{< include /includes/objectives/_session2.qmd >}}

## TODO: Section

::: callout-note
## Reading task: \~4 minutes

TODO: Reading text
:::

## :technologist: Exercise: Title

**Time: ~NUM minutes.**

Text

{{< text_snippet sticky_up >}}

## :speech_balloon: Discussion activity: Title

**Time: ~Num minutes.**

Text

## :book: Reading task: Title

**Time: ~Num minutes.**

Text

{{< text_snippet sticky_up >}}

## Summary

-   TODO: List of summary items

## Survey

{{< include /includes/_survey.qmd >}}








## :book: Reading task: Working with messy and tidy data

::: {.callout-note collapse="true"}
## Instructor note

For the reading section, emphasize the characteristics of a "tidy"
dataset.
:::

**Time: \~12 minutes**

The concept of "tidy" data was popularized in an
[article](https://www.jstatsoft.org/v59/i10/paper) [@Wickham2014] by
Hadley Wickham and described in more detail in the [Tidy Data
chapter](https://r4ds.hadley.nz/data-tidy.html) of the *R for Data
Science* online book. Before we continue with tidy data, we need to
cover something that is related to the concept of "tidy" and that has
already come up in this workshop: the `{tidyverse}`. The `{tidyverse}`
is an ecosystem of R packages that are designed to work well together,
that all follow a strong "[design
philosophy](https://design.tidyverse.org/)" and common [style
guide](https://style.tidyverse.org/). This makes combining these
packages in the tidyverse much easier. We teach the tidyverse because of
these reasons.

Part of the "tidy" part of tidyverse revolves around tidy data. A tidy
dataset is when:

-   Each variable has its own column (e.g. "Body Weight").
-   Each observation has its own row (e.g. "Person").
-   Each value has its own cell (e.g. "Body weight for a person at a
    specific date").

Take a look at the example "tidy" and "messy" data frames (also called
"tibbles" in the tidyverse) below. Think about why each may be
considered "tidy" or "messy". What do you notice between the tidy
versions and the messier versions?

```{r}
library(tidyr)
```

```{r}
# Datasets come from tidyr
# Tidy:
table1

# Partly tidy:
table2

# Messier:
table3

# Messy:
table4a

# Messy:
table4b
```

The "most" tidy version is `table1` as each column describes their
values (e.g. population is population size), each row is unique (e.g.
first row is for values from Afghanistan from 1999), and each cell is an
explicit value representative of its column and row.

`table2` is a "long" version of `table1` so it is partly "tidy", but it
doesn't satisfy the rule that each variable has a column, since `count`
represents both cases and population size.

On the other hand, `table3` is messy because the `rate` column values
are a composite of two other column values (cases and population), when
it should be a single number (a percent). Both `table4a` and `table4b`
have columns with ambiguous values inside. For example, you can't tell
from the data what the values in the `1999` column contain.

Tidy data has a few notable benefits:

1.  Time spent preparing your data to be tidy from the beginning can
    save days of added work and frustration in the long run.
2.  "Tidy data" is a conceptual framework that allows you to easily
    build off and wrangle (i.e. "manipulate", "clean up", "manage") data
    in simpler and easy-to-interpret ways, especially when used within
    the framework of the tidyverse.
3.  Tidier data is easier to visualize and "map" onto different axes of
    the plot, like the x and y axes.

The concept of tidy data also gives rise to "tidy code" for visualizing
and wrangling. By using "verbs" (R functions) and chaining them together
in "sentences" (in a sequential pipeline), you can construct meaningful
and readable code that describes in plainer English what you are doing
to the data. This is one simple way that you can enhance the
reproducibility of your code. There are other ways to make your code
tidier and more readable, all of which we will cover in this workshop.

-   Document what you did to your data and why you did it, to help you
    remember later on by using comments with `#` hashes as well as using
    Markdown text to describe things that can't be easily explained by
    the code, for example, the "why" behind the code.
-   Keep the code simple: Don't be clever nor overly concise, always be
    clear even if it means more code. Clear code is easier to understand
    than clever and sometimes overly-complex or -concise code.

Whether working with either messy or tidy data, there are a few
principles to follow:

-   You should always save your original raw dataset in the `data-raw/`
    folder.
    -   *Note*: Whether or not you save data to `data-raw/` depends on
        how you collected the data and how many collaborators are on
        your team. You may end up storing and processing the data in
        another folder as a project of its own.
-   **Never** edit your raw data directly. If you have to edit it to fix
    mistakes, have some version control of it so you can track it and
    see what has been changed.
-   Only work with your raw data using R code. *As much as possible or
    is reasonable, don't manually edit it*. Manual editing doesn't leave
    a history of what you've done to it, so you can't go back and see
    what you've done. Always keep a history of any changes you've made
    to the data, preferably by using R code.
-   Save the edited data as another dataset and store it in the `data/`
    folder.

We are saving to `data/` because the dataset is already collected and
published, so the original raw data can always be downloaded again.

{{< text_snippet hats_up >}}

## :speech_balloon: Discussion activity: How tidy is our data?

**Time: \~10**

Now that we've learned about the concept of "tidy" data, let's discuss
how tidy our data is. At a glance, there are approximately 7 things in
this data that are mildly untidy or could be improved. Can you identify
them?

1.  On your own, take 2 minutes to look over the data and try to find
    what could be improved.
2.  With your neighbour, take 5 minutes and discuss what you found and
    see if you have overlap in agreement or if you found different
    things.
3.  As a group, we will take 3 minutes and we will share what you all
    discussed and then go over the 7 things.

::: {.callout-note appearance="minimal" collapse="true"}
## :teacher: Instructor note

The 6 main things are:

1.  Column names are not tidy or standardised. For instance, there are
    `.` in the names and some use abbreviations.
2.  Some column names also contain data. For instance, all the columns
    with numbers at the end are time points, like `Insulin1` is insulin
    measured at 1 minute after eating the meal. This data is in the wide
    format, when it should be in the long format. We'll cover long and
    wide in @sec-wrangling.
3.  Some columns that contain a number at the end, likely indicating
    time, also contain `..` instead of `.`. This likely means it was
    minus the number (`-5`, 5 minutes before the test). So it should be
    renamed to something like `_minus5`.
4.  The data could actually be split into two or three datasets, one in
    the long format for the measurements taken at regular minute
    intervals, one for the measurements taken at 30 minute intervals
    (the OGTT).
5.  There are two ID columns, `id` and `OFS.ID`.
6.  The `Group` column has abbreviations as `FDR` and `CTR` instead of
    the full names.
7.  There are two similar columns for insulin from the OGTT:
    `Insulin.0.OGTT.X` and `Insulin.0.OGTT`.
:::

## Summary

-   Use `read_csv()` or `read_csv2()` to import data from a `.csv`
    (comma or tab separated) file.
-   Use `here::here()` to make file paths easier to manage in your R
    project.
-   Use `glimpse()` to get an overview of the data.
-   Use the Quarto code chunk options like `echo`, `message`, or
    `warning` to control what is shown in the output of the code chunk.
-   Use tab auto-completion when writing code.
-   With tidy data, each variable has its own column, each observation
    has its own row, and each value has its own cell.
-   Never edit raw data. Instead, use R code to make changes and clean
    up the raw data, rather than manually editing the dataset.

## Survey

Please complete the survey for this session:
