# A Poissonous counting trick {#sec-poisson}

> "Choose to be optimistic, it feels better." --- Dalai Lama

> "Choose to model positive outcomes, it feels better for count outcomes." ---
> Daniel Skak Mazhari-Jensen

```{r}
#| label: setup
#| include: false

#library(tidyverse)
#ddagit782594dalibrary(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)
library(tidyr)
library(broom)
library(marginaleffects)

hd_data <- readr::read_csv(
  "https://raw.githubusercontent.com/DanMazJen/medicinsk-statistik/main/data/hd_data.csv"
)
```

## Pree-class Literature

- [Today's dataset](s)
- [Compendium](s)
- [GLM](s)

# Slides

> [**Session 12 slides**](../slides/day12.html)

<div>

```{=html}
<iframe class="slide-deck" src="../slides/day12.html"></iframe>
```

</div>

::: {.callout-note collapse="true"}
## Teacher note

The slides contain speaking notes that you can view by pressing 'S' on the
keyboard.
:::

## Case session 12: ""

## Review of Logistic regression

See session 9 on logistic regression by Carsten.

- We use the binomial distribution with logit link to model probabilities
- These are discrete counts, but we are interested in the overall probability of
  getting outcome 0/1
- Only two outcomes: 0 or 1 (heads or tails, dead or alive, left or right)
- The logit link transform our outcome variable to a linear space called
  log-odds
- The logistic function transform out data back to the outcome space
  (probabilities)

## Reading Task: When the odds are low, go Poisson!

While the binomail distribution is great when *N* is known (trial count), there
exists studies where you don't know the upper limit (counting fish, selling a
product, etc.) --- what is the upper bound of fish in a lake we haven't fully
accounted for?

> If we don't know *N*, we cannot use the binomial!

Luckily, there's a trick for this. When N becomes huge and *p* becomes really
small, the binomial distribution takes a special shape --- the Poisson
distribution.

Explained mathematically:

The expected value of a binomial distribution is just *Np* (probability scale),
and its variance $Np(1-p)$. But when *N* is very large and *p* is very small,
the variance and expected value is the same.

E.g., you have 1 in 1000 probability of getting a price.

```{r}
#| label: binom-to-poisson

y <- rbinom(1e5, 1000, 1/1000)
c( mean(y), var(y))
```

As a result, the distribution only has one parameter, λ.

$$
  y_i \sim Poisson(λ)
$$

This solves several issues:

- We can now model counts and probabilities (actually, *rates*, but more on this
  later) without being in control of the experiment and knowing the finite trial
  count
- We can model events with extremely low probability in counts per sample
  instead of probability per trial
- We only need one parameter for both the mean and the variance, effectively
  saving a degree of freedom.

But how can we do this in practice?

### Which link?

As a link function, we often use log.

$$
  y_i \sim Poisson(λ_i) \\
  log(λ_i) = α + β(x_i - \bar{x})
$$
Why log?

```{r}
#| label: why-log

# Poisson regression:
# log(mu) = beta0 + beta1*x

dat <- data.frame(
  x = seq(-5, 5, length.out = 200)
) %>%
  mutate(
    log_mu = -1 + 0.5*x,       # can be negative
    mu = exp(log_mu),           # always positive
  ) %>%
  pivot_longer(
    cols = c(log_mu, mu),
    names_to = "scale",
    values_to = "value"
  ) %>%
  mutate(
    scale = recode(
      scale,
      log_mu = "Log-linear (Poisson) scale:\nlog(E(Y))",
      mu = "Outcome scale:\nE(Y)"
    )
  )

ggplot(dat, aes(x = x, y = value)) +
  geom_line(
    linewidth = 1.3,
    color = "#2166AC"
  ) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "grey40"
  ) +
  facet_wrap(~scale, scales = "free_y") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Why Use a Log Link in Poisson Regression?",
    subtitle = "Negative values are allowed on the log scale, but exponentiation guarantees positive counts",
    x = "Predictor (x)",
    y = "Outcome (y)"
  ) #+
 # annotate(
 #   "text",
 #   x = -2.5,
 #   y = -2.8,
 #   label = "log(μ) can be negative\n(no restriction)",
 #   size = 5,
 #   color = "black"
 # ) +
 # annotate(
 #   "text",
 #   x = 2,
 #   y = 0.25,
 #   label = "μ = exp(log(μ))\n> 0 always",
 #   size = 5,
 #   color = "black"
 # ) +
 # annotate(
 #   "segment",
 #   x = -1,
 #   xend = 1,
 #   y = -1.5,
 #   yend = 0.2,
 #   arrow = arrow(length = unit(0.2, "cm")),
 #   color = "firebrick"
 # ) +
 # annotate(
 #   "segment",
 #   x = 0,
 #   xend = 1,
 #   y = 0.4,
 #   yend = 1.5,
 #   arrow = arrow(length = unit(0.2, "cm")),
 #   color = "firebrick"
 # )
```

Note that the log-link has the same flaws and benefits from the binomial
distribution.

Pros:

- ensures positive values (for prediction)

Cons:

- implies exponential relationship with predictors
  - Grow very rapidly --- therefore, very sensitive to range
  - it cannot keep being exponential!

## Data Example

Epilepsy seizure count or survival of graduate students in different faculty?

Or maybe hispital length of stay data?

$$
  Seizures_i \sim Poisson(λ_i) \\
  log(λ_i) = α * β_1 predrugseizures + β_2 drug  \\
$$

Heres a DAG:
<!-- panache-ignore-format-start -->
          Age
           │
           ▼
Baseline seizures ─────► Outcome (Seizures)
        │                  ▲
        │                  │
        ▼                  │
   Treatment ──────────────┘
        ▲
        │
 Unmeasured severity
<!-- panache-ignore-format-end -->
```{r}
#| label: dag

dagitty::dagitty("
dag {

Age -> base

Age -> y

base -> y

trt -> y

severity -> base

severity -> y

}
") |>
    ggdag::tidy_dagitty() |>
    ggdag::ggdag()
```

Let's explore the dataset a bit --- but before we do, I'll do some tidying of
the dataset using `{dplyr}` for wrangling. I'm not gonna explain this bit, as
you can go back to session 8 by Rasmus for working with `{dplyr}` for wrangling.

```{r}
#| label: pre-wrangle

data(epil)

epil <- epil |>
    rename(
        seizure_count = y,
        drug = trt,
        baseline_seizures = base,
        trial_end = V4,
        log_base = lbase,
        log_age = lage
  ) |>
    #filter(period < 2 | period > 3) |>
    filter(period > 3) |>
    mutate(
        seizure_count = as.integer(seizure_count),
        period = factor(period,
            levels = c(4),
            labels = c("week_4")
            )
    )

```

```{r}
#| label: just-looking

epil |>
    tibble::as.tibble()

epil |>
    count(drug)
```

```{r}
#| label: explore1

library(patchwork)
library(MASS)

lims <- range(c(epil$baseline_seizures, epil$seizure_count))

baseline_data_plot <- epil |>
    ggplot() +
    geom_histogram(aes(x=baseline_seizures)) +
  coord_cartesian(xlim = lims)

posttrial_data_plot <- epil |>
    ggplot() +
    geom_histogram(aes(x=seizure_count)) +
  coord_cartesian(xlim = lims)

baseline_data_plot + posttrial_data_plot
```

```{r}
#| label: test

epil |>
    ggplot(aes(x=y)) +
    geom_histogram() +
    facet_grid(~ V4)

fit <- glm(y ~ trt * base, V4, # + lage
    family = poisson,
    data = epil
)

predict(fit) |>
    data.frame(
        mu_hat = _,
        mu_obs = epil$y
    ) |>
    ggplot(aes(x=exp(mu_hat))) +
    geom_histogram()

predict(fit, type = "response") == exp(predict(fit))

epil |>
  mutate(
    predicted = predict(fit, type = "response")
  ) |>
  ggplot(aes(predicted, y)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, linetype = 2, colour = "red") +
  labs(
    x = "Predicted count",
    y = "Observed count"
  ) +
  theme_bw()
```

```{r}
#| label: exploration

epil |>
    count(trt)

epil |>
    ggplot(aes(seizure_count)) +
    geom_histogram(binwidth = 2,
                   fill="steelblue",
                   color="white") +
    theme_bw()

epil |>
    ggplot(aes(baseline_seizures,seizure_count,color=drug)) +
    geom_point(alpha=0.6) +
    geom_smooth(method="loess",se=FALSE) +
    theme_bw()
```

```{r}
#| label: model-fit

m1 <- glm(
    seizure_count ~ drug * baseline_seizures * age,
    family = poisson(link="log"),
    data = epil
)

summary(m1)
```

Yeah, we did it! But... What does the coefficients actually mean? Did we come
any closer to answering our medical question: did the drug reduce epilepsy?

## GLMs are notoriously difficult to interpret

- like a classification problem of predicting tides --- many small-world
  continuous processes that build up when all stars align to give either GO or
  NO-GO!

Binomial regression

- counts of two possible with known *n* (outcome often in per cent)
- dead/alive, heads/tails, accept/reject, left/right,

Poisson regression

- counts of unknown max value (no theoretical upper limit) aka with unknown n
  (outcome in counts)
- elephants in Kenya, applicants for a grant, reported *p*-values in a paper,
  etc.

```{r}
#| label: inference

broom::tidy(m1,
     exponentiate = TRUE,
     conf.int = TRUE)
```

## Reporting odds-ratios or counts

From an absolute perspective, the musquito is the most dangerous animal in the
world! It causes most deaths a year. However, if I were to battle a musquito or
a shark, I would bet my life on the musquito and not the shark! It is relatively
much more dangerous! However, conditional on being in the real-life situation of
facing a musquito or a shark, the former is much more probable.

```{r}
#| label: predictions

epil$pred <- predict(
    m1,
    type="response"
)

ggplot(epil,
       aes(pred,seizure_count))+
    geom_point()+
    geom_abline(slope=1,
                intercept=0,
                color="red")+
    theme_bw()
```

```{r}
#| label: counterfactuals

library(marginaleffects)

m1 |>
    plot_predictions(
    condition="baseline_seizures"
    ) +
    labs(
    y="Predicted seizures"
)

cf <- bind_rows(
#
    predictions(
    m1,
    newdata=transform(epil,
    drug="placebo")) |>
    mutate(group="Placebo"),
#
    predictions(
    m1,
    newdata=transform(epil,
    drug="progabide")) %>%
    mutate(group="Treatment")
)

cf |>
    ggplot(aes(estimate,fill=group))+
    geom_density(alpha=0.4)+
    theme_bw()

cmp <- comparisons(
    m1,
    variables="drug"
)

# Each bar is the estimated reduction in seizures if that patient switched treatments.
cmp |>
    ggplot(aes(estimate))+
    geom_histogram(
    bins=30,
    fill="steelblue")+
    theme_bw()


m2 <- glm(y ~ trt * base + age + V4,
    family=poisson,
    data=epil
)

plot_predictions(
    m2,
    condition=c("base","trt")
)
```

```{r}
#| label: cut-off-of-when-to-use

placebo <- predictions(
    m1,
    newdata = transform(epil, drug = "placebo")
)

treated <- predictions(
    m1,
    newdata = transform(epil, drug = "progabide")
)

epil$placebo <- placebo$estimate
epil$treated  <- treated$estimate

epil$benefit <- epil$placebo - epil$treated

epil |>
    filter(benefit >= 3)

grid <- datagrid(
    model=m1,
    baseline_seizures = seq(0, 100, by = 1),
    age = mean(epil$age), #28 yo
    #V4 = 1, # follow-up
    drug = c("placebo", "progabide")
)

pred <- marginaleffects::predictions(m1, newdata = grid)

cutoff <- pred |>
    select(baseline_seizures, drug, estimate) |>
    tidyr::pivot_wider(
        names_from = drug,
        values_from = estimate) |>
    mutate(
        reduction = placebo - progabide
    )

ggplot(cutoff,
       aes(baseline_seizures, reduction)) +
    geom_line(size = 1.2) +
    geom_hline(yintercept = -3,
               linetype = 2,
               colour = "red") +
    labs(
        x = "Baseline seizures",
        y = "Expected reduction",
        title = "Expected treatment benefit"
    ) +
    theme_bw() +
    geom_label(
        data = cutoff_pt,
        aes(
            label = paste0(
            "Baseline = ", baseline_seizures,
            "\nReduction = ", round(reduction, 2))
        ),
  nudge_y = -30
)

cutoff_pt <- cutoff |>
  slice_min(abs(reduction - (-3)), n = 1)

```

manual predictions:

```{r}
#| label: manual-predict

x_seq <- seq(min(epil$base), max(epil$base), length.out = 100)

b <- coef(fit)

mu_hat <- exp(
  b[1] +
  b["base"] * x_seq +
  b["trtprogabide"] * 0       # reference treatment
  #b["base"] * mean(epil$lbase)
)



plot(x_seq, mu_hat, type = "l")

# placebo
newdata <- data.frame(
  base = x_seq,
  trt = "placebo"
)
newdata$mu_hat <- predict(
  fit,
  newdata = newdata,
  type = "response"
)

# progabide
newdata2 <- data.frame(
    base = x_seq,
    trt = "progabide"
)
predict(fit, newdata2, type = "response")

mu_prog <- exp(
    b["(Intercept)"] +
    b["trtprogabide"] +
    (b["base"] + b["trtprogabide:base"]) * x_seq
)
data.frame(x=x_seq,y=mu_prog) |>
    ggplot(aes(x=x,y=y)) +
    geom_line()

#X <- model.matrix(~ trt + base + `trtprogabide:base`, newdata)
#X <- model.matrix(~ trt * base, newdata)
tt <- delete.response(terms(fit))

X <- model.matrix(
  tt,
  newdata,
  xlev = fit$xlevels
)

newdata$mu_hat <- exp(X %*% coef(fit))
```

This last expression,

`exp(X %*% coef(fit))`

is essentially what predict.glm(type = "response") computes internally. It's the
most general way to think about GLM predictions: construct the design matrix X,
multiply by the coefficient vector β to get the linear predictor η=Xβ, then
apply the inverse link function. For a Poisson model with the log link, the
inverse link is simply exp().

The predicted values in a generalized linear model are obtained using **matrix
multiplication** of the design matrix and the estimated coefficient vector.

Let the design matrix be

$$
  X =
  \begin{bmatrix}
  1 & 0 & 6.0 & 0\\
  1 & 0 & 7.46 & 0\\
  \vdots & \vdots & \vdots & \vdots\\
  1 & 0 & 151 & 0
  \end{bmatrix},
$$

and let the estimated regression coefficients be

$$
  \hat{\beta} =
  \begin{bmatrix}
  1.29584\\
  -0.24923\\
  0.02145\\
  0.00042
  \end{bmatrix}.
$$

The linear predictor is obtained by matrix multiplication:

$$
  \eta = X\hat{\beta}.
$$

In R, this is computed as

```r
eta <- X %*% coef(fit)
```

Here,

- $X$ is a $100 \times 4$ design matrix,
- $\hat{\beta}$ is a $4 \times 1$ coefficient vector,
- $\eta$ is a $100 \times 1$ vector of linear predictors.

For the first observation, the linear predictor is computed as the product of
the first row of $X$ and the coefficient vector:

$$
  \eta_1
  =
  \begin{bmatrix}
  1 & 0 & 6 & 0
  \end{bmatrix}
  \begin{bmatrix}
  1.29584\\
  -0.24923\\
  0.02145\\
  0.00042
  \end{bmatrix}
  =
  1(1.29584)
  +0(-0.24923)
  +6(0.02145)
  +0(0.00042)
  =
  1.42453.
$$

This row-by-column multiplication is an **inner product**, also called a **dot
product**, between the first row of the design matrix and the coefficient
vector.

Thus,

- The overall computation,
  $$
  \eta = X\hat{\beta},
  $$
  is **matrix multiplication**.

- Each individual fitted value,
  $$
  \eta_i = x_i^\top \hat{\beta},
  $$
  is an **inner product (dot product)**.

Since this is a Poisson generalized linear model with a log link, the predicted
mean response is obtained by applying the inverse link function:

$$
  \hat{\mu} = \exp(\eta).
$$

In R:

```r
eta <- X %*% coef(fit)   # matrix multiplication
mu  <- exp(eta)          # inverse log link
```

A **cross product** is **not** being computed here. In vector algebra, the cross
product is defined only for three-dimensional vectors and produces another
vector. Although R has a function named `crossprod()`, it computes

$$
  X^\top Y,
$$

which is a matrix product, not the geometric cross product.

### Exposure and the offset

The parameter λ is the expected value of a Poisson model --- but also commonly
interpreted as a *rate*.

The rate interpretation is nice for allowing varying expoure time.

Implizitly, λ is equal to an expected number of events, μ, per unit time or
distance, τ. This implies that λ = μ/τ, which lets us redefine the link:

$$
  y_i \sim Poisson(λ_i) \\
  log λ_i = log \frac{μ_i}{τ_i} = α + βx_i
$$

Since the logarithm of a ratio is the same as the difference of logarithms:

$$
  log λ_i = log μ_i - log τ_i = α + βx_i
$$

These τ are the "exposures". If different observations *i* have different
exposures, the expected value on row *i* is given by

$$
  log μ_i = log τ_i + α + βx_i
$$

When τ_i = 1, then logτ_i = 0 and we're back to standard Poisson!

But when the exposure is not 1, τ does effectively *scales* the expected number
of events for each case *i* in the unit x = 1 (1 day, year, whatever...).
Therefore, we write:

$$
  y_i \sim Poisson(μ_i) \\
  log μ_i = log τ_i + α + βx_i
$$

Where τ is a column in the data. It's just like adding a predictor (taking its
log) without adding a parameter for it!

If you put a parameter in front of τ, you're able to model the hypothesis that
the rate is not constant with time.

*Remember:* When the length of observations, area of sampling, or intensity of
sampling varies, the counts we observe also naturally vary! Poisson assumes
stationarity (time/space) --- here, we add an offset! That is --- the logarithm
of the exposure unit.

Think of days of monitoring a patient...

```{r}
#| label: offset-example

set.seed(636961739)
num_days <- 30 # days of the experiment

hospital1 <- rpois (num_days, 1.5) # 1.5 events per day

num_weeks <- 4 # weeks of the experiment

hospital2 <- rpois(num_weeks, 0.5*7) # events * lowst unit (i.e., days)

events <- c(hospital1, hospital2)
exposure <- c(rep(1,30), rep(7,4))
hospital_id <- c(rep(0,30), rep(1,4))
hospital_data <- data.frame(
  events=events,
  days=exposure,
  hospital_id=hospital_id
)

hospital_data$log_days = log(hospital_data$days)

sim_m1 <- glm( events ~ 1 + hospital_id,
    family = poisson,
    offset = log_days,
    data = hospital_data
)

broom::tidy(sim_m1,exponentiate = TRUE)
# We simulated 1.5 for hospital1 (intercept) and 0.5 (a third or 33%) for hospital2. So OR of .31
```

### Negative binomial (gamma-Poisson) models

Poisson models often have a lot of unexplained variance, mainly from ignoring
*rate heterogeneity*.

The gamma-Poisson is a mixture of different Poisson distributions (similar to
how Student-t is a mixture of Gaussians).

```{r}
#| label: assumptions

performance::check_overdispersion(m1)

performance::check_model(m1)

fit <- MASS::glm.nb(y ~ trt * base, V4,
    data = epil
)

performance::check_model(fit)

fit <- glm(y ~ trt * base, V4,
    family = poisson,
    data = epil
)
```

# Other dataset

https://search.r-project.org/CRAN/refmans/dobson/html/doctors.html?utm_source=chatgpt.com
https://search.r-project.org/CRAN/refmans/ACSWR/html/bs1.html?utm_source=chatgpt.com
https://github.com/agbarnett/dobson/tree/main/data

```{r}
url <- "https://raw.githubusercontent.com/agbarnett/dobson/main/data/doctors.rda"
load(url(url))
doctors

glm(
  deaths ~ age + smoking,
  family = poisson(),
  offset = log(personyears),
  data = doctors
)

doctors2 <- doctors |>
  mutate(
    age_group = factor(
      age,
      levels = 1:5,
      labels = c("35–44", "45–54", "55–64", "65–74", "75–84")
    ),
    rate = deaths / personyears,
    rate_1000 = 1000 * rate,
    offsets = log(personyears)
  ) |>
  select(age_group, smoking, deaths, personyears, rate, rate_1000, offsets)

doctors2
```

```{r}
ggplot(doctors2,
       aes(age_group, rate_1000,
           colour = smoking,
           group = smoking)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  labs(
    x = "Age group",
    y = "Deaths per 1,000 person-years",
    colour = "Smoking"
  ) +
  theme_bw()

ggplot(doctors2,
       aes(personyears, deaths,
           colour = smoking)) +
  geom_point(size = 3) +
  scale_x_log10() +
  theme_bw()

ggplot(doctors2,
       aes(age_group,
           log(rate),
           colour = smoking,
           group = smoking)) +
  geom_point(size = 3) +
  geom_line() +
  labs(y = "Log mortality rate")
```

```{r}
m1 <- glm(
  deaths ~ offset(offsets) + smoking + age_group,
  #offset = offsets,
  family = poisson(),
  data = doctors2
)

summary(m1)
```

```{r}
avg_predictions(
    m1,
    variables = "smoking",
    type = "response"
)

cf <- bind_rows(

    mutate(doctors2,
           smoking = "smoker",
           scenario = "Everyone smokes"),

    mutate(doctors2,
           smoking = "non-smoker",
           scenario = "Nobody smokes")

)

pred <- predictions(
    m1,
    newdata = datagrid(
        age_group = levels(doctors2$age_group),
        smoking = levels(doctors2$smoking),
        personyears = 1000
    )
)

ggplot(pred,
       aes(age_group,
           estimate,
           colour = smoking,
           group = smoking)) +
    geom_line() +
    geom_point(size = 3) +
    ylab("Predicted deaths per 1,000 person-years")

avg_comparisons(
    m1,
    variables = "smoking",
    comparison = "ratio"
)
```

## Exploratory analysis and modelling strategy

The outcome is the number of graduates who survived, while the total number of
graduates represents the population at risk. The analysis therefore focuses on
modelling survival *rates* rather than raw counts by including the log of the
total number of graduates as an offset in Poisson regression. This estimates
multiplicative effects on the survival rate while accounting for differences in
cohort size.

The exploratory analysis first examines temporal patterns, differences between
faculties, and differences by sex to assess whether systematic trends or effect
modification are plausible. Two Poisson models are then compared: a main-effects
model and a model allowing the temporal trend to differ by faculty. Because
Poisson models assume equality of the mean and variance, overdispersion is
assessed before interpreting results. Finally, marginal effects summarize
adjusted contrasts, while counterfactual predictions visualize model-implied
survival rates under hypothetical scenarios across years and demographic groups.

```{r}
# https://github.com/agbarnett/dobson/blob/main/data/chronic.rda
url <- "https://raw.githubusercontent.com/agbarnett/dobson/main/data/graduates.rda"
load(url(url))
doctors

#load("/Users/ze18rp/Downloads/chronic.rda")
#chronic

load("/Users/ze18rp/Downloads/graduates.rda")
graduates

library(tidyverse)
library(performance)
library(marginaleffects)

graduates2 <- graduates |>
  mutate(
    survive = as.integer(survive),
    total = as.integer(total),
    year_c = year - min(year),
    faculty = as.factor(faculty)
  )

## -------------------------------------------------
## EDA
## -------------------------------------------------

# 1. Survival rate over time
ggplot(graduates2,
       aes(year, survive / total,
           colour = faculty,
           group = interaction(faculty, sex),
           shape = sex)
        ) +
  geom_line(alpha = 0.7) +
  geom_point() +
  labs(
    y = "Survival rate",
    x = "Year"
  ) +
  theme_bw()

# 2. Distribution of survival rates by faculty and sex
ggplot(graduates2,
       aes(faculty, survive / total, fill = sex)) +
  geom_boxplot() +
  labs(y = "Survival rate") +
  theme_bw()

# 3. Cohort sizes over time
ggplot(graduates2,
       aes(year, total,
           colour = faculty,
           shape = sex)) +
  geom_point(size = 2) +
  geom_line(aes(group = interaction(faculty, sex))) +
  labs(y = "Total graduates") +
  theme_bw()

## -------------------------------------------------
## Poisson GLMs
## -------------------------------------------------

m1 <- glm(
  survive ~ year_c + faculty + sex + offset(log(total)),
  family = poisson,
  data = graduates2
)

m2 <- glm(
  survive ~ year_c * faculty + sex + offset(log(total)),
  family = poisson,
  data = graduates2
)

summary(m1)
summary(m2)

# Check Poisson assumption
performance::check_overdispersion(m1)
performance::check_overdispersion(m2)

## -------------------------------------------------
## Marginal effects
## -------------------------------------------------

# Average adjusted comparison between faculties
avg_comparisons(m2, variables = "faculty")
avg_comparisons(
  m2,
  variables = list(faculty = "pairwise")
)
nd <- datagrid(
  year_c = mean(graduates2$year_c),
  sex = levels(factor(graduates2$sex)),
  faculty = levels(graduates2$faculty),
  total = 100
)
avg_comparisons(
  m2,
  variables = list(faculty = "pairwise"),
  newdata = datagrid(
    year_c = mean(graduates2$year_c),
    sex = unique,
    faculty = unique,
    total = 100
  )
)

# Average slope for year
avg_slopes(m2, variables = "year_c")

## -------------------------------------------------
## Counterfactual predictions
## -------------------------------------------------

# Predicted survival rate over time by faculty
plot_predictions(
  m2,
  condition = c("year_c", "faculty"),
  newdata = datagrid(
    year_c = seq(min(graduates2$year_c),
                 max(graduates2$year_c),
                 length.out = 100),
    faculty = unique(graduates2$faculty),
    sex = "men",
    total = 100
  ),
  type = "response"
)

# Predicted survival rate by year and sex
plot_predictions(
  m2,
  condition = c("year_c", "sex"),
  newdata = datagrid(
    year_c = seq(min(graduates2$year_c),
                 max(graduates2$year_c),
                 length.out = 100),
    faculty = "medicine",
    sex = unique(graduates2$sex),
    total = 100
  ),
  type = "response"
)
```
