# Marginalizing the problem {#sec-marginal-effects}

> "x." --- John Doe

```{r}
#| label: setup
#| include: false

#library(tidyverse)
#ddagit782594dalibrary(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)

#hd_data <- readr::read_csv(
#  "https://raw.githubusercontent.com/DanMazJen/medicinsk-statistik/main/data/hd_data.csv"
#)
data(hd_data, package = "medisr")
```

## Pree-class Literature

{{< include /includes/curriculum/_session15.qmd >}}

## Slides

> [**Session 15 slides**](../slides/day15.html)

<div>

```{=html}
<iframe class="slide-deck" src="../slides/day15.html"></iframe>
```

</div>

::: {.callout-note collapse="true"}
## Teacher note

The slides contain speaking notes that you can view by pressing 'S' on the
keyboard.
:::

## Case session 15: "If you're already dead, your odds of dying don't increase."

At a PhaRmaceutic RStudio AS, a new compound is being evaluated for toxicity
before it can move further through development. In an early preclinical
experiment, groups of laboratory animals are exposed to increasing doses of the
compound, and investigators record how many animals die at each dose.

At the weekly safety meeting, the lead researcher summarizes the results:

> "As the dose increases, mortality increases."

One colleague looks at the highest-dose groups and asks:

> "But if almost all the animals are already dead, can another increase in dose
> really have the same effect?"

The team looks more closely at the data. Mortality increases rapidly through the
middle of the dose range, but begins to level off at high doses.

The investigators fit a logistic regression model because the outcome is binary:
each animal either died or survived. The model produces a coefficient for dose.

> "A one-unit increase in dose increases mortality by 8.7," the lead researcher
> says.

The biostatistician stops them.

> "The coefficient is on the log-odds scale. That's not a change in
> probability."

The team now faces a more useful question:

- How does increasing dose change the probability of death?

And because the dose-response relationship is nonlinear, another question
follows:

- At what dose are we asking about the effect?

Your task is to investigate the dose-response relationship using a quadratic
logistic regression model. You will move from regression coefficients to
predicted probabilities and marginal slopes, examining how the effect of dose
changes across the dose range.

## Reading Task: Communicating results of regression models

*find something from marginaleffects* ...

## The danger of coefficients

::: {.callout-warning}
In nonlinear models, you cannot generally interpret the regression coefficient
as the change in probability.
:::

::: {.callout-important}
The purpose of `{marginaleffects}` is essentially to help us make this
translation: From a fitted generalized model → to quantities that answer the
scientific question.
:::

## Marginalizing the problem - our workflow:

Throughout this session, we'll use this workflow:

1. What is my scientific question --- **the estimand**?
2. Choose machinery to model data --- **the estimator**?
3. What did the regression estimate --- **the coefficient**?
4. What quantity do I actually want to communicate?
5. At what values?

- A particular patient?
- A clinically meaningful scenario?
- The average population?
- Different levels of an interacting variable?

6. Post-processing model coefficients --- **the estimate**
7. Communicate in clinically meaningful units

::: {#fig-marginal-effects-workflow .column-page fig-cap="From the scientific question to a clinically interpretable marginal effect."}
```{mermaid}
%%| label: workflow-diag

flowchart LR
    A["1. Scientific question<br/><br/>What happens to P(Y)<br/>when X changes?"]
    --> B["2. Regression estimate<br/><br/>log-odds"]

    B --> C["3. Quantity to communicate<br/><br/>Probability"]

    C --> D{"4. Who/where are we estimating the effect?"}

    D --> D1["Particular patient"]
    D --> D2["Clinically meaningful<br/>scenario"]
    D --> D3["Average population"]
    D --> D4["Different levels of<br/>an interacting variable"]

    D1 --> E["5. Marginal effects"]
    D2 --> E
    D3 --> E
    D4 --> E

    E --> E1["avg_slopes(model)"]
    E --> E2["slopes(model, ...)"]
    E --> E3["avg_comparisons(model, ...)"]

    E1 --> F["6. Communicate in<br/>clinically meaningful units"]
    E2 --> F
    E3 --> F

    F --> G["1.8 percentage-point increase<br/>in predicted probability"]

    style A fill:#E8F1FB,stroke:#3B6EA5,stroke-width:2px
    style B fill:#F3F3F3,stroke:#666,stroke-width:2px
    style C fill:#E8F5E9,stroke:#4C8C4A,stroke-width:2px
    style D fill:#FFF4D6,stroke:#C79A24,stroke-width:2px
    style E fill:#EDE7F6,stroke:#7656A6,stroke-width:2px
    style F fill:#E8F5E9,stroke:#4C8C4A,stroke-width:2px
    style G fill:#DDF3E0,stroke:#3D8B40,stroke-width:3px
```
:::

As shown in @fig-marginal-effects-workflow, we move from the **scientific
question**, through the **model scale**, to a **marginal effect on the
probability scale**.

## Reading Task: Investigating one variable at a time

...

> How does `[RISK FACTOR]` affect the probability of `[DISEASE/OUTCOME]`?

### Context

- Population: `[POPULATION]`
- Risk factor: `[X]`
- Outcome: `[Y]`
- Relevant clinical context: `[CONTEXT]`
- Dataset/source: `[DATA SOURCE]`

> **If a patient's `[RISK FACTOR]` increases by `[X UNITS]`, how much does their
> probability of disease increase?**

## The estimand

We are interested in the probability of death:

$$
  P(death = 1 | X)
$$

for an RCT study looking at dose-effect of death on beetles:

```{r}
#| label: smoking-dag
#| echo: false

dag <- ggdag::dagify(
  died ~ dose
)

ggdag::ggdag(dag) +
  ggdag::geom_dag_point(size = 20) +
  ggdag::geom_dag_text(size = 3) +
  ggdag::theme_dag()
```

Here, there are no confounding. We randomized the study and controlled all
variables.

For smoking cigarettes, using observational data:

```{r}
#| label: beetles-dag
#| echo: false

dag <- ggdag::dagify(
  deaths ~ smoking + age,
  smoking ~ age
)

ggdag::ggdag(dag) +
  ggdag::geom_dag_point(size = 20) +
  ggdag::geom_dag_text(size = 3) +
  ggdag::theme_dag()
```

Here, age is confounding the effect of smoking on death (specifically, it's a
fork). Therefore, we need to adjust for age.

## The estimator

$$
  y_i = Binomial(n,p_i) \\
  logit(p_i) = α + βx_i
$$
where *n* is the number of observations and *p* is the probability of death.

The logit function is defined as the *log-odds*:

$$
  logit(p_i) = log(\frac{p_i}{1 - p_i})
$$

The "odds" of an event is just the probability it happens divided by the
probability it does not happen. In other words:

$$
  log(\frac{p_i}{1 - p_i}) = α + βx_i
$$
To figure out the definition of *p_i*, we do some algebra to solve for $p_i$:

$$
  p_i = \frac{exp( α + βx_i)}{1 + exp(α + βx_i)}
$$
The above function is the **logistic**

- Note that the log-odds space is linear (what we model)
- The logistic space is the sigmoidal shape (the transformation)

### The statistical models

```{r}
#| label: doctors-model

age_labels <- c("35 to 44 years", "45 to 54 years", "55 to 64 years", "65 to 74 years", "75 to 84 years")
doctors <- medisr::doctors |>
    mutate(age = factor(age, levels=c(1:5), labels=age_labels)) |>
    select(-agesq)

smoking_model <- glm(
  deaths ~ smoking * age + offset(log(personyears)),
  family = poisson(link = "log"),
  data = doctors
)
```

If you did the exercises from @sec-ex-poisson you are very familiar with this
dataset and model. Therefore, I won't explain it in detail. Here, we use the
offset to compute a fixed rate per year (see @sec-poisson).

Here's the model output:

```{r}
#| label: smoking-output

smoking_model |>
    parameters::parameters(exponentiate=TRUE)
```

- coefficient for `[X]`
- odds ratio for `[X]`
- statistical uncertainty, if relevant

**Logistic example**

We also make a logistic regression:

```{r}
#| label: beetles-data

beetle_dose <- medisr::beetle_dose |>
    mutate(
        survived = n-died#,
        #dose_sq = I(dose^2)
        )

beetles_model <- glm( cbind(died, survived) ~ dose, family = binomial(link =
"logit"), data = beetle_dose )
```

Binomial data can be stored as **aggregated** and **disagreggated**. The former
is compact and show the total success and failures:

<!--panache-ignore-start-->
| Coin    |  Heads    | Tails |
| -------- | ------- | ------- |
| Small    | 49      | 56      |
| Medium   | 51      | 55      |
| Large    | 50      | 47      |
<!--panache-ignore-end-->

The disaggregated dataset would show a long list with each coinflip coming out
as either `heads = 1` or `heads = 0` so the first 5 tosses could be:

<!--panache-ignore-start-->
| Coin    | Heads |
|---------|-------|
| Small   | 1     |
| Small   | 1     |
| Small   | 0     |
| Small   | 1     |
| Small   | 0     |
<!--panache-ignore-end-->

Because our dataset is *aggregated*, we need to tell `glm()` what's the success
and failure --- in this case, how many died and how many did not from the dose
concentration. We use `cbind(died, survived)` as the outcome. Effetively this is
a way to say that 6 out of 59 --- or ~10% of the beetles died of the lowest does
of 49.05689 mg/L of gaseous carbon disulphide (CS₂) --- values are in log_10 CS₂
mg/L.

```{r}
#| label: beetle-output

beetles_model |>
    parameters::parameters(exponentiate=TRUE)
```

## Clear communication

Suppose:

OR = 1.50

What does this mean?

A 1-unit increase in age is associated with a 50% increase in the odds of
cardiovascular disease, holding the other variables constant.

But the more important question is:

Does this mean that a 1-year increase in age increases the probability of
disease by 50 percentage points?

No.

An odds ratio describes a change in odds, not directly a change in probability.

### Why the odds ratio is not enough

The clinical question is usually about risk/probability, not odds.

Consider two patients:

- Patient A has a predicted disease probability of 5%.
- Patient B has a predicted disease probability of 50%.

Suppose both experience the same change in the predictor.

The resulting change in probability will not necessarily be the same.

This is the key complication of logistic regression:

> The effect of X on probability depends on where you start.

The relationship between X and probability is nonlinear.

The logistic curve is relatively flat near 0 and 1, and steepest in the middle.

Therefore:

> There is not necessarily one single "effect of X" on probability.

## Step 4: What quantity do I actually want to communicate

In this case, we're interested in probability of death for the beetles and count
of deaths per personyear for the smoking dataset.

## Step 5: At what values?

- A particular patient?
- A clinically meaningful scenario?
- The average population?
- Different levels of an interacting variable?

## Step 6: Post-processing model coefficients --- **the estimate**

```{r}
#| label: outcome-scale
#|
library(marginaleffects)

plot_predictions(beetles_model, condition =  "dose")

slopes(
  beetles_model,
  variables = "dose",
  type = "response"
)

plot_slopes(
  beetles_model,
  variables = "dose",
  condition = "dose",
  type = "response"
)


plot_comparisons(
  beetles_model,
  variables = "dose",
  condition = list("dose" = 1:2),
  type = "response"
)



smoking_model

plot_predictions(smoking_model, by="age")

slopes(
  smoking_model,
  variables = "smoking",
  type = "response"
)

plot_slopes(
  smoking_model,
  variables = "smoking",
  condition = "age",
  type = "response"
)

plot_comparisons(
  smoking_model,
  variables = "smoking",
  condition = "age",
  newdata = doctors,
  type = "response"
)

my_grid <- expand.grid(
  age = levels(doctors$age),
  smoking = levels(doctors$smoking),
  personyears = 10000
)

plot_comparisons(
  smoking_model,
  variables = "smoking",
  condition = "age",
  newdata = my_grid,
  type = "response"
)

plot_comparisons(
  smoking_model,
  variables = "smoking",
  condition = "age",
  newdata = my_grid,
  comparison = "ratio",
  type = "response"
)


marginaleffects::avg_predictions(
  smoking_model,
  variables = "smoking",
  by = "smoking"
)

marginaleffects::avg_predictions(
  smoking_model,
  variables = "smoking",
  by = "age"
)


ATE = marginaleffects::avg_comparisons(
  smoking_model,
  variables = "smoking",
  newdata = doctors
)
ATE

ATT <- marginaleffects::avg_comparisons(
  smoking_model,
  variables = "smoking",
  newdata = subset(smoking == 'smoker')
)
ATT

ATU <- marginaleffects::avg_comparisons(
  smoking_model,
  variables = "smoking",
  newdata = subset(smoking == 'non-smoker')
)
ATU

CATE <- marginaleffects::avg_comparisons(
  smoking_model,
  variables = "smoking",
  by = "age"
)
CATE

```

### What is a marginal effect?

A marginal effect asks:

*How much does predicted probability change when X changes?*

For a continuous variable:

*What happens to predicted probability if X increases by 1 unit?*

For example:

If age increases by 1 year, how much does predicted probability of disease
change?

In a nonlinear model, this is calculated on the probability scale, rather than
the log-odds scale.

With {marginaleffects}:

```{r}
#| label: marg-eff
#| eval: false

library(marginaleffects)

avg_slopes(model)
```

Conceptually, the output answers: A 1-unit increase in age → Δ predicted
probability

If the result were:

age: 0.012

the interpretation would be:

A one-year increase in age is associated with a 1.2 percentage-point increase in
predicted probability, on average.

Notice the distinction:

0.012 probability

1.2 percentage points

### The effect depends on the patient

Suppose we want to know the effect of age for a particular patient.

We might ask:

What is the marginal effect of age for a 50-year-old non-smoker with normal
blood pressure?

versus:

What is the marginal effect of age for a 75-year-old smoker with high blood
pressure?

These can be different.

We can calculate effects for specified covariate values:

```{r}
#| label: grid-eff
#| eval: false

slopes(
  model,
  newdata = datagrid(
    age = c(50, 75),
    smoking = c(0, 1),
    blood_pressure = c(120, 160)
  )
)
```

The important concept is not the syntax.

It is:

A marginal effect is conditional on the characteristics of the person for whom
we calculate it.

## Predicted probabilities versus marginal effects

These are different questions.

Predicted probability What is this patient's predicted probability of disease?

```{r}
#| label: pred
#| eval: false

predictions(
  model,
  newdata = datagrid(
    age = 60,
    smoking = 0,
    blood_pressure = 130
  )
)
```

Marginal effect How much does the predicted probability change if age increases?

```{r}
#| label: pro-marg
#| eval: false

slopes(
  model,
  newdata = datagrid(
    age = 60,
    smoking = 0,
    blood_pressure = 130
  )
)
```

**Predictions:**

> "What is the risk?"

**Marginal effects:**

> "How much does the risk change?"

Both are useful, but they answer different scientific questions.

## Average marginal effects

Often we do not want the effect for one particular patient.

We want:

What is the average effect in this population?

That is the average marginal effect (AME).

```{r}
#| label: avg-slopes
#| eval: false

avg_slopes(model)
```

Conceptually:

| Patient 1: effect of X = ...
| Patient 2: effect of X = ...
| Patient 3: effect of X = ...
|        ⋮
| Patient N: effect of X = ...

Average = AME

The interpretation is:

On average, a one-unit increase in X changes predicted probability by Δ
percentage points.

This is often much easier to communicate than an odds ratio.

## Interactions: when the effect depends on something else

Suppose our model includes:

Age × Smoking

The scientific question becomes:

Does the effect of age on disease probability differ between smokers and
non-smokers?

The model might be:

```{r}
#| label: model1
#| eval: false

model <- glm(
  disease ~ age * smoking + blood_pressure,
  data = data,
  family = binomial
)
```

The temptation is to look at the interaction coefficient and interpret it
directly.

Don't.

The interaction coefficient is on the log-odds scale.

Instead, calculate the marginal effect of age at different smoking statuses:

```{r}
#| label: int-model1
#| eval: false

slopes(
  model,
  variables = "age",
  by = "smoking"
)
```

Now the result directly addresses the scientific question:

Effect of age among non-smokers Effect of age among smokers

And you can ask:

Is the difference between those marginal effects clinically meaningful?

## Visualizing the interaction

```{r}
#| label: plotting1
#| eval: false

plot_predictions(
  model,
  condition = c("age", "smoking")
)
```

## Comparing clinically meaningful changes

A one-unit change is not always clinically meaningful.

For example, instead of asking:

What is the effect of a 1-unit increase in blood pressure?

we might ask:

What is the change in predicted probability when blood pressure increases from
120 to 140 mmHg?

For this, use a comparison:

```{r}
#| label: clin-inform
#| eval: false

avg_comparisons(
  model,
  variables = list(
    blood_pressure = c(120, 140)
  )
)
```

This produces a quantity directly corresponding to the clinical question:

Predicted probability at 140 − Predicted probability at 120 = change in
probability

## Uncertainty A marginal effect is an estimate, not a known quantity.

Therefore we also report uncertainty:

- Estimate
- 95% CI
- p-value

For example:

AME = 0.018 (95% CI 0.010--0.026)

Interpretation:

On average, a one-unit increase in X is associated with a 1.8 percentage-point
increase in predicted probability, with a 95% confidence interval from 1.0 to
2.6 percentage points.

::: {.hidden}
Reading task: Reporting incidence rate ratio or counts

> Should we report the incidence rate ratio or the counts for each group?

What would you prefer and why?

From an absolute perspective, the musquito is the most dangerous animal in the
world! It causes most deaths a year. However, if I were to battle a musquito or
a shark, I would bet my life on the musquito and not the shark! It is relatively
much more dangerous! However, conditional on being in the real-life situation of
facing a musquito or a shark, the former is much more probable.

Now, consider the interpretation of the coefficients and model we just made in
@sec-poisson-interp. How would you inform a patient or a clinican about the
results?

<details>
 <summary>**Click for a tip**</summary>

*Tip*: consider conditioning on `baseline_seizures`.

Is `round((1 - exp(m1_intercept$coefficients[2])) * 100, 1)`% lower the same
when you have 1 seizure every 8 weeks?
What about 100? Would it be a meaningful unit to make national policies?

Now, what if we take the average absolute change:
`(1-exp(m1_intercept$coefficients[2]))*mean(subset(epil,drug=="placebo")$seizure_count)`
when you have 1 seizure every 8 weeks? What about 100? Would it be a meaningful
unit to make national policies?.

 </details>
:::

::: {.hidden}
Extra madness - counterfactuals for clinical inference

This is an extra part (not curriculum material).

Suppose you wanted to know a cutoff for when the drug prevents *at least* three
seizures:

```r
#| label: cut-off-of-when-to-use

grid <- datagrid(
    model=m2_baseline,
    baseline_seizures = seq(0, 100, by = 1),
    age = mean(epil$age), #28 yo
    #V4 = 1, # follow-up
    drug = c("placebo", "progabide")
)

pred <- marginaleffects::predictions(m2_baseline, newdata = grid)

cutoff <- pred |>
    dplyr::select(baseline_seizures, drug, estimate) |>
    tidyr::pivot_wider(
        names_from = drug,
        values_from = estimate) |>
    mutate(
        reduction = placebo - progabide
    )

cutoff_pt <- cutoff |>
  slice_min(abs(reduction - (3)), n = 1)

worsening <- cutoff |>
  slice_min(abs(reduction - (0)), n = 1)

ggplot(cutoff,
       aes(baseline_seizures, reduction)) +
    geom_line(linewidth = 1) +
    geom_hline(yintercept = 3,
               linetype = 2,
               colour = "red") +
    labs(
        x = "Baseline seizures",
        y = "Expected reduction",
        title = "Expected treatment benefit"
    ) +
    theme_bw() +
    geom_label(
        data = cutoff_pt,
        aes(
            label = paste0(
            "Baseline = ", baseline_seizures,
            "\nReduction = ", round(reduction, 2))
        ),
        nudge_y = -1.5
    ) +
    geom_point(aes(baseline_seizures,reduction),
        cutoff_pt,
        colour = "steelblue",
        size = 3)
```

Ask the teacher if you're curious...
:::

<!--

### Reading Task: Conditional venom?

Read [this Australian Geographics
blog](https://www.australiangeographic.com.au/blogs/creatura-with-bec-crew/2025/09/funnel-web-spiders-are-australias-deadliest-not-for-our-pets/)
on the Funnel-web spider.



-->
