# Marginalizing the problem {#sec-marginal-effects}

> "x." --- John Doe

```{r}
#| label: setup
#| include: false

#library(tidyverse)
#ddagit782594dalibrary(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)

#hd_data <- readr::read_csv(
#  "https://raw.githubusercontent.com/DanMazJen/medicinsk-statistik/main/data/hd_data.csv"
#)
data(hd_data, package = "medisr")
```

## Pree-class Literature

{{< include /includes/curriculum/_session15.qmd >}}

# Slides

> [**Session 15 slides**](../slides/day15.html)

<div>

```{=html}
<iframe class="slide-deck" src="../slides/day15.html"></iframe>
```

</div>

::: {.callout-note collapse="true"}
## Teacher note

The slides contain speaking notes that you can view by pressing 'S' on the
keyboard.
:::

## Case session 14: ""

```{r}
hd_data <- hd_data |>
    mutate(
        ExAng = factor(ExAng,
            levels = c(0, 1),
            labels = c("no", "yes")
        ),
        HD = factor(HD,
            levels = c("No", "Yes"),
            labels = c("no", "yes")
        ),
        ChestPain = factor(ChestPain),
        Thal = factor(Thal),
        fasting_blood_glucose = factor(Fbs,
            levels = c(0, 1),
            labels = c("normal", "high")
        )
    )

hd_data |>
  ggplot(ggplot2::aes(Age, MaxHR, color = HD)) +
  geom_point() +
  geom_smooth(method = "lm")
```

## Reading Task: Investigating one variable at a time

...

> How does `[RISK FACTOR]` affect the probability of `[DISEASE/OUTCOME]`?

### Context

- Population: `[POPULATION]`
- Risk factor: `[X]`
- Outcome: `[Y]`
- Relevant clinical context: `[CONTEXT]`
- Dataset/source: `[DATA SOURCE]`

> **If a patient's `[RISK FACTOR]` increases by `[X UNITS]`, how much does their
> probability of disease increase?**

## The scientific model

We model the probability of disease:

```text
P(Disease = 1 | X)
```

using logistic regression:

logit(P(Y=1))=β0+β1X1+β2X2+⋯\text{logit}(P(Y=1)) = \beta\_0 + \beta\_1X_1 +
\beta\_2X_2 + \cdots

### The statistical model

```r
model <- glm(
  disease ~ risk_factor + covariate_1 + covariate_2,
  data = data,
  family = binomial
)
```

```text
[INSERT MODEL OUTPUT]
```

- coefficient for `[X]`
- odds ratio for `[X]`
- statistical uncertainty, if relevant

## Clear communication

Suppose:

> **OR = [INSERT OR]**

> What does this mean?

The more important question:

> **Does this mean that a [X]-unit increase in `[X]` increases the probability
> of disease by [OR-related percentage] percentage points?**

recall...

An odds ratio describes a change in **odds**, not directly a change in
**probability**.
