# Evaluating a crystal ball - part 1 {#sec-classification}

> The future is now

```{r}
#| label: setup
#| include: false

#library(tidyverse)
#ddagit782594dalibrary(tidyverse)
library(dplyr)
library(ggplot2)
library(yardstick)
library(probably)

#hd_data <- readr::read_csv(
#  "https://raw.githubusercontent.com/DanMazJen/medicinsk-statistik/main/data/hd_data.csv"
#)
data(hd_data, package = "medisr")
```

## Pree-class Literature

- [Marginal effects chatpter 5
  ](https://marginaleffects.com/chapters/predictions.html)
  - read

# Slides

> [**Session 15 slides**](../slides/day15.html)

<div>

```{=html}
<iframe class="slide-deck" src="../slides/day15.html"></iframe>
```

</div>

::: {.callout-note collapse="true"}
## Teacher note

The slides contain speaking notes that you can view by pressing 'S' on the
keyboard.
:::

## Case session 19: ""

## Content (still being worked on)

The established ABCD framework is

- A = calibration-in-the-large,
- B = calibration slope,
- C = discrimination/C-statistic,
- D = clinical usefulness/decision-curve analysis.

[see framework here](https://pmc.ncbi.nlm.nih.gov/articles/PMC4155437/)

The `segment_logistic` data are particularly convenient because they already
contain observed classes and out-of-sample predicted probabilities from a
logistic regression model.

### Evaluating a Prediction Model Evaluating a Prediction Model

A prediction model can be evaluated from two things:

- what actually happened;
- and what the model predicted.

Here we use predictions from a logistic regression model. The model has already
been trained, so we do not fit a model. We simply evaluate its predictions.

```{r}
#| label: get-data

data(segment_logistic)
```

Class is the observed outcome. `.pred_good` is the model's predicted probability
that the outcome is "good".

```{r}
#| label: looking-at-data

segment_logistic
```

### Accuracy

Accuracy answers a simple question:

> How often does the model get the class right?

To calculate accuracy, we first turn probabilities into classes. Here we use 0.5
as the threshold.

```{r}
#| label: accuracy

predictions <- segment_logistic |>
    mutate(
        .pred_class = probably::make_two_class_pred(
            .pred_good,
            levels(Class),
            threshold = 0.5 )
        )

accuracy(predictions, Class, .pred_class)
```

Accuracy is easy to understand, but it depends on the threshold and can be
misleading when the classes are unbalanced.

### Specificity

Specificity asks:

Among people who are not "good", how many does the model correctly identify?

```{r}
#| label: specificity

specificity(predictions, Class, .pred_class)
```

A highly specific model produces few false positives.

### Sensitivity

Sensitivity asks:

Among people who are "good", how many does the model correctly identify?

```{r}
#| label: sensitivity

sensitivity(predictions, Class, .pred_class)
```

A highly sensitive model produces few false negatives.

Sensitivity and specificity depend on the chosen probability threshold. There is
nothing special about 0.5 in clinical practice: the useful threshold depends on
the consequences of making each type of error.

### ROC and AUC

Instead of choosing one threshold, we can look at model discrimination across
all possible thresholds.

The ROC curve shows sensitivity against 1 − specificity.

```{r}
#| label: roc

roc_curve( segment_logistic, Class, .pred_good, event_level = "first" ) |>
    autoplot()
```

The area under this curve is the AUC.

```{r}
#| label: auc

roc_auc( segment_logistic, Class, .pred_good)
```

AUC can be interpreted as the probability that the model assigns a higher
predicted probability to a randomly chosen "good" observation than to a randomly
chosen non-"good" observation.

AUC measures discrimination. It does not tell us whether the predicted
probabilities themselves are accurate.

### Calibration

Calibration asks a different question:

When the model says 70%, does the event happen about 70% of the time?

A simple calibration plot compares predicted probability with the observed
proportion.

The probably package provides this directly:

```{r}
#| label: calibration

cal_plot_breaks( segment_logistic, Class, .pred_good )
```

A well-calibrated model follows the diagonal line.

We can also fit a smooth calibration curve:

```{r}
#| label: cal-smooth

cal_plot_logistic( segment_logistic, Class, .pred_good, smooth = FALSE )
```

Calibration is about the accuracy of the probabilities. A model can have
excellent discrimination and poor calibration.

### ABCD of model validation

A useful way to remember prediction-model evaluation is the ABCD framework:

- A --- Calibration-in-the-large: are predictions systematically too high or too
  low?
- B --- Calibration slope: are predictions too extreme or not extreme enough?
- C --- C-statistic: how well does the model discriminate between outcomes?
- D --- Decision curve: does using the model improve clinical decisions?

[This framework](https://pmc.ncbi.nlm.nih.gov/articles/PMC4155437/) separates
statistical performance from clinical usefulness.

For example, the calibration model can be written very simply as:

```{r}
#| label: cal-model

calibration_model <- segment_logistic |>
    mutate(
        outcome = Class == "good",
        logit_p = log(.pred_good / (1 - .pred_good))
    ) |>
        glm(outcome ~ logit_p, data = _, family = binomial)

broom::tidy(calibration_model)
```

The intercept is related to calibration-in-the-large (A) and the coefficient of
`logit_p` is the calibration slope (B).

Ideally, we'd see that the calibration-in-the-large is (close to) 0 and that the
calibration slope is (close to) 1.

This is called the **C-statistic** of the AUC.

**D**

The D component asks whether the predictions are useful for making decisions.
That requires specifying a clinical action and the range of probability
thresholds at which that action would be taken.

### Effect size estimation

Prediction performance and effect estimation are different questions.

A logistic regression coefficient describes the association between a predictor
and the outcome. For example:

How much does the odds of "good" change for a one-unit increase in age?

The dataset used here contains only the predictions, not the original
predictors. Therefore, we cannot recover the original regression coefficients or
their odds ratios from `segment_logistic`.

What we can estimate is the association between the model's predicted risk and
the observed outcome. The calibration slope above is one such measure.

This distinction is important:

Prediction asks: How accurately can we predict what will happen?

Effect estimation asks: How does changing a particular variable relate to what
happens?

A model can be excellent for prediction without giving an unbiased estimate of a
causal effect.

### Putting it all together

A simple evaluation therefore looks like this:

<!-- panache-ignore-format-start -->
|Question |  Measure|
|---------|---------|
|Does the model classify correctly at this threshold?	|Accuracy|
|Does it avoid false positives?	                        |Specificity|
|Does it avoid false negatives?	                        |Sensitivity|
|Can it distinguish the two outcomes?	                |ROC / AUC|
|Are the probabilities accurate?	                    |Calibration|
|Does it improve clinical decisions?	                |Decision curve|
|What is the effect of a predictor?	                    |Effect estimate|
<!-- panache-ignore-format-end -->

The key idea is that no single metric evaluates a prediction model completely.

A useful model should discriminate well, produce sensible probabilities, and ---
most importantly --- lead to better decisions.
