# Marginalizing the problem {#sec-marginal-effects}

> "x." --- John Doe

```{r}
#| label: setup
#| include: false

#library(tidyverse)
#ddagit782594dalibrary(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)

#hd_data <- readr::read_csv(
#  "https://raw.githubusercontent.com/DanMazJen/medicinsk-statistik/main/data/hd_data.csv"
#)
data(hd_data, package = "medisr")
```

## Pree-class Literature

{{< include /includes/curriculum/_session15.qmd >}}

# Slides

> [**Session 15 slides**](../slides/day15.html)

<div>

```{=html}
<iframe class="slide-deck" src="../slides/day15.html"></iframe>
```

</div>

::: {.callout-note collapse="true"}
## Teacher note

The slides contain speaking notes that you can view by pressing 'S' on the
keyboard.
:::

## Case session 14: ""

```{r}
hd_data <- hd_data |>
    mutate(
        ExAng = factor(ExAng,
            levels = c(0, 1),
            labels = c("no", "yes")
        ),
        HD = factor(HD,
            levels = c("No", "Yes"),
            labels = c("no", "yes")
        ),
        ChestPain = factor(ChestPain),
        Thal = factor(Thal),
        fasting_blood_glucose = factor(Fbs,
            levels = c(0, 1),
            labels = c("normal", "high")
        )
    )

hd_data |>
  ggplot(ggplot2::aes(Age, MaxHR, color = HD)) +
  geom_point() +
  geom_smooth(method = "lm")
```

### Reading Task: Conditional venom?

Read [this Australian Geographics
blog](https://www.australiangeographic.com.au/blogs/creatura-with-bec-crew/2025/09/funnel-web-spiders-are-australias-deadliest-not-for-our-pets/)
on the Funnel-web spider.

## The danger of coefficients

::: {.callout-warning}
In nonlinear models, you cannot generally interpret the regression coefficient
as the change in probability.
:::

::: {.callout-important}
The purpose of `{marginaleffects}` is essentially to help us make this
translation: From a fitted generalized model → to quantities that answer the
scientific question.
:::

## Marginalizing the problem - our workflow:

Throughout this session, we'll use this workflow:

**Step 1** --- What is my scientific question?

What happens to the probability of Y when X changes?

**Step 2** --- What did the regression estimate?

log-odds

**Step 3** --- What quantity do I actually want to communicate?

probability

**Step 4** --- At what values?

- A particular patient?
- A clinically meaningful scenario?
- The average population?
- Different levels of an interacting variable?

**Step 5** --- Use marginal effects

avg_slopes(model)

or

slopes(model, ...)

or

avg_comparisons(model, ...)

**Step 6** --- Communicate in clinically meaningful units

Prefer:

1.8 percentage-point increase in predicted probability

over:

AME = 0.018

::: {#fig-marginal-effects-workflow .column-page fig-cap="From the scientific question to a clinically interpretable marginal effect."}
```{mermaid}
%%| label: workflow-diag

flowchart LR
    A["1. Scientific question<br/><br/>What happens to P(Y)<br/>when X changes?"]
    --> B["2. Regression estimate<br/><br/>log-odds"]

    B --> C["3. Quantity to communicate<br/><br/>Probability"]

    C --> D{"4. Who/where are we estimating the effect?"}

    D --> D1["Particular patient"]
    D --> D2["Clinically meaningful<br/>scenario"]
    D --> D3["Average population"]
    D --> D4["Different levels of<br/>an interacting variable"]

    D1 --> E["5. Marginal effects"]
    D2 --> E
    D3 --> E
    D4 --> E

    E --> E1["avg_slopes(model)"]
    E --> E2["slopes(model, ...)"]
    E --> E3["avg_comparisons(model, ...)"]

    E1 --> F["6. Communicate in<br/>clinically meaningful units"]
    E2 --> F
    E3 --> F

    F --> G["1.8 percentage-point increase<br/>in predicted probability"]

    style A fill:#E8F1FB,stroke:#3B6EA5,stroke-width:2px
    style B fill:#F3F3F3,stroke:#666,stroke-width:2px
    style C fill:#E8F5E9,stroke:#4C8C4A,stroke-width:2px
    style D fill:#FFF4D6,stroke:#C79A24,stroke-width:2px
    style E fill:#EDE7F6,stroke:#7656A6,stroke-width:2px
    style F fill:#E8F5E9,stroke:#4C8C4A,stroke-width:2px
    style G fill:#DDF3E0,stroke:#3D8B40,stroke-width:3px
```
:::

As shown in @fig-marginal-effects-workflow, we move from the **scientific
question**, through the **model scale**, to a **marginal effect on the
probability scale**.

## Reading Task: Investigating one variable at a time

...

> How does `[RISK FACTOR]` affect the probability of `[DISEASE/OUTCOME]`?

### Context

- Population: `[POPULATION]`
- Risk factor: `[X]`
- Outcome: `[Y]`
- Relevant clinical context: `[CONTEXT]`
- Dataset/source: `[DATA SOURCE]`

> **If a patient's `[RISK FACTOR]` increases by `[X UNITS]`, how much does their
> probability of disease increase?**

## The scientific model

We model the probability of disease:

```text
P(Disease = 1 | X)
```

using logistic regression:

$$
  {logit}(P(Y = 1)) = \beta\_0 + \beta\_1X_1 + \beta\_2X_2 + \cdots
$$

### The statistical model

```r
model <- glm(
  disease ~ risk_factor + covariate_1 + covariate_2,
  data = data,
  family = binomial
)
```

```text
[INSERT MODEL OUTPUT]
```

- coefficient for `[X]`
- odds ratio for `[X]`
- statistical uncertainty, if relevant

## Clear communication

Suppose:

OR = 1.50

What does this mean?

A 1-unit increase in age is associated with a 50% increase in the odds of
cardiovascular disease, holding the other variables constant.

But the more important question is:

Does this mean that a 1-year increase in age increases the probability of
disease by 50 percentage points?

No.

An odds ratio describes a change in odds, not directly a change in probability.

### Why the odds ratio is not enough

The clinical question is usually about risk/probability, not odds.

Consider two patients:

- Patient A has a predicted disease probability of 5%.
- Patient B has a predicted disease probability of 50%.

Suppose both experience the same change in the predictor.

The resulting change in probability will not necessarily be the same.

This is the key complication of logistic regression:

> The effect of X on probability depends on where you start.

The relationship between X and probability is nonlinear.

The logistic curve is relatively flat near 0 and 1, and steepest in the middle.

Therefore:

> There is not necessarily one single "effect of X" on probability.

### What is a marginal effect?

A marginal effect asks:

*How much does predicted probability change when X changes?*

For a continuous variable:

*What happens to predicted probability if X increases by 1 unit?*

For example:

If age increases by 1 year, how much does predicted probability of disease
change?

In a nonlinear model, this is calculated on the probability scale, rather than
the log-odds scale.

With {marginaleffects}:

```{r}
#| label: marg-eff
#| eval: false

library(marginaleffects)

avg_slopes(model)
```

Conceptually, the output answers: A 1-unit increase in age → Δ predicted
probability

If the result were:

age: 0.012

the interpretation would be:

A one-year increase in age is associated with a 1.2 percentage-point increase in
predicted probability, on average.

Notice the distinction:

0.012 probability

1.2 percentage points

### The effect depends on the patient

Suppose we want to know the effect of age for a particular patient.

We might ask:

What is the marginal effect of age for a 50-year-old non-smoker with normal
blood pressure?

versus:

What is the marginal effect of age for a 75-year-old smoker with high blood
pressure?

These can be different.

We can calculate effects for specified covariate values:

```{r}
#| label: grid-eff
#| eval: false

slopes(
  model,
  newdata = datagrid(
    age = c(50, 75),
    smoking = c(0, 1),
    blood_pressure = c(120, 160)
  )
)
```

The important concept is not the syntax.

It is:

A marginal effect is conditional on the characteristics of the person for whom
we calculate it.

## Predicted probabilities versus marginal effects

These are different questions.

Predicted probability What is this patient's predicted probability of disease?

```{r}
#| label: pred
#| eval: false

predictions(
  model,
  newdata = datagrid(
    age = 60,
    smoking = 0,
    blood_pressure = 130
  )
)
```

Marginal effect How much does the predicted probability change if age increases?

```{r}
#| label: pro-marg
#| eval: false

slopes(
  model,
  newdata = datagrid(
    age = 60,
    smoking = 0,
    blood_pressure = 130
  )
)
```

**Predictions:**

> "What is the risk?"

**Marginal effects:**

> "How much does the risk change?"

Both are useful, but they answer different scientific questions.

## Average marginal effects

Often we do not want the effect for one particular patient.

We want:

What is the average effect in this population?

That is the average marginal effect (AME).

```{r}
#| label: avg-slopes
#| eval: false

avg_slopes(model)
```

Conceptually:

| Patient 1: effect of X = ...
| Patient 2: effect of X = ...
| Patient 3: effect of X = ...
|        ⋮
| Patient N: effect of X = ...

Average = AME

The interpretation is:

On average, a one-unit increase in X changes predicted probability by Δ
percentage points.

This is often much easier to communicate than an odds ratio.

## Interactions: when the effect depends on something else

Suppose our model includes:

Age × Smoking

The scientific question becomes:

Does the effect of age on disease probability differ between smokers and
non-smokers?

The model might be:

```{r}
#| label: model1
#| eval: false

model <- glm(
  disease ~ age * smoking + blood_pressure,
  data = data,
  family = binomial
)
```

The temptation is to look at the interaction coefficient and interpret it
directly.

Don't.

The interaction coefficient is on the log-odds scale.

Instead, calculate the marginal effect of age at different smoking statuses:

```{r}
#| label: int-model1
#| eval: false

slopes(
  model,
  variables = "age",
  by = "smoking"
)
```

Now the result directly addresses the scientific question:

Effect of age among non-smokers Effect of age among smokers

And you can ask:

Is the difference between those marginal effects clinically meaningful?

## Visualizing the interaction

```{r}
#| label: plotting1
#| eval: false

plot_predictions(
  model,
  condition = c("age", "smoking")
)
```

## Comparing clinically meaningful changes

A one-unit change is not always clinically meaningful.

For example, instead of asking:

What is the effect of a 1-unit increase in blood pressure?

we might ask:

What is the change in predicted probability when blood pressure increases from
120 to 140 mmHg?

For this, use a comparison:

```{r}
#| label: clin-inform
#| eval: false

avg_comparisons(
  model,
  variables = list(
    blood_pressure = c(120, 140)
  )
)
```

This produces a quantity directly corresponding to the clinical question:

Predicted probability at 140 − Predicted probability at 120 = change in
probability

## Uncertainty A marginal effect is an estimate, not a known quantity.

Therefore we also report uncertainty:

- Estimate
- 95% CI
- p-value

For example:

AME = 0.018 (95% CI 0.010--0.026)

Interpretation:

On average, a one-unit increase in X is associated with a 1.8 percentage-point
increase in predicted probability, with a 95% confidence interval from 1.0 to
2.6 percentage points.
