# Evaluating a crystal ball - part 1 {#sec-classification}

> The future is now

```{r}
#| label: setup
#| include: false

#library(tidyverse)
#ddagit782594dalibrary(tidyverse)
library(dplyr)
library(ggplot2)
library(yardstick)
library(probably)
library(tidyr)
#hd_data <- readr::read_csv(
#  "https://raw.githubusercontent.com/DanMazJen/medicinsk-statistik/main/data/hd_data.csv"
#)
data(hd_data, package = "medisr")
```

## Pree-class Literature

- [Marginal effects chatpter 5
  ](https://marginaleffects.com/chapters/predictions.html)
  - read

# Slides

> [**Session 15 slides**](../slides/day15.html)

<div>

```{=html}
<iframe class="slide-deck" src="../slides/day15.html"></iframe>
```

</div>

::: {.callout-note collapse="true"}
## Teacher note

The slides contain speaking notes that you can view by pressing 'S' on the
keyboard.
:::

## Case session 19: ""

## Content (still being worked on)

The established ABCD framework is

- A = calibration-in-the-large,
- B = calibration slope,
- C = discrimination/C-statistic,
- D = clinical usefulness/decision-curve analysis.

[see framework here](https://pmc.ncbi.nlm.nih.gov/articles/PMC4155437/)

The `ad_validation_logistic` data are particularly convenient because they
already contain observed classes and out-of-sample predicted probabilities from
a logistic regression model.

### Dataset context:

```{r}
#| label: prep-dataset
#| echo: false
#| output: false

#ad_data <- medisr::ad_data
ad_data <- medisr::ad_data

ad_data$diagnosis <- factor(ad_data$diagnosis,
  levels = c("Impaired", "Control"),
  labels = c("impaired", "control"),
)

ad_data$Genotype <- factor(ad_data$Genotype,
  levels = c("E3E3", "E3E4", "E4E4", "E2E3", "E2E4", "E2E2"),
  labels = c("E2E2", "E2E3", "E3E3", "E3E4", "E2E4", "E4E4")
)

# Create genotype dummy variables
# Currently not used!
#ad_data$E2 <- ifelse(grepl("2", ad_data$Genotype), 1, 0)
#ad_data$E3 <- ifelse(grepl("3", ad_data$Genotype), 1, 0)
#ad_data$E4 <- ifelse(grepl("4", ad_data$Genotype), 1, 0)

# Split data: every second observation
test <- ad_data[seq(1, nrow(ad_data), by = 2), ]
train  <- ad_data[seq(2, nrow(ad_data), by = 2), ]
```

The dataset was made based on the following logistic regression classifier:

```r
# Logistic regression
model <- glm(
  impairment ~ age + male + tau + p_tau + Ab_42 + Genotype,
  data = train,
  family = binomial
)
```

So

$$
  impairment_i \sim age + male + tau + p\_tau + Ab\_42 + genotype
$$
where:

- `age`: Normalized unit
- `male`: Reported sex was male
- `tau`: Tau protein stabilizes microtubules in axons. They can form insoluble
  neurofibrillary tangles when floating around freely. Here, the concentration
  was measured in the cerebrospinal fluid via lumbar puncture.
- `p_tau`: Phosphorylated tau neurofibrillary tangles in the cerebrospinal
  fluid. Here, the concentration was measured in the cerebrospinal fluid via
  lumbar puncture.
- `Ab_42`: Amyloid beta 42 (Aβ42) is a 42-amino-acid peptide derived from the
  amyloid precursor protein that forms the plaques found in the brains of people
  with Alzheimer's disease.
- `genotype`: The APOE allele pair. Can be E2E2, E2E3, E3E3, E3E4, E2E4, or
  E4E4. .

We want to predict patients with cognitive impairment defined by a gold-standard
neurophyschological test (such as the Wechsler's Adult Intelligence Score).

```{r}
#| label: see-data

test |>
    select(diagnosis,age,male,tau,p_tau,Ab_42,Genotype)
```

```{r}
#| label: data-plt1

p1 <- ggplot(test, aes(tau, p_tau, colour = diagnosis)) +
  geom_point(
    aes(alpha = factor(male)),
    size = 2.4,
    position = position_jitter(width = 0.03, height = 0.03)
  ) +
  facet_wrap(~ Genotype) +
  scale_alpha_manual(
    values = c(`0` = 0.55, `1` = 0.9),
    name = "Male"
  ) +
  labs(
    x = "Tau",
    y = "p-tau",
    colour = "Diagnosis"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

p1
```

```{r}
#| label: data-plt2

biomarkers <- test |>
  pivot_longer(
    cols = c(tau, p_tau, Ab_42),
    names_to = "biomarker",
    values_to = "value"
  )

p2 <- ggplot(
  biomarkers,
  aes(diagnosis, value, fill = diagnosis)
) +
  geom_violin(
    alpha = 0.25,
    colour = NA,
    trim = FALSE
  ) +
  geom_boxplot(
    width = 0.18,
    outlier.shape = NA,
    alpha = 0.65
  ) +
  geom_jitter(
    width = 0.08,
    height = 0,
    size = 1.4,
    alpha = 0.55
  ) +
  facet_wrap(~ biomarker, scales = "free_y") +
  labs(
    x = NULL,
    y = NULL,
    fill = "Diagnosis"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

p2
```

### Evaluating a Prediction Model

A prediction model can be evaluated from two things:

- what actually happened;
- and what the model predicted.

Here we use predictions from a logistic regression model. The model has already
been trained, so we do not fit a model. We simply evaluate its predictions.

```{r}
#| label: get-data

ad_validation_logistic <- medisr::ad_validation_logistic
```

Class is the observed outcome. `.pred_impaired` is the model's predicted
probability that the patient will show cognitive impairment.

```{r}
#| label: looking-at-data

ad_validation_logistic
```

### A gentle overview

How many was correctly classified?

We can make a **confusion matrix** --- a table that shows how many observations
were classified correctly and how many were assigned to the wrong class.

```{r}
#| label: conf-mat

conf_mat(
  predictions,
  truth = Class,
  estimate = .pred_class
)
```

A confusion matrix is useful because it lets us see what kinds of mistakes the
model is making, rather than giving us just one summary number.

We spot an interesting pattern:

- There are many more control participants than cognitively impaired
  participants.
- labeling the cognitive impairment was not a success --- only 25 out of 44 were
  labeled correctly!
- labeling the control participants much better --- 109 out of 123 were labeled
  correctly!

We can quickly calculate that `r 109/123`% of the controls were labeled
correctly while only `r 25/44`% of the cognitively impaired were labeled
correctly.

This seems like a reasonable way to evaluate the model. But there is a problem:
these percentages answer a slightly different question for each class, and they
can be strongly affected by how common each class is.

In particular, there are many more controls than cognitively impaired
participants. A model can therefore appear to perform well simply by being good
at predicting the more common class.

So rather than looking only at the percentage of observations classified
correctly, we can use metrics that explicitly describe different types of
classification errors.

### Accuracy

Accuracy answers a simple question:

> How often does the model get the class right?

Accuracy is the proportion of all observations that were classified correctly:

$$ Accuracy = \frac{TP + TN}{TP + TN + FP + FN}

$$

Here, TP means true positives, TN means true negatives, FP means false
positives, and FN means false negatives.

To calculate accuracy, we first turn probabilities into classes. Here we use 0.5
as the threshold: if the predicted probability of impairment is at least 0.5, we
classify the participant as impaired; otherwise, we classify them as a control.

```{r}
#| label: accuracy

predictions <- ad_validation_logistic |>
    mutate(
        .pred_class = probably::make_two_class_pred(
            .pred_impaired,
            levels(Class),
            threshold = 0.5 )
        )

accuracy(predictions, Class, .pred_class)
```

The accuracy was 0.802. This means that about 80.2% of all participants were
classified correctly.

Accuracy is easy to understand, but it has an important limitation: it does not
tell us which class the model gets wrong. It can also be misleading when the
classes are unbalanced.

For example, imagine that 95% of participants were controls. A model that always
predicted "control" would have 95% accuracy, despite completely failing to
identify anyone with cognitive impairment.

This is why it is useful to look separately at the model's performance for the
different classes.

### Specificity

Specificity asks:

> Among people who are not "impaired", how many does the model correctly
> identify?

Specificity measures how well the model identifies the negative class --- here,
the control participants.

In other words, specificity is the proportion of actual controls that the model
correctly labels as controls.

$$
  Sensitivity = \frac{TP}{TP + FN}
$$

```{r}
#| label: specificity

specificity(predictions, Class, .pred_class)
```

The specificity was 0.886. This means that 88.6% of the participants who were
actually controls were correctly identified as controls.

A highly specific model produces few false positives: it rarely labels a control
participant as cognitively impaired.

Specificity is therefore particularly useful when false positives are costly or
problematic. For example, in a screening context, a false positive might lead to
unnecessary further testing or concern.

### Sensitivity

Sensitivity asks:

> Among people who are "impaired", how many does the model correctly identify?

Sensitivity measures how well the model identifies the positive class --- here,
cognitively impaired participants.

$$
  Specificity = \frac{TN}{TN + FP}
$$

In other words, sensitivity is the proportion of participants who are actually
impaired that the model correctly identifies as impaired.

```{r}
#| label: sensitivity

sensitivity(predictions, Class, .pred_class)
```

The sensitivity was 0.568. This means that 56.8% of the participants who were
actually cognitively impaired were correctly identified by the model.

A highly sensitive model produces few false negatives: it rarely misses a
participant who is actually impaired.

This distinction between sensitivity and specificity is important. A model can
have high specificity while having relatively low sensitivity, as we see here.
Our model is quite good at recognizing controls, but it misses a substantial
proportion of participants with cognitive impairment.

### Discussion task: Setting a proability threshold

Sensitivity and specificity depend on the chosen probability threshold. We used
0.5 above, but there is nothing inherently special about 0.5.

Changing the threshold changes the balance between false positives and false
negatives. A lower threshold makes the model more willing to classify someone as
impaired. This will generally increase sensitivity, because fewer impaired
participants are missed, but it can also decrease specificity, because more
controls may be incorrectly labeled as impaired.

Conversely, a higher threshold makes the model more conservative about calling
someone impaired. This can increase specificity but may reduce sensitivity.

In a clinical application, the useful threshold should therefore be chosen based
on the consequences of the two types of errors, rather than simply because 0.5
is a convenient number.

This is one reason why reporting accuracy alone is often not enough: two models
can have similar accuracy while having very different sensitivity and
specificity, and those differences can matter a great deal in practice.

**Discucss with your neighbor:**

- which diagnosis would you prefer a lower probability? Why?
- which diagnosis would you prefer a higher probability? Why?

Consider consequences of the diagnosis, like:

- Cost of treatment
- Side-effects
- Effectiveness of treatment

### ROC and AUC

Instead of choosing one threshold, we can look at model discrimination across
all possible thresholds.

The ROC curve shows sensitivity against 1 − specificity.

```{r}
#| label: roc

roc_curve(ad_validation_logistic, Class, .pred_impaired, event_level = "first" ) |>
    autoplot()
```

The area under this curve is the AUC.

```{r}
#| label: auc

roc_auc(ad_validation_logistic, Class, .pred_impaired)
```

AUC can be interpreted as the probability that the model assigns a higher
predicted probability to a randomly chosen "impaired" observation than to a
randomly chosen "control" observation.

AUC measures discrimination. It does not tell us whether the predicted
probabilities themselves are accurate.

### Calibration

Calibration asks a different question:

When the model says 70%, does the event happen about 70% of the time?

A simple calibration plot compares predicted probability with the observed
proportion.

The probably package provides this directly:

```{r}
#| label: calibration

cal_plot_breaks(ad_validation_logistic, Class, .pred_impaired)
```

A well-calibrated model follows the diagonal line.

We can also fit a smooth calibration curve:

```{r}
#| label: cal-smooth

cal_plot_logistic(ad_validation_logistic, Class, .pred_impaired, smooth = FALSE)
```

Calibration is about the accuracy of the probabilities. A model can have
excellent discrimination and poor calibration.

### ABCD of model validation

A useful way to remember prediction-model evaluation is the ABCD framework:

- A --- Calibration-in-the-large: are predictions systematically too high or too
  low?
- B --- Calibration slope: are predictions too extreme or not extreme enough?
- C --- C-statistic: how well does the model discriminate between outcomes?
- D --- Decision curve: does using the model improve clinical decisions?

[This framework](https://pmc.ncbi.nlm.nih.gov/articles/PMC4155437/) separates
statistical performance from clinical usefulness.

For example, the calibration model can be written very simply as:

```{r}
#| label: cal-model

calibration_model <- ad_validation_logistic |>
    mutate(
        outcome = Class == "impaired",
        logit_p = log(.pred_impaired / (1 - .pred_impaired))
    ) |>
        glm(outcome ~ logit_p, data = _, family = binomial)

broom::tidy(calibration_model)
```

The intercept is related to calibration-in-the-large (A) and the coefficient of
`logit_p` is the calibration slope (B).

Ideally, we'd see that the calibration-in-the-large is (close to) 0 and that the
calibration slope is (close to) 1.

This is called the **C-statistic** of the AUC.

**D**

The D component asks whether the predictions are useful for making decisions.
That requires specifying a clinical action and the range of probability
thresholds at which that action would be taken.

### Effect size estimation

Prediction performance and effect estimation are different questions.

A logistic regression coefficient describes the association between a predictor
and the outcome. For example:

How much does the odds of "impaired" change for a one-unit increase in age?

The dataset used here contains only the predictions, not the original
predictors. Therefore, we cannot recover the original regression coefficients or
their odds ratios from `ad_validation_logistic`.

What we can estimate is the association between the model's predicted risk and
the observed outcome. The calibration slope above is one such measure.

This distinction is important:

Prediction asks: How accurately can we predict what will happen?

Effect estimation asks: How does changing a particular variable relate to what
happens?

A model can be excellent for prediction without giving an unbiased estimate of a
causal effect.

### Putting it all together

A simple evaluation therefore looks like this:

<!-- panache-ignore-format-start -->
|Question |  Measure|
|---------|---------|
|Does the model classify correctly at this threshold?	|Accuracy|
|Does it avoid false positives?	                        |Specificity|
|Does it avoid false negatives?	                        |Sensitivity|
|Can it distinguish the two outcomes?	                |ROC / AUC|
|Are the probabilities accurate?	                    |Calibration|
|Does it improve clinical decisions?	                |Decision curve|
|What is the effect of a predictor?	                    |Effect estimate|
<!-- panache-ignore-format-end -->

The key idea is that no single metric evaluates a prediction model completely.

A useful model should discriminate well, produce sensible probabilities, and ---
most importantly --- lead to better decisions.
