# Conditioning and polynomial madness {#sec-interaction}

> "x." --- John Doe

```{r}
#| label: setup
#| include: false

#library(tidyverse)
#ddagit782594dalibrary(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)

hd_data <- readr::read_csv(
  "https://raw.githubusercontent.com/DanMazJen/medicinsk-statistik/main/data/hd_data.csv"
)
```

## Pree-class Literature

- [s](s)

## Case session 11: ""

At AalboR Statistical Hospital, a new exercise-based cardiac assessment protocol
has recently been introduced following an initial data review suggesting that
patients with lower max heart rate are more likely to have underlying heart
disease. Based on this, some clinicians have started using the max heart rate
during exercise as a quick indicator when evaluating chest pain patients.

However, during a recent internal meeting, a cardiologist raises a concern:

> *"I'm not convinced this applies to our patients. Especially those who develop
> chest pain during exercise --- their heart rate response might behave
> differently. If that's true, we could be misinterpreting results in a subgroup
> of patients. I want to see this association for myself"*

You are asked to explore the hospital's patient data to assess whether the
relationship between Age and max heart rate --- and the interpretation of max
heart rate more generally --- appears consistent across patients, or whether it
differs depending on exercise-induced angina, sex, or other patient
characteristics.

You are tasked with the decision: *Should this new assessment approach be used
broadly, or are there patient groups where it may be misleading?*

```{r}
hd_data <- hd_data |>
    mutate(
        ExAng = factor(ExAng,
            levels = c(0, 1),
            labels = c("no", "yes")
        ),
        HD = factor(HD,
            levels = c("No", "Yes"),
            labels = c("no", "yes")
        ),
        ChestPain = factor(ChestPain),
        Thal = factor(Thal),
        fasting_blood_glucose = factor(Fbs,
            levels = c(0, 1),
            labels = c("normal", "high")
        )
    )

hd_data |>
  ggplot(ggplot2::aes(Age, MaxHR, color = HD)) +
  geom_point() +
  geom_smooth(method = "lm")
```

## Reading Task: Conditional manatees?

...

## Adding the variable to a regression model is not enough

```{r}
model_additive <-
    lm(MaxHR ~ Age + HD,
    data = hd_data)

library(marginaleffects)

plot_predictions(
    model_additive,
    condition = c("Age", "HD")
)

model_additive |>
    avg_slopes()
```

## Adding an interaction

```{r}
model_interact <-
    lm(MaxHR ~ Age + HD + Age:HD,
    data = hd_data)

plot_predictions(
    model_interact,
    condition = c("Age", "HD")
)
```

## Interpreting an interaction estimate

```{r}
model_interact |>
    summary()
```

```{r}
model_params <- model_interact |>
    parameters::model_parameters()
```

$$
Y^=213.93−1.06(age)−53.84(hd)+0.689(age×hd)
$$

where:

- age = age in years -
- hd = 0 (no heart disease) or 1 (heart disease)

```{r}
model_params <-
    data.frame(
        intercept_est = model_params[1,2],
        age_est = model_params[2,2],
        hd_est = model_params[3,2],
        age_hd_interact_est = model_params[4,2]
    )

model_params
```

### Step 1:

Step 1: Write separate equations for each group No heart disease (hd = 0)

Substitute hd = 0:

$ ⁄bar{Y} = 213.93−1.06(age) $

Interpretation:

For people without heart disease, each additional year of age is associated with
about 1.06 units lower predicted outcome. Heart disease (hd = 1)

Substitute hd = 1:

$ Y^ =213.93−1.06(age)−53.84+0.689(age) $

Combine terms:

$Y^=160.09−0.368(age)$

Interpretation:

For people with heart disease, each additional year of age is associated with
only about 0.37 units lower predicted outcome. The age effect is less negative
because of the positive interaction term.

### Step 2: Plug in specific ages

**Age = 40**

*No heart disease:* 213.93−1.06(40)=171.67

*Heart disease:* 160.09−0.368(40)=145.37

*Difference:* 145.37−171.67=−26.30

> At age 40, heart disease is associated with about 26 units lower predicted
> outcome.

**Age = 70**

*No heart disease:* 213.93−1.06(70)=140.00

*Heart disease:* 160.09−0.368(70)=134.34

*Difference:* 134.34−140.00=−5.66

> At age 70, heart disease is associated with only about 6 units lower predicted
> outcome.

### Step 3: Interpret the interaction

The interaction coefficient is: **0.689**

This means that the effect of heart disease becomes 0.689 units less negative
for every additional year of age.

In other words: The difference between the heart disease and no-heart-disease
groups shrinks by about 0.689 units for each additional year of age.

We can again plug the values in to examine this:

```r
Age HD=0    HD=1    Difference (HD−No HD)
40  171.67  145.37  -26.30
50  161.11  141.69  -19.41
60  150.54  138.01  -12.53
70  139.98  134.32  -5.66
```

Students often find the table easier than interpreting the interaction
coefficient directly because they can see how the gap between groups changes
with age.

```{r}
plot_slopes(model_interact, variables = "HD", condition = "Age") +
    geom_hline(yintercept = 0, linetype = "dotted") +
    labs(
        y = "expected difference (MaxHR | HD)"
    )

slopes(model_interact,
  variables = "HD",
  newdata = datagrid(Age = range))

slopes(model_interact,
  variables = "HD",
  newdata = datagrid(Age = range),
  hypothesis = "b2 - b1 = 0"
)
```

# Symmetry of the linear interaction

Consider for example the GDP and terrain ruggedness problem. The interaction
there has two equally valid phrasings.

How much does the influence of ruggedness (on GDP) depend upon whether the
nation is in Africa? How much does the influence of being in Africa (on GDP)
depend upon ruggedness?

## Symmetry of the Linear Interaction

Consider McElreath's GDP and terrain ruggedness example.

One might ask:

1. How much does the effect of terrain ruggedness on GDP depend on whether a
   nation is in Africa?
2. How much does the effect of being in Africa on GDP depend on terrain
   ruggedness?

These sound like different questions, but in a linear model they are
mathematically identical.

The original model is

$$
\log(\text{GDP}_i) \sim \text{Normal}(\mu_i,\sigma)
$$

with

$$
\mu_i = \alpha + \gamma_i \,\text{rugged}_i + \beta_2 \,\text{Africa}_i
$$

and

$$
\gamma_i = \beta_1 + \beta_3 \,\text{Africa}_i.
$$

Substituting the second equation into the first gives

$$
\mu_i = \alpha + (\beta_1 + \beta_3 \,\text{Africa}_i)\,\text{rugged}_i + \beta_2 \,\text{Africa}_i.
$$

Expanding:

$$
\mu_i = \alpha + \beta_1 \,\text{rugged}_i + \beta_3 \,\text{rugged}_i \times \text{Africa}_i + \beta_2 \,\text{Africa}_i.
$$

Now collect the terms containing $\text{Africa}_i$:

$$
\mu_i = \alpha + \beta_1 \,\text{rugged}_i + (\beta_2 + \beta_3 \,\text{rugged}_i)\,\text{Africa}_i.
$$

Define

$$
G_i = \beta_2 + \beta_3 \,\text{rugged}_i.
$$

Then

$$
\mu_i = \alpha + \beta_1 \,\text{rugged}_i + G_i\,\text{Africa}_i.
$$

The quantity $G_i$ plays the same role for the Africa effect that $\gamma_i$
played for the ruggedness effect.

### Key takeaway

The interaction coefficient $\beta_3$ simultaneously describes

- how the effect of ruggedness changes across Africa/non-Africa, and
- how the effect of Africa changes across levels of ruggedness.

In a linear model, interactions are symmetric. The choice of which variable is
called the "predictor" and which is called the "moderator" is often a matter of
interpretation rather than mathematics.

# Continuous interactions

```{r}
library(sjPlot)
data(efc)

m <- lm(
  neg_c_7 ~ c12hour + sex + c12hour:sex,
  data = efc
)

# fit model with 3-way-interaction
fit <- lm(neg_c_7 ~ c12hour * barthtot * c161sex, data = efc)

# select only levels 30, 50 and 70 from continuous variable Barthel-Index
sjPlot::plot_model(fit, type = "pred", terms = c("c12hour", "barthtot [30,50,70]", "c161sex"))
```

where:

neg_c_7 = caregiver burden (outcome) c12hour = hours of care provided sex =
caregiver sex

Then ask the two interaction questions:

How much does the association between caregiving hours and burden depend on
caregiver sex?

and

How much does the difference between male and female caregivers depend on
caregiving hours?

Students usually perceive these as different questions, but the interaction
coefficient is the same parameter in both cases.

```{r}
library(marginaleffects)

marginaleffects::slopes(
  fit,
  variables = "c12hour",
  by = "c161sex"
)

marginaleffects::avg_slopes(
  fit,
  variables = "c12hour",
  by = list("c12hour", "barthtot" = c(30,50,70))
)

marginaleffects::plot_slopes(
  fit,
  variables = "c12hour",
  condition = list("c12hour", "barthtot" = c(30,50,70))
)

marginaleffects::plot_predictions(
  fit,
  variables = "c12hour",
  condition = list("c12hour", "barthtot" = c(30,50,70))
)
```

```{r}
marginaleffects::comparisons(
  fit,
  variables = "c161sex",
  by = "c12hour"
)
```
