# Marginalizing the problem {#sec-marginal-effects}

```{r}
#| label: setup
#| include: false

#library(tidyverse)
#ddagit782594dalibrary(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)
library(marginaleffects)
library(tidyr)

#hd_data <- readr::read_csv(
#  "https://raw.githubusercontent.com/DanMazJen/medicinsk-statistik/main/data/hd_data.csv"
#)
#data(hd_data, package = "medisr")
```

## Pree-class Literature

{{< include /includes/curriculum/_session15.qmd >}}

## Slides

> [**Session 15 slides**](../slides/day15.html)

<div>

```{=html}
<iframe class="slide-deck" src="../slides/day15.html"></iframe>
```

</div>

::: {.callout-note collapse="true"}
## Teacher note

The slides contain speaking notes that you can view by pressing 'S' on the
keyboard.
:::

## Case session 15: "The odds ratio is 0.65... and so what?"

At the stroke research meeting, the biostatistician puts a slide on the screen.

> Thrombectomy vs. no thrombectomy: OR = 0.65

The clinical team pauses.

> "That sounds promising," someone says. "But what does 0.65 actually mean for a
> patient?"

The analysis comes from an observational study of patients with acute stroke.
The outcome is death within one year, and the researchers have fitted a logistic
regression model to account for differences between patients who received
thrombectomy and those who did not.

The statistician explains that the regression coefficient corresponds to an odds
ratio of 0.65.

Another question immediately follows:

> "But what we really want to know is whether patients have a lower probability
> of dying if they receive thrombectomy. Why are we talking about odds?"

Now the room gets filled with discussion.

Suppose 20% of patients in the control group die within one year. What would you
actually want to show a clinician---or a patient?

- An odds ratio of 0.65?
- The probability of death under each treatment?
- The absolute difference in probability of death?
- Something else?

And whose probability are we talking about?

- A particular patient with a particular age, diabetes status, and clinical
  profile?
- An older adult with diabetes?
- The average patient in this study?
- Or the population-level probability we would see if everyone received
  thrombectomy, compared with what we would see if no one received it?

These are related questions, but they are not the same question. The clinical
team decides to work through the problem one scientific question at a time.

1. Is thrombectomy associated with a different probability of death?
2. Does the effect differ for older adults with diabetes?
3. What does the model imply for the average stroke patient?
4. What if everyone could participate twice --- both in the treatment and the
   no-treatment group?

The team notices something --- the regression model has produced one
coefficient, but the researchers have several different scientific questions.

So they ask:

> What exactly is the model estimating---and how do we turn that into the
> quantity we actually want to know?

Your task is to make sure that the quantity we calculate and communicate answers
the scientific question we started with.

<!--
:::{.hidden}
Case session 15: "If you're already dead, your odds of dying don't increase."

At a PhaRmaceutic RStudio AS, a new compound is being evaluated for toxicity
before it can move further through development. In an early preclinical
experiment, groups of laboratory animals are exposed to increasing doses of the
compound, and investigators record how many animals die at each dose.

At the weekly safety meeting, the lead researcher summarizes the results:

> "As the dose increases, mortality increases."

One colleague looks at the highest-dose groups and asks:

> "But if almost all the animals are already dead, can another increase in dose
> really have the same effect?"

The team looks more closely at the data. Mortality increases rapidly through the
middle of the dose range, but begins to level off at high doses.

The investigators fit a logistic regression model because the outcome is binary:
each animal either died or survived. The model produces a coefficient for dose.

> "A one-unit increase in dose increases mortality by 8.7," the lead researcher
> says.

The biostatistician stops them.

> "The coefficient is on the log-odds scale. That's not a change in
> probability."

The team now faces a more useful question:

- How does increasing dose change the probability of death?

And because the dose-response relationship is nonlinear, another question
follows:

- At what dose are we asking about the effect?

Your task is to investigate the dose-response relationship using a quadratic
logistic regression model. You will move from regression coefficients to
predicted probabilities and marginal slopes, examining how the effect of dose
changes across the dose range.
:::
-->

## Reading Task: Communicating results of regression models

*The following text was synthesized from @marginalbook Chapter 2 and adjusted to
the topic of the lecture.*

A fitted statistical model produces coefficients, but these coefficients are not
necessarily the quantities we are scientifically interested in. The **estimand**
is the quantity that answers the research question---the thing we actually want
to learn about. A model coefficient, by contrast, is simply a statistical
quantity used to represent patterns in the data. It is an abstract output of the
model, not medical or clinical information in itself.

For example, a logistic regression produces coefficients on the log-odds scale.
Although this may be useful for fitting the model, a log-odds coefficient is
often difficult to interpret in practical or clinical terms. It does not
directly tell us how a patient's probability of an outcome changes.

Even probabilities, which are much more familiar, can be surprisingly difficult
to interpret. People do not always reason about probabilities intuitively, and
research in psychology and behavioral economics has documented many ways in
which our judgments about probabilities can be misleading.

This raises an important question: if simple probabilities can be difficult to
understand, why make our results even harder to interpret by focusing on
quantities such as odds ratios or other model coefficients? The problem becomes
even greater with more complicated models involving nonlinear relationships,
interactions, random effects, Bayesian models, or machine-learning methods.

The key point is that the model coefficient should usually not be the end of the
analysis. We should start with the scientific question and the estimand we want
to learn about, and then use the statistical model as a tool for **estimating**
that quantity. When the raw model coefficients are difficult to interpret, we
can transform the model results into quantities that are easier to understand
and that directly address the research question.

For example, rather than reporting only a logistic regression coefficient or
odds ratio, we might report predicted probabilities for clinically meaningful
values of the predictors and compare those probabilities. The model is then
doing what it is meant to do: providing a statistical representation of the data
that helps us learn about a scientifically meaningful quantity.

## Discussion exercise: "What do we actually want to know?"

> An observational study retrospecively investigates whether a new treatment
> reduces the risk of a complication after surgery. The researchers fit a
> logistic regression model and obtain $β_1$ (odds ratio) as 0.65 for treatment
> vs. control.

**Discuss with our neighbor:**

1. What is the scientific question? What would you actually want to know about
   the treatment?

2. What is the model giving us? What exactly is the odds ratio of 0.65? Is it
   itself a clinical outcome or something else?

3. What would be easier to interpret? Imagine the complication occurs in 20% of
   patients in the control group. What kind of result would you rather show a
   clinician or patient: the odds ratio, a risk/probability, an absolute risk
   difference, or something else? Why?

4. What is the estimand? Try to state, in one sentence, the quantity you would
   ideally like to learn about.

5. If the model coefficient is not the scientific answer, what role does the
   model play?

## Marginalizing the problem - our workflow:

Throughout this session, we'll use this workflow:

1. What is my scientific question --- **the estimand**?
2. Choose machinery to model data --- **the estimator**?
3. What did the regression estimate --- **the coefficient**?
4. What quantity do I actually want to communicate?
5. At what values?

- A particular patient?
- A clinically meaningful scenario?
- The average population?
- Different levels of an interacting variable?

6. Post-model transformations --- **the estimate**
7. Communicate in clinically meaningful units

::: {#fig-marginal-effects-workflow .column-page fig-cap="From the scientific question to a clinically interpretable marginal effect."}
```{mermaid}
%%| label: workflow-diag

flowchart LR
    A["1. Scientific question<br/><br/>What happens to P(Y)<br/>when X changes?"]

    A --> B["2. Estimator<br/><br/>logistic regression"]

    B --> C["3. Regression estimate<br/><br/>log-odds"]

    C --> D["4. Quantity to communicate<br/><br/>Probability"]

    D --> E{"5. Who/where are we estimating the effect?"}

    E --> E1["Particular patient"]
    E --> E2["Clinically meaningful<br/>scenario"]
    E --> E3["Average population"]
    E --> E4["Different levels of<br/>an interacting variable"]

    E1 --> F["6. Marginal effects"]
    E2 --> F["6. Marginal effects"]
    E3 --> F["6. Marginal effects"]
    E4 --> F["6. Marginal effects"]

    F --> G["7. Communicate in<br/>clinically meaningful units"]

    style A fill:#E8F1FB,stroke:#3B6EA5,stroke-width:2px
    style B fill:#F3F3F3,stroke:#666,stroke-width:2px
    style C fill:#E8F5E9,stroke:#4C8C4A,stroke-width:2px
    style D fill:#FFF4D6,stroke:#C79A24,stroke-width:2px
    style E fill:#EDE7F6,stroke:#7656A6,stroke-width:2px
    style F fill:#E8F5E9,stroke:#4C8C4A,stroke-width:2px
    style G fill:#DDF3E0,stroke:#3D8B40,stroke-width:3px
```
:::

As shown in @fig-marginal-effects-workflow, we move from the **scientific
question**, through the **model scale**, to a **marginal effect on the
probability scale**.

## Step 1: Four scientific questions

Throughout this session, we'll tackle four research questions one-by-one:

1. Does the population who got thrombectory have lower probability of death at
   1-year follow-up after stroke compared to the population who did not get
   thrombectomy?

2. What is the effect of thrombectory in older adults with diabetes, compared to
   older adults without diabetes?

3. What is effect of the average person profile in the sample?

4. What would be the average probability of death, had all participants received
   thrombectomy, compared to not receiving it?

::: {.callout-warning}
## From model to meaning

In nonlinear models, you cannot generally interpret the regression coefficient
as the change in probability.

The purpose of `{marginaleffects}` is essentially to help us make this
translation: From a fitted generalized model → to quantities that answer the
scientific question.
:::

### Context

Go ahead and read more about the dataset

```r
?medisr::thrombectomy
```

### The estimand

We are interested in the probability of death at 1-year follow-up:

$$
  P(death = 1 | X)
$$

for an observational study looking at the effect of thrombectomy after stroke:

```{r}
#| label: smoking-dag
#| echo: false

dag <- ggdag::dagify(
  death ~ trt + age + diabetes + hyp_ten,
  trt ~ age + diabetes + hyp_ten,
  hyp_ten ~ diabetes + age,
  diabetes ~ sex + age
)

ggdag::ggdag(dag) +
  ggdag::geom_dag_point(size = 20) +
  ggdag::geom_dag_text(size = 3) +
  ggdag::theme_dag()

# ggdag::ggdag(dag, text = FALSE, use_labels = "name")

# dagitty::adjustmentSets(dag,exposure="trt","death")
```

We need to adjust for age, hypertension and diabetes. We will still incl. sex,
as the dataset is unbalanced and we want to adjust for that bias.

## Step 2: The estimator

We follow individuals for one year and there is no censoring. The probability of
death is relative high (as we'll see in a minute) and *n* can be defined as the
sample size. So we are able to use logistic regression.

::: {.callout-warning}
## What about Poisson regression?

In many ways, this would actually have been the natural choice here, as the
dataset has a time aspect and we follow people for longer and some may become
censored my migrating etc.

Here, the choice was purely educational --- you'll use poisson regression on the
length of stay in the exercises 😄
:::

$$
  y_i = Binomial(n,p_i) \\
  logit(p_i) = α + βx_i
$$
where *n* is the number of observations and *p* is the probability of death.

The logit function is defined as the *log-odds*:

$$
  logit(p_i) = log(\frac{p_i}{1 - p_i})
$$

The "odds" of an event is just the probability it happens divided by the
probability it does not happen. In other words:

$$
  log(\frac{p_i}{1 - p_i}) = α + βx_i
$$
To figure out the definition of *p_i*, we do some algebra to solve for $p_i$:

$$
  p_i = \frac{exp( α + βx_i)}{1 + exp(α + βx_i)}
$$
The above function is the **logistic**

- Note that the log-odds space is linear (what we model)
- The logistic space is the sigmoidal shape (the transformation)

### The statistical models

We want to adjust for competing predictors of death and potential counfounders
in our model.

$$
  logit( \text{death at 1 year followup}) = α + β1_{thrombectomy}
                                            + β2_{diabites} + β3_{age} \\
  + β4_{hypertention} + β5_{sex} + β6_{thrombectomy×diabites} \\
  + β7_{thrombectomy×age}
  + β8_{thrombectomy×hyptertension} + β9_{thrombectomy×sex} \\
$$

We implement this model in R using `glm()`:

```{r}
#| label: thrombectomy-model

thrombectomy <- medisr::thrombectomy |>
    mutate(death_1year_followup = if_else(time_to_death < 365 & death ==1,1,0),
    age_at_diagnosis = as.numeric(age_at_diagnosis)
    )

th_model <- glm(
  death_1year_followup ~ thrombectomy +
        prior_diabetes +
        age_at_diagnosis +
        prior_hypertention +
        sex +
        thrombectomy:prior_diabetes +
        thrombectomy:age_at_diagnosis +
        thrombectomy:prior_hypertention +
        thrombectomy:sex,
        family = binomial(link = "logit"),
  data = thrombectomy
)
```

## Step 3: Stock with regression coefficients...

Here's the model output:

```{r}
#| label: smoking-output

th_model |>
    parameters::parameters(exponentiate=TRUE) |>
    select(Parameter, Coefficient, SE) |>
    rename(Odds_ratio=Coefficient)
```

Cool, we got an OR = 9.06 for `thrombectomy`

> What does this mean?

### Rookie mistake #1

> Don't read this coefficient, thinking it generalizes to the sample!

This sample has odds of 0.00182 for the outcome when thrombectomy = 0 and **all
other coefficients are 0**.

In other words, the intercept describes the model's baseline odds for a
0-year-old female without diabetes or hypertension, and without thrombectomy.
Such a patient didn't participate in the study. Thus, this is out-of-sample
prediction --- and a pretty silly and unjustified extrapolation from what we're
trying to answer scientifically here.

The thrombectomy OR of 9.06 means that, holding all the other variables
constant, the odds are multiplied by 9.06 when thrombectomy changes from 0 to 1.

### Rookie mistake #2:

Thinking that an OR of 9.06 means a 9-fold increase in probability.

It doesn't. An odds ratio describes a change in odds, not probability. To get
the probability, you need to convert the log-odds back using the inverse logit
function `g()`.

For example, the model's baseline odds are:

$0.00182$

With thrombectomy:

$0.00182 × 9.06 ≈ 0.0165$

which corresponds to a probability of approximately 1.62%, not 9.06%.

So, OR = 9.06 does not mean that the probability of the outcome is 9.06 times
higher. It means that, for two otherwise identical patients, the patient with
`thrombectomy = 1` has 9.06 times the odds of the outcome compared with the
patient with thrombectomy = 0.

### Rookie mistake #3

A 1-unit increase in age is associated with a 0.7% increase in the odds of
cardiovascular disease, holding the other variables constant.

But the more important question is:

Does this mean that a 1-year increase in age increases the probability of
disease by 0.7% percentage points?

The answer is a stricking no!

An odds ratio describes a change in odds, not directly a change in probability.

The clinical question is usually about risk/probability, not odds.

Consider two patients:

- Patient A has a predicted disease probability of 5%.
- Patient B has a predicted disease probability of 50%.

Suppose both experience the same change in the predictor.

The resulting change in probability will not necessarily be the same.

This is the key complication of logistic regression:

> The effect of X on probability depends on where you start.

The relationship between X and probability is nonlinear.

The logistic curve is relatively flat near 0 and 1, and steepest in the middle.

Therefore:

> There is not necessarily one single "effect of X" on probability.

```{r}
#| label: slop-maybe
#| layout-ncol: 2
#| fig-subcap:
#|   - the odds ratio change as a function of 1-unit change in age
#|   - the probability change as a function of 1-unit change in age

plot_slopes(
  th_model,
  condition = "age_at_diagnosis",
  variables = "age_at_diagnosis",
  type = "link"
)
plot_slopes(
  th_model,
  condition = "age_at_diagnosis",
  variables = "age_at_diagnosis",
  type = "response"
)
```

### Thought experiment:

If you're already likely to die, will adding risk change something?

```{r}
#| label: slop-maybe1
#| layout-ncol: 2
#| fig-subcap:
#|   - the odds ratio change as a function of 1-unit change in age
#|   - the probability change as a function of 1-unit change in age

grid1 = datagrid(
    age_at_diagnosis = 100,
    prior_diabetes = 1,
    thrombectomy = 0,
    model = th_model
)

plot_predictions(
  th_model,
  newdata = grid1,
  variable = "prior_hypertention",
  by = "prior_hypertention"
)

grid2 = datagrid(
    age_at_diagnosis = 70,
    prior_diabetes = 1,
    thrombectomy = 0,
    model = th_model
)

plot_predictions(
  th_model,
  newdata = grid2,
  variable = "prior_hypertention",
  by = "prior_hypertention"
)
```

So only a 6% point absolute difference (78.1% to 84.1%) probability for a
100-year-old with diabetes but a 9.1 point absolute difference (33.4% to 42.5%)
probability for a 70-year-old with diabetes, both not receiving thrombectomy.

## Step 4: Clear communication --- transforming coefficients

Recall @sec-coef-interp where we learned how to plug in numbers from a fitted
regression in order to fit expectations or *predictions* of specific patients
profiles in the dataset.

We will do this again - but this time, we're fitting a logistic regression

```{r}
#| label: extract-bs

b = coef(th_model)
b
```

```{r}
#| label: profile-one

linpred_treatment_diabetes_50yo_female = b[1] +
    b[2] * 1 + b[3] * 1 + b[4] * 50 + b[5] * 0 +
    b[6] * 0 + b[7] * 1 + b[8] * 50 + b[9] * 0 + b[10] * 0

linpred_treatment_diabetes_50yo_female |> unname()
```

<!--panache-ignore-start-->
| Term | Value | Comment |
|---|---|---|
| `b[1]` | `* 1` | intercept |
| `b[2]` | `* 1` | thrombectomy = 1 |
| `b[3]` | `* 1` | diabetes = 1 |
| `b[4]` | `* 50` | age = 50 |
| `b[5]` | `* 0` | hypertension = 0 |
| `b[6]` | `* 0` | female = 0 |
| `b[7]` | `* 1` | thrombectomy * diabetes = 1*1 |
| `b[8]` | `* 50` | thrombectomy * age = 1*50 |
| `b[9]` | `* 0` | thrombectomy * hypertension = 1*0 |
| `b[10]` | `* 0` | thrombectomy * sex = 1*0 |

<!--panache-ignore-end-->

The other will be

```{r}
#| label: profile-two

linpred_no_treatment_diabetes_40yo_male = b[1] +
    b[2] * 0 + b[3] * 1 + b[4] * 40 + b[5] * 0 +
    b[6] * 1 + b[7] * 0 + b[8] * 0 + b[9] * 0 + b[10] * 0

linpred_no_treatment_diabetes_40yo_male |> unname()
```

<!--panache-ignore-start-->
| Term | Value | Comment |
|---|---|---|
| `b[1]` | `* 1` | intercept |
| `b[2]` | `* 0` | thrombectomy = 0 |
| `b[3]` | `* 1` | diabetes = 1 |
| `b[4]` | `* 40` | age = 40 |
| `b[5]` | `* 0` | hypertension = 0 |
| `b[6]` | `* 1` | male = 1 |
| `b[7]` | `* 0` | thrombectomy * diabetes = 0 |
| `b[8]` | `* 0` | thrombectomy * age = 0*40 |
| `b[9]` | `* 0` | thrombectomy * hypertension = 0 |
| `b[10]` | `* 0` | thrombectomy * sex = 0 |

<!--panache-ignore-end-->

But we're still in log-odds space. Let's convert

The letter *g* represents the standard logistic function

$$
  g(x) = \frac{1}{1 + e^{-x}}
$$

Applying this function ensures that the linear component inside the parentheses
of Equation 5.1 gets rescaled to the interval in order to respect the natural
(probability) scale of the binary outcome.

In R, we can make a function to do this for us:

```{r}
#| label: glm

g = \(x) 1 / (1 + exp(-x)) |> unname()

g(linpred_treatment_diabetes_50yo_female)

g(linpred_no_treatment_diabetes_40yo_male)
```

```{r}
#| label: use-margeff-link

grid = data.frame(
    thrombectomy = c(1, 0),
    prior_diabetes = c(1, 1),
    age_at_diagnosis = c(50, 40),
    prior_hypertention = c(0, 0),
    sex = c(0, 1)
)
grid
```

Now, the `{marginaleffects}` function `predictions()` can take care of all this
for us:

```{r}
#| label: with-link

library(marginaleffects)

predictions(th_model, newdata = grid, type = "link")
```

If we don't specify `type = "link"`, the default value is `type = "response"`,
which has already transformed the log-odds to probability (or log-ratio to
count, was the model poisson regression):

```{r}
#| label: inv-link

predictions(th_model, newdata = grid)
```

## Step 5-7: Model-based transformations

To get the best estimate of our estimand, we need to do some post-hoc
transformations to our model coefficients.

Our model is conditional on our sample --- we cannot do anything about that. We
can, however, do some post-hoc weigh on certain participant profiles, so that
our estimate is targeting a specific aspect of the model inference.

Questions that form such discussions can be:

- A particular patient?
- A clinically meaningful scenario?
- The average population?
- Different levels of an interacting variable?

Here, we tackle one scientific question at a time in order to address these
questions:

### RQ 1: Our sample as reality

> "Does the population who got thrombectory have lower probability of death at
> 1-year follow-up after stroke compared to the population who did not get
> thrombectomy?"

Here, we're interested in the sample we have. We're not asking questions whether
these people are representative, biased, or we have any confounds.

The approach we **STRONGLY RECOMMEND NOT TO USE** is to simply report the
coefficients:

```{r}
#| label: raw-coeff

th_model |>
    parameters::parameters(exponentiate=TRUE)
```

Here, the odds ratio is 2.20 ...

> when all interacting variables are at their reference/zero values.

In particular, the interpretation depends on how age is coded/centered and on
the reference categories for diabetes, hypertension, and sex.

> But how intuitive is that? 👀

Instead, lets make predictions on the outcome scale of probabilities using
`{marginaleffects}`.

If we don't supply a `newdata` grid, the `predictions()` function will take all
participants in our sample and compute predictions for them, one by one:

```{r}
#| label: empirical-grid

p = predictions(th_model)
p
```

You can see that the row length is the same as the original dataset:

```{r}
#| label: nrow

nrow(p)
nrow(thrombectomy)
```

A table with `r nrow(p)` rows is not super intuitive and nice to make inference
from. Why can we do about that?

We could either

1. Make a plot to illustrate it all visually, or
2. take the average of all these predictions

**Keeping all information**

Let's start by the first approach and visualize all information:

```{r}
#| label: empirical-data-viz

library(patchwork)

p = predictions(th_model)

# Histogram
p1 = ggplot(p) +
  geom_histogram(aes(estimate, fill = factor(thrombectomy))) +
  labs(x = "Pr(outcome = 1)", y = "Count", fill = "thrombectomy") +
  scale_fill_grey()

# Empirical Cumulative Distribution Function
p2 = ggplot(p) +
  stat_ecdf(aes(estimate, colour = factor(thrombectomy))) +
  labs(x = "Pr(outcome = 1)",
       y = "Cumulative Probability",
       colour = "thrombectomy") +
  scale_colour_grey()

# Combined plots
p1 + p2
```

The plot to the left shows the histogram. I suddenly becomes clear that fewer
patients received thrombectomy than patients who did not.

Indeed, that is correct:

```{r}
#| label: arm-balance

table(thrombectomy$thrombectomy)
```

about `r mean(thrombectomy$thrombectomy)` received thrombectomy.

From the right plot, we can see the cumulative density plot. Here, it seems that
overall... <!-- TODO: Write text here -->

**Marginalizing over the predictions**

We can compute the average for them by using `avg_predictions()`

```{r}
#| label: avg-pred

avg_predictions(th_model)
```

So the average probability of death at the 1-year follow-up is 26.7%.

This is a reasonable estimate for answering the question

> "how many Danes die at 1-year follow-up after stroke?"

However, as we now know that the groups are unbalanced, this is not a good
estimate for the treatment effect! And it doesn't have the granularity to
address more nuanced questions of sub-population level.

Finally, to compute the average for the patients receiving thrombectomy and
those who did not, we can group the average for these conditions

```{r}
#| label: avg-pred-trt

avg_predictions(th_model, by="thrombectomy")
```

Based on the data we have, patients that are offered and successfully manage to
receive thrombectomy have a 30.7% and 21.4% chance of death measured at 1-year
follow-up.

This is the final estimate for our question:

> "Do Danish patients who receive thrombectomy have better probability of
> survival at 1-year follow-up after stroke, compared to patients not receiving
> thrombectomy?"

However, a better estimate might be the **risk difference** or the **risk
ratio**.

$$
  RD = 0.214 - 0.307 = -9.3
$$
or a 9.3% points lower predicted risk of death on the absolute scale.

$$
  RR = \frac{0.214}{0.307} = 0.697
$$
or a 30% lower predicted risk of death on the relative scale.

So we conclude that:

> The model-estimated 1-year mortality was 30.7% in the absence of thrombectomy
> and 21.4% under thrombectomy, corresponding to an un-adjusted absolute risk
> difference of −9.3 percentage points

### RQ 2: Treatment effect of older adults conditional on diabetes status

> "What is the effect of thrombectory in older adults with diabetes, compared to
> older adults without diabetes?"

Here, we're interested specifically in a sub-population. So now, we need to
consider multiple variables (`diabetes` and `thrombectory`) and the proportions
who have this profile in our sample.

```{r}
#| label: grid-oi

datagrid(
    prior_diabetes = 1,
    thrombectomy = c(0, 1),
    age_at_diagnosis=c(80:90),
    model = th_model
)
```

```{r}
#| label: grid-oi-build

p = predictions(
    th_model,
    newdata = datagrid(
        prior_diabetes = 1,
        thrombectomy = c(0, 1),
        age_at_diagnosis=c(80:90)
    )
)

p
```

```{r}
#| label: just-50

predictions(
    th_model,
    newdata = datagrid(
        age_at_diagnosis = 80,
        prior_diabetes = unique,
        thrombectomy = max
    )
)
```

```{r}
#| label: grid-30-to-50

predictions(
    th_model,
    newdata = datagrid(
        age_at_diagnosis=c(80:90),
        prior_diabetes = unique,
        thrombectomy = max)
)
```

```{r}
#| label: marginal-grid-30-to-50

avg_predictions(
    th_model,
    newdata = datagrid(
        age_at_diagnosis=c(80:90),
        prior_diabetes = unique,
        thrombectomy = max)
)
```

```{r}
#| label: marginal-plot

grid_oi <- datagrid(
    age_at_diagnosis=c(80:90),
    prior_diabetes = unique,
    thrombectomy = unique,
    model = th_model
)

p1 = plot_predictions(th_model, newdata = grid_oi, by = "thrombectomy")
p2 = plot_predictions(th_model, newdata = grid_oi, by = c("thrombectomy", "prior_diabetes"))
p1 + p2
```

It becomes clear that the effect of thrombectomy on 1-year follow-up mortality
is negated substantially by the co-occurrence of diabetes in old age.

The effect is substantial for non-diabetic old patient.

```{r}
#| label: final-est-old-diab

avg_predictions(
    th_model,
    by = c("thrombectomy", "prior_diabetes"),
    newdata = datagrid(
        age_at_diagnosis=c(80:90),
        prior_diabetes = unique,
        thrombectomy = unique
    )
)
```

from 42.3% probability of death without treatment to 26.7% for non-diabetic.

from 66.2% probability of death without treatment to 61.6% for non-diabetic.

So our estimates are

### RQ:3 "The average person"

> What is effect of the average person profile in the sample?

Until now, we have considered our observations as meaningful populations that we
want to investigate. However, if we're interested in thrombectomy as an
experimental procedure for any possible candidate patient, we have until now
treated the groups unfairly.

But realistictly, people who gets hit by a stroke and need thrombectomy are not
"any one". These people are already a specific sub-population of Danish people.
Therefore, a perfectly balanced age group, 50% diabetes etc. is extremely
unrealistic and not representative of the Danish population.

First of all, males are nearly 48% propensity of receiving thrombectory or not,
while it's only 38% of females.

```{r}
#| label: sub-groups

xtabs(~ thrombectomy + sex, data=thrombectomy)
```

We have many more males in the treatment group and

```{r}
#| label: sub-diabetes

 xtabs(~ thrombectomy + prior_diabetes, data=thrombectomy)
```

We don't have equal amount of patients with diabetes in the treatment group.

About 10-15% of our sample have diabetes (while the Danish prevalence is
[6,4%](https://diabetes.dk/forskning/viden-om-diabetes/diabetes-i-danmark/)). A
balanced design would have 50% in each group (although this is unrealistic in
reality).

We might be interested in covering some representative and balanced parts of the
population.

```{r}
#| label: exp-balance

age_mean <- mean(thrombectomy$age_at_diagnosis)
age_sd <- sd(thrombectomy$age_at_diagnosis)

grid_oi <- datagrid(
    model = th_model,
    prior_diabetes = unique,
    prior_hypertention = unique,
    sex = unique,
    thrombectomy = unique,
    age_at_diagnosis = c(age_mean-age_sd, age_mean, age_mean+age_sd)
    )

predictions(
    th_model,
    by = "thrombectomy",
    newdata = grid_oi
)
```

This grid is perfectly balanced at our representative selected age: 60, 73, and
86 years old. We have the same amount of diabetes, hypertension, males, and
thrombectomy. This is indeed a completely made-up example. But so is "the
average person"! 👀

::: {.callout-warning}
### "The average person" doesn't exist!

In some samples, "the average person" doesn't exist. I mean this literally 😅

Consider the case where we have a bimodal distribution --- either people love or
hate coriander/cilantro. There is noone who is "meh". So computing the average
person would be the awkward average of "meh", while in reality, people are
either excited or disgusted.

Imagine a scenario where a patient either benefits a lot or not at all from a
treatment. The "average person" would then get "some moderate effect" --- but
that would be completely against the clinical reality experienced by the
clinicians! 👀
:::

```{r}
#| label: data-viz

p1 = plot_predictions(
    th_model,
    newdata = grid_oi,
    condition = "age_at_diagnosis"
)
p2 = plot_predictions(
    th_model,
    newdata = grid_oi,
    condition = c("age_at_diagnosis", "thrombectomy")
)
p3 = plot_predictions(
    th_model,
    newdata = grid_oi,
    condition = c("age_at_diagnosis", "thrombectomy", "prior_diabetes")
)
(p1 + p2) / p3
```

### RQ 4: "What if" 🤔

The research questions was:

> What would be the average probability of death, had all participants received
> thrombectomy, compared to not receiving it?

Finally, we might be interested in the imaginary scenario: what had happened,
given all patients received both the treatment and no treatment, and which would
have resulted in the best probability of being alive at the 1-year follow-up?

We can actually ask this question too:

```{r}
#| label: counterfactual

p = predictions(th_model, variables = list(thrombectomy = 0:1))
nrow(p)
```

Notice how our dataset just doubled its rows. This is no coicidence!

In fact, we now have all observed participants with the "*counterfactual*"
reality that they both got thrombectomy and not. It is called counterfactual
because is its "counter to the fact". The fact was that each patient either did
or did not revieve the thrombectomy. Counter to this fact, we're interested in
the imaginary scenario that they got both!

These predictions are interesting, because they give us a first look at the
kinds of counterfactual (*potentially causal*) queries.

We can ask:

> For each individual in the `Syntheric DanStroke` `{medisr}` sample, what is
> the predicted probability of death in the counterfactual world where they
> receive thrombectory, and where they do not?

Let's marginalize over all participant profiles and get the average probability
of death, had everyone in the sample been in both scenarios of the
treatment/no-treatment.

```{r}
#| label: counterfactual-est

avg_predictions(th_model, variables = list(thrombectomy = 0:1))
```

Notice how different the probabilities are now. Remember that the dataset is not
balanced --- but rather symmetric between the two groups now. The sample is
still biased by propensity for having diabetes, hypertension etc. But the
treatment groups share all these biases now.

```{r}
#| label: counterfactual-plot

plot_predictions(
    th_model,
    newdata=datagrid(
        thrombectomy = c(0,1),
        grid_type="counterfactual"
    ),
    by="thrombectomy"
)
```

This is the best estimate we have until now for the question:

> "Would Danish stroke patients have lower probability of death at 1-year
> follow-up after stroke if everyone had received thrombectomy compared to not
> receiving it?"

However, a better estimate might be the **risk difference** or the **risk
ratio**.

```{r}
#| label: relative-diff

avg_comparisons(
    th_model,
    variables = "thrombectomy"
    )
```

So a risk difference of -0.0317, favoring thrombectomy (negative probability
subtracting the probability of death for patients revieving thrombectomy -
patients not revieving thrombectomy).

::: {.callout}
## The average treatment effect (ATE)

Given certain assumptions (i.e., exchangeability, positivity, non-interference,
and consistency), the risk difference above is actually equivalent to performing
*g*-computation that yields the *average treatment effect*. However, this theory
is beyond the scope of this lecture 😄
:::

The `comparisons()` function above computed the predicted probability of
mortality (`death_1year_followup == 1`) for each observed row of the data in two
counterfactual cases: when `thrombectomy == 1`, and when `thrombectomy == 0`.
Then, it computed the differences between these two sets of predictions.
Finally, it computed the population-average of risk differences.

Instead of *risk differences*, we could compute population-averaged (marginal)
adjusted *risk ratios*:

```{r}
#| label: relative-ratio

avg_comparisons(
    th_model,
    variables="thrombectomy",
    comparison = "lnratioavg", # Calculates adjusted Risk Ratio
    transform = exp # then exponentiates
    )
```

Which gives the estimate of a 0.883 adjusted risk ratio after revieving
thrombectomy.

So we conclude that:

> The model-estimated 1-year mortality was 27.1% in the absence of thrombectomy
> and 23.9% under thrombectomy, corresponding to an adjusted risk ratio of
> 0.883.

## Summary

This session covered going from estimand to estimate.

To communicate clearly about results, we need to apply transformations to our
model coefficients. Predictions can be computed on different scales, depending
on the type of statistical model. The quantity we use should be intuitive and
relevant for clinical practice (going from OR to model-based probabilities)

Analysts can make predictions for different combinations of predictor values, or
grids: empirical, interesting, representative, balanced, or counterfactual. The
grid depends on the research estimatand.

To simplify the presentation of results, it often makes sense to report average
predictions (i.e. marginalizing over all predictions).

<!--
```r
#| label: outcome-scale

library(marginaleffects)

plot_predictions(th_model, condition =  "thrombectomy")

plot_predictions(th_model,
    by =  "thrombectomy"
)

# interesting
datagrid(prior_diabetes = 1, thrombectomy = c(0, 1), model = th_model)

p = predictions(th_model,
  newdata = datagrid(prior_diabetes = 1, thrombectomy = c(0, 1))
)
p

predictions(th_model,
  newdata = datagrid(age = 50, prior_diabetes = unique, thrombectomy = max)
)

# Representative grid
p = predictions(th_model, newdata = "mean")
p

#Balanced grid

# predictions(mod, newdata = datagrid(
#   agecat = unique, incentive = unique, distance = mean
# ))

predictions(th_model, newdata = "balanced")

# Counterfactual grid
p = predictions(th_model, variables = list(thrombectomy = 0:1))
dim(p)

p = data.frame(
  Control = p[p$thrombectomy == 0, "estimate"],
  Treatment = p[p$thrombectomy == 1, "estimate"])

ggplot(p, aes(Control, Treatment)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  geom_point() +
  labs(x = "Pr(outcome=1) when thrombectomy = 0",
       y = "Pr(outcome=1) when thrombectomy = 1") +
  xlim(0, 1) + ylim(0, 1) + coord_equal()

# Aggregation

p = avg_predictions(th_model, newdata = "balanced", by = "thrombectomy")
p

table(thrombectomy$thrombectomy)

avg_predictions(th_model, by = "thrombectomy")

avg_predictions(
    th_model,
    variables = list(thrombectomy = c(0, 1)),
    by = "thrombectomy")


## Visualization



## Marginal predictions
avg_predictions(th_model, by = "thrombectomy")

p1 = plot_predictions(th_model, by = "thrombectomy")
p2 = plot_predictions(th_model, by = c("thrombectomy", "prior_diabetes"))
p1 + p2

plot_predictions(th_model, by = "thrombectomy", newdata = "balanced")

# Conditional predictions
p1 = plot_predictions(th_model,
  condition = "age_at_diagnosis"
)
p2 = plot_predictions(th_model,
  condition = c("age_at_diagnosis", "thrombectomy")
)
p3 = plot_predictions(th_model,
  condition = c("age_at_diagnosis", "thrombectomy", "prior_diabetes")
)
(p1 + p2) / p3

plot_predictions(th_model, condition = list(
    "age_at_diagnosis", "prior_diabetes" = 1, "thrombectomy" = 0
))
```

```r
#| label: outcome-scale

library(marginaleffects)

plot_predictions(beetles_model, condition =  "dose")

slopes(
  beetles_model,
  variables = "dose",
  type = "response"
)

plot_slopes(
  beetles_model,
  variables = "dose",
  condition = "dose",
  type = "response"
)


plot_comparisons(
  beetles_model,
  variables = "dose",
  condition = list("dose" = 1:2),
  type = "response"
)



smoking_model

plot_predictions(smoking_model, by="age")

slopes(
  smoking_model,
  variables = "smoking",
  type = "response"
)

plot_slopes(
  smoking_model,
  variables = "smoking",
  condition = "age",
  type = "response"
)

plot_comparisons(
  smoking_model,
  variables = "smoking",
  condition = "age",
  newdata = doctors,
  type = "response"
)

my_grid <- expand.grid(
  age = levels(doctors$age),
  smoking = levels(doctors$smoking),
  personyears = 10000
)

plot_comparisons(
  smoking_model,
  variables = "smoking",
  condition = "age",
  newdata = my_grid,
  type = "response"
)

plot_comparisons(
  smoking_model,
  variables = "smoking",
  condition = "age",
  newdata = my_grid,
  comparison = "ratio",
  type = "response"
)


marginaleffects::avg_predictions(
  smoking_model,
  variables = "smoking",
  by = "smoking"
)

marginaleffects::avg_predictions(
  smoking_model,
  variables = "smoking",
  by = "age"
)


ATE = marginaleffects::avg_comparisons(
  smoking_model,
  variables = "smoking",
  newdata = doctors
)
ATE

ATT <- marginaleffects::avg_comparisons(
  smoking_model,
  variables = "smoking",
  newdata = subset(smoking == 'smoker')
)
ATT

ATU <- marginaleffects::avg_comparisons(
  smoking_model,
  variables = "smoking",
  newdata = subset(smoking == 'non-smoker')
)
ATU

CATE <- marginaleffects::avg_comparisons(
  smoking_model,
  variables = "smoking",
  by = "age"
)
CATE

```

### What is a marginal effect?

A marginal effect asks:

*How much does predicted probability change when X changes?*

For a continuous variable:

*What happens to predicted probability if X increases by 1 unit?*

For example:

If age increases by 1 year, how much does predicted probability of disease
change?

In a nonlinear model, this is calculated on the probability scale, rather than
the log-odds scale.

With {marginaleffects}:

```r
#| label: marg-eff
#| eval: false

library(marginaleffects)

avg_slopes(model)
```

Conceptually, the output answers: A 1-unit increase in age → Δ predicted
probability

If the result were:

age: 0.012

the interpretation would be:

A one-year increase in age is associated with a 1.2 percentage-point increase in
predicted probability, on average.

Notice the distinction:

0.012 probability

1.2 percentage points

### The effect depends on the patient

Suppose we want to know the effect of age for a particular patient.

We might ask:

What is the marginal effect of age for a 50-year-old non-smoker with normal
blood pressure?

versus:

What is the marginal effect of age for a 75-year-old smoker with high blood
pressure?

These can be different.

We can calculate effects for specified covariate values:

```r
#| label: grid-eff
#| eval: false

slopes(
  model,
  newdata = datagrid(
    age = c(50, 75),
    smoking = c(0, 1),
    blood_pressure = c(120, 160)
  )
)
```

The important concept is not the syntax.

It is:

A marginal effect is conditional on the characteristics of the person for whom
we calculate it.

## Predicted probabilities versus marginal effects

These are different questions.

Predicted probability What is this patient's predicted probability of disease?

```r
#| label: pred
#| eval: false

predictions(
  model,
  newdata = datagrid(
    age = 60,
    smoking = 0,
    blood_pressure = 130
  )
)
```

Marginal effect How much does the predicted probability change if age increases?

```r
#| label: pro-marg
#| eval: false

slopes(
  model,
  newdata = datagrid(
    age = 60,
    smoking = 0,
    blood_pressure = 130
  )
)
```

**Predictions:**

> "What is the risk?"

**Marginal effects:**

> "How much does the risk change?"

Both are useful, but they answer different scientific questions.

## Average marginal effects

Often we do not want the effect for one particular patient.

We want:

What is the average effect in this population?

That is the average marginal effect (AME).

```r
#| label: avg-slopes
#| eval: false

avg_slopes(model)
```

Conceptually:

| Patient 1: effect of X = ...
| Patient 2: effect of X = ...
| Patient 3: effect of X = ...
|        ⋮
| Patient N: effect of X = ...

Average = AME

The interpretation is:

On average, a one-unit increase in X changes predicted probability by Δ
percentage points.

This is often much easier to communicate than an odds ratio.

## Interactions: when the effect depends on something else

Suppose our model includes:

Age × Smoking

The scientific question becomes:

Does the effect of age on disease probability differ between smokers and
non-smokers?

The model might be:

```r
#| label: model1
#| eval: false

model <- glm(
  disease ~ age * smoking + blood_pressure,
  data = data,
  family = binomial
)
```

The temptation is to look at the interaction coefficient and interpret it
directly.

Don't.

The interaction coefficient is on the log-odds scale.

Instead, calculate the marginal effect of age at different smoking statuses:

```r
#| label: int-model1
#| eval: false

slopes(
  model,
  variables = "age",
  by = "smoking"
)
```

Now the result directly addresses the scientific question:

Effect of age among non-smokers Effect of age among smokers

And you can ask:

Is the difference between those marginal effects clinically meaningful?

## Visualizing the interaction

```r
#| label: plotting1
#| eval: false

plot_predictions(
  model,
  condition = c("age", "smoking")
)
```

## Comparing clinically meaningful changes

A one-unit change is not always clinically meaningful.

For example, instead of asking:

What is the effect of a 1-unit increase in blood pressure?

we might ask:

What is the change in predicted probability when blood pressure increases from
120 to 140 mmHg?

For this, use a comparison:

```r
#| label: clin-inform
#| eval: false

avg_comparisons(
  model,
  variables = list(
    blood_pressure = c(120, 140)
  )
)
```

This produces a quantity directly corresponding to the clinical question:

Predicted probability at 140 − Predicted probability at 120 = change in
probability

## Uncertainty A marginal effect is an estimate, not a known quantity.

Therefore we also report uncertainty:

- Estimate
- 95% CI
- p-value

For example:

AME = 0.018 (95% CI 0.010--0.026)

Interpretation:

On average, a one-unit increase in X is associated with a 1.8 percentage-point
increase in predicted probability, with a 95% confidence interval from 1.0 to
2.6 percentage points.

::: {.hidden}
Reading task: Reporting incidence rate ratio or counts

> Should we report the incidence rate ratio or the counts for each group?

What would you prefer and why?

From an absolute perspective, the musquito is the most dangerous animal in the
world! It causes most deaths a year. However, if I were to battle a musquito or
a shark, I would bet my life on the musquito and not the shark! It is relatively
much more dangerous! However, conditional on being in the real-life situation of
facing a musquito or a shark, the former is much more probable.

Now, consider the interpretation of the coefficients and model we just made in
@sec-poisson-interp. How would you inform a patient or a clinican about the
results?

<details>
 <summary>**Click for a tip**</summary>

*Tip*: consider conditioning on `baseline_seizures`.

Is `round((1 - exp(m1_intercept$coefficients[2])) * 100, 1)`% lower the same
when you have 1 seizure every 8 weeks?
What about 100? Would it be a meaningful unit to make national policies?

Now, what if we take the average absolute change:
`(1-exp(m1_intercept$coefficients[2]))*mean(subset(epil,drug=="placebo")$seizure_count)`
when you have 1 seizure every 8 weeks? What about 100? Would it be a meaningful
unit to make national policies?.

 </details>
:::

::: {.hidden}
Extra madness - counterfactuals for clinical inference

This is an extra part (not curriculum material).

Suppose you wanted to know a cutoff for when the drug prevents *at least* three
seizures:

```r
#| label: cut-off-of-when-to-use

grid <- datagrid(
    model=m2_baseline,
    baseline_seizures = seq(0, 100, by = 1),
    age = mean(epil$age), #28 yo
    #V4 = 1, # follow-up
    drug = c("placebo", "progabide")
)

pred <- marginaleffects::predictions(m2_baseline, newdata = grid)

cutoff <- pred |>
    dplyr::select(baseline_seizures, drug, estimate) |>
    tidyr::pivot_wider(
        names_from = drug,
        values_from = estimate) |>
    mutate(
        reduction = placebo - progabide
    )

cutoff_pt <- cutoff |>
  slice_min(abs(reduction - (3)), n = 1)

worsening <- cutoff |>
  slice_min(abs(reduction - (0)), n = 1)

ggplot(cutoff,
       aes(baseline_seizures, reduction)) +
    geom_line(linewidth = 1) +
    geom_hline(yintercept = 3,
               linetype = 2,
               colour = "red") +
    labs(
        x = "Baseline seizures",
        y = "Expected reduction",
        title = "Expected treatment benefit"
    ) +
    theme_bw() +
    geom_label(
        data = cutoff_pt,
        aes(
            label = paste0(
            "Baseline = ", baseline_seizures,
            "\nReduction = ", round(reduction, 2))
        ),
        nudge_y = -1.5
    ) +
    geom_point(aes(baseline_seizures,reduction),
        cutoff_pt,
        colour = "steelblue",
        size = 3)
```

Ask the teacher if you're curious...
:::



### Reading Task: Conditional venom?

Read [this Australian Geographics
blog](https://www.australiangeographic.com.au/blogs/creatura-with-bec-crew/2025/09/funnel-web-spiders-are-australias-deadliest-not-for-our-pets/)
on the Funnel-web spider.



-->
