# Conditioning and polynomial madness {#sec-interaction}

> "x." --- John Doe

```{r}
#| label: setup
#| include: false

#library(tidyverse)
#ddagit782594dalibrary(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)
library(medisr)
library(tidyr)


#hd_data <- readr::read_csv(
#  "https://raw.githubusercontent.com/DanMazJen/medicinsk-statistik/main/data/hd_data.csv"
#)
data(hd_data, package = "medisr")
```

## Pree-class Literature

{{< include /includes/curriculum/_session11.qmd >}}

## Slides

> [**Session 11 slides**](../slides/day11.html)

<div>

```{=html}
<iframe class="slide-deck" src="../slides/day11.html"></iframe>
```

</div>

::: {.callout-note collapse="true"}
## Teacher note

The slides contain speaking notes that you can view by pressing 'S' on the
keyboard.
:::

## Case session 11: ""

At AalboR Statistical Hospital, a new exercise-based cardiac assessment protocol
has recently been introduced following an initial data review suggesting that
patients with lower max heart rate are more likely to have underlying heart
disease. Based on this, some clinicians have started using the max heart rate
during exercise as a quick indicator when evaluating chest pain patients.

However, during a recent internal meeting, a cardiologist raises a concern:

> *"I'm not convinced this applies to all patients. Especially older patients ---
> their heart rate response already decrease. If that's true, we could be
> misinterpreting results in a subgroup of patients.*

You are asked to explore the hospital's patient data to assess whether the
relationship between heart disease and max heart rate appears consistent across
patients, or whether it differs depending on their age.

You are tasked with the decision: *Should this new assessment approach be used
broadly, or are there patient groups where it may be misleading?*

::: {.hidden}
At AalboR Statistical Hospital, a new exercise-based cardiac assessment protocol
has recently been introduced following an initial data review suggesting that
patients with lower max heart rate are more likely to have underlying heart
disease. Based on this, some clinicians have started using the max heart rate
during exercise as a quick indicator when evaluating chest pain patients.

However, during a recent internal meeting, a cardiologist raises a concern:

> *"I'm not convinced this applies to our patients. Especially those who develop
> chest pain during exercise --- their heart rate response might behave
> differently. If that's true, we could be misinterpreting results in a subgroup
> of patients. I want to see this association for myself"*

You are asked to explore the hospital's patient data to assess whether the
relationship between Age and max heart rate --- and the interpretation of max
heart rate more generally --- appears consistent across patients, or whether it
differs depending on exercise-induced angina, sex, or other patient
characteristics.

You are tasked with the decision: *Should this new assessment approach be used
broadly, or are there patient groups where it may be misleading?*
:::

```{r}
#| label: why-this
#| echo: false
#| output: false

hd_data <- hd_data |>
    mutate(
        ExAng = factor(ExAng,
            levels = c(0, 1),
            labels = c("no", "yes")
        ),
        HD = factor(HD,
            levels = c("No", "Yes"),
            labels = c("no", "yes")
        ),
        ChestPain = factor(ChestPain),
        Thal = factor(Thal),
        fasting_blood_glucose = factor(Fbs,
            levels = c(0, 1),
            labels = c("normal", "high")
        )
    )

hd_data |>
  ggplot(ggplot2::aes(Age, MaxHR, color = HD)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x)
```

### Reading Task: Conditional venom?

Read [this Australian Geographics
blog](https://www.australiangeographic.com.au/blogs/creatura-with-bec-crew/2025/09/funnel-web-spiders-are-australias-deadliest-not-for-our-pets/)
on the Funnel-web spider.

## Reading task: The additive and linear assumption

In a standard linear regression model, we make two important assumptions about
how the predictors X relate to the response Y [@james2021introduction]:

1. Additivity assumption

> Each predictor affects Y independently of the other predictors. In other
> words, the effect of $X_j$ ​does not depend on the values of the other X's.

Example: Suppose we predict salary using years of experience and education. The
additivity assumption says that the effect of one additional year of experience
is the same regardless of a person's education level.

2. Linearity assumption

> The effect of a predictor on Y is constant across its range. A one-unit
> increase in $X_j$ is assumed to produce the same change in Y, no matter what
> the current value of $X_j$ is.

Example: If each additional year of experience increases salary by €2,000, the
model assumes this €2,000 increase is the same for someone with 2 years of
experience and someone with 20 years.

We need to bend these assumptions in order to produce an interaction effect. You
will see how we routinely do this using simple tricks of multiplication and
squaring the predictor variables.

### Adding the variable to a regression model is not enough

We will use the `lm()` function for linear model and the `{marginaleffects}`
package to make a quick `{ggplot2}` figure. While we could do this in
`{ggplot2}`, the `{marginaleffects}` package has several tricks for making
regression plots nice and simple to code and inspect.

```{r}
#| label: addition-not-enough

model_additive <-
    lm(MaxHR ~ Age + HD,
    data = hd_data)

library(marginaleffects)

plot_predictions(
    model_additive,
    condition = c("Age", "HD")
)

model_additive |>
    avg_slopes()
```

Wait --- the lines are completely parallel?!

This is exactly the additive and linear assumption: the same effect is constant
for heart disease and for age. We need to allow some flexibility in our model to
consider whether the slopes might change dependent on heart disease.

### Making non-linear terms in linear models

Let's start at the conceptual level.

Suppose we want to predict salary using two variables:

- ($X_1$) = years of experience
- ($X_2$) = education level

We might first think:

- More venom → higher mortality.
- One species → more vulnerable.

But remember the additivity assumption: all species scale similarly on mortality
with increasing venom. Do you think it applies here? I'm not sure...

> The deadliness of venom dose might matter differently depending on the
> species.

That is an interaction. And it is per definition violating the additivity
assumption!

But how do we make a *linear model* capture **non-linear** relationships 👀

Think of this problem as not "breaking" the rules but rather "bending" the
rules. We need some way to give the linear model extra flexibility to capture
relationships that are not additive.

#### 1. What does multiplication mean intuitively?

Think about ordinary multiplication first.

If you have:

$$
  2 \times 3 = 6
$$

and increase 2 to 4:

$$
  4 \times 3 = 12.
$$

The result depends on *both numbers*. This process is linear *conditioned* on
the other number being the same.

Now imagine:

$$
  X_1 \times X_2.
$$

If ($X_1$) changes, the product changes --- but *how much it changes depends on
($X_2$)*.

For example:

::: {#tbl-multiplication-table}
  | Venom dose ($X_1$) | Education ($X_2$) | ($X_1X_2$) |
  | -----------------: | ----------------: | ---------: |
  |                  2 |                 1 |          2 |
  |                  3 |                 1 |          3 |
  |                  2 |                 3 |          6 |
  |                  3 |                 3 |          9 |
:::

Notice what happens when experience increases from 2 to 3:

- If education = 1, the product increases by **1**.
- If education = 3, the product increases by **3**.

So multiplication naturally creates a situation where:

> The effect of one variable depends on the value of the other variable.

That is exactly what we mean by an **interaction**.

To allow this, we create a new variable:

$$
  γ = X_1X_2
$$
our new variable, γ (gamma), *multiply the two variables together*.

> Notice that multiplication is non-linear now!

When we say *linear regression*, "linear" does *not* mean that we are forever
doomed to model only linear and additive phenomena.

Instead, it means that the model *coefficients* are linear --- the numbers that
the regression estimates.

You can think of γ as we create a *new predictor* --- the variable is producing
another linear coefficient. The regression can then use this new variable just
like any other predictor.

This new variable consider the *conditional relationship* between two values. If
$β_1$ is small, the value might be small independent of $β_2$. But if both $β_1$
and $β_2$ is large, the value might always be large (see
@tbl-multiplication-table)

There are therefore two different ideas:

- Additive models: Where the effect of one predictor does not depend on the
  other predictor.
- Interaction models: Where he effect of one predictor can depend on the other
  predictor.

Multiplication is useful because it naturally creates this dependency.

But how do we do that in R(eal life)?

### Adding an interaction

We can literally make a new variable called `age_by_hd` using `{dplyr}`:

```{r}
#| label: interaction1

model_interact_manual <- hd_data |>
    mutate(
        HD_num = ifelse(HD == "yes", 1, 0),
        age_by_hd = Age * HD_num
    ) |>
    lm(MaxHR ~ Age + HD + age_by_hd,
    data = _)

coef(model_interact_manual)
```

However, this is such a common routine in R that we don't need to make the new
variable. Instead, we can indicate this multiplication by using the following
notation:

`Age:HD`

We will wait with interpreting the coefficients till the @sec-coef-interp. For
now, confirm that these numbers are identical to the model below.

```{r}
#| label: interaction1-r-trick

model_interact <-
    lm(MaxHR ~ Age + HD + Age:HD,
    data = hd_data)

coef(model_interact)
```

Finally, let's take a look at the interaction plot:

```{r}
#| label: plot-interaction

plot_predictions(
    model_interact,
    condition = c("Age", "HD")
)
```

Cool! Now, the lines are definitely *not* parallel! This is in fact a trick for
spotting interactions in plots.

::: {.callout-caution}
## Many ways of doing the same thing

You will also see other people using this notation `Age * HD`. This expands to
`Age + HD + Age:HD`. However, we recommend using the `Age:HD`, as adding more
variables and `*` notation quickly explodes.

```{r}
#| label: interaction1-r-trick2

model_interact <-
    lm(MaxHR ~ Age * HD,
    data = hd_data)
```
:::

### Interpreting an interaction estimate {#sec-coef-interp}

In order to interpret the regression model, we need to recover the regression
coefficient. From there, we will have to plug in numbers depending on what we're
interesting in estimating. This section goes through that process step by step.

#### Step 1: Recovering the regression equation

The first thing you can do to inspect a regression model is to use the
`summary()` function:

```{r}
#| label: summary-int1

model_interact |>
    summary()
```

Note that we are missing the confidence intervals. A better tool is
`parameters::model_parameters()`:

```{r}
#| label: params

model_params <- model_interact |>
    parameters::model_parameters()
model_params
```

This is great --- we could make it a bit simpler and only look at the estimates
(here names `_est`):

```{r}
#| label: model-coeff-interp
#| echo: false

model_params <-
    data.frame(
        intercept_est = model_params[1,2],
        age_est = model_params[2,2],
        hd_est = model_params[3,2],
        age_hd_interact_est = model_params[4,2]
    )

model_params
```

You might wonder how this relates to the regression equation. Let's plug the
values in:

$$
  \bar{Y} = 213.93 − 1.06 * age − 53.84 * hd + 0.689 * age\_by\_hd
$$

where:

- age = age in years
- hd = 0 (no heart disease) or 1 (heart disease)

And there we have the regression equation 🎉

#### Step 2: Write separate equations for each group

For this case, we would want the estimate for a heart disease vs. no heart
disease.

Recall that no heart disease was coded `hd = 0`:

$$
  \bar{Y} = 213.93 − 1.06 * age − 53.84 * ZERO + 0.689 * ZERO
$$
$53.84 * 0$ cancels out --- so does 0.689 \* ZERO.

$\bar{Y} = 213.93 − 1.06(age)$

**Interpretation:**

For people without heart disease, each additional year of age is associated with
about 1.06 units lower predicted outcome.

If we now look at the heart disease group (`hd = 1`) we get:

$ \bar{Y} = 213.93 − 1.06(age) − 53.84 + 0.689(age) $

because $53.84 * 1$ is 53.84 and $0.689 * age × 1$ is $0.689(age)$.

We can combine the terms:

$\bar{Y} = 160.09 − 0.371(age)$

**Interpretation:**

For people with heart disease, each additional year of age is associated with
only about 0.37 units lower predicted outcome. The age effect is less negative
because of the positive interaction term.

#### Step 3: Plug in specific ages

**Age = 40**

*No heart disease:* $213.93 − 1.06(40) = 171.67$

*Heart disease:* $160.09 − 0.368(40) = 145.37$

*Difference:* $145.37 − 171.67 = −26.30$

> At age 40, heart disease is associated with about 26 units (BPM max heart
> rate) lower predicted outcome.

**Age = 70**

*No heart disease:* $213.93 − 1.06(70) = 140.00$

*Heart disease:* $160.09 − 0.368(70) = 134.34$

*Difference:* $134.34 − 140.00 = −5.66$

> At age 70, heart disease is associated with only about 6 units (BPM max heart
> rate) lower predicted outcome.

#### Step 4: Interpret the interaction

The interaction coefficient is: **0.689**

This means that the effect of heart disease becomes 0.689 units (BPM max heart >
rate) less negative for every additional year of age.

In other words: The difference between the heart disease and no-heart-disease
groups shrinks by about 0.689 units for each additional year of age.

We can again plug the values in to examine this:

```r
Age HD=0    HD=1    Difference (HD−No HD)
40  171.67  145.37  -26.30
50  161.11  141.69  -19.41
60  150.54  138.01  -12.53
70  139.98  134.32  -5.66
```

Often, graphics are easier to interpret than tables or coefficients. Let's try
to visualize the difference between max heart rate and having a heart disease
when it change across age.

```{r}
#| label: delta-hr-of-hd-across-age
#| warning: false

plot_slopes(model_interact, variables = "HD", condition = "Age") +
    geom_hline(yintercept = 0, linetype = "dotted") +
    labs(
        y = "expected difference (MaxHR | HD)"
    )
```

```{r}
#| label: ekstra-for-margeff-later
#| include: false

slopes(model_interact,
  variables = "HD",
  newdata = datagrid(Age = range))

slopes(model_interact,
  variables = "HD",
  newdata = datagrid(Age = range),
  hypothesis = "b2 - b1 = 0"
)
```

## Symmetry of the linear interaction

Consider for example the max heart rate and age problem. The interaction there
has two equally valid phrasings.

- How much does the influence of age (on max heart rate) depend on whether the
  patient has a heart disease?
- How much does the influence of having a heart disease (on max heart rate)
  depend on age?

These sound like different questions, but in a linear model they are
mathematically identical.

The original model is

$$
  \text{max heart rate}_i \sim α + β_1age + β_2HD + β_3age\_by\_HD + ε\\
$$

For now, let's make the model more compressed at the expense of nesting some
parameters within a new variable γ (gamma). We will not consider the noise, ε,
so we will only deal with the expected value, μ.

$$
  \mu_i = \alpha + \gamma_i \,\text{age}_i + \beta_2 \,\text{HD}_i
$$

where

$$
  \gamma_i = \beta_1 + \beta_3 \,\text{HD}_i.
$$

Here, $\gamma_i$ is the effect of age for a patient with a given heart-disease
status. The interaction allows this effect to differ between patients with and
without heart disease.

Substituting the second equation into the first gives

$$
  \mu_i = \alpha + (\beta_1 + \beta_3 \,\text{HD}_i)\,\text{age}_i
          + \beta_2 \,\text{HD}_i.
$$

The substitution makes explicit where the interaction enters the model.

Expanding:

$$
  \mu_i = \alpha + \beta_1 \,\text{age}_i + \beta_3 \,\text{age}_i
          \times \text{HD}_i + \beta_2 \,\text{HD}_i.
$$

The interaction term is the same regardless of which variable we initially
treated as the predictor.

Now collect the terms containing $\text{HD}_i$:

$$
  \mu_i = \alpha + \beta_1 \,\text{age}_i
          + (\beta_2 + \beta_3 \,\text{age}_i)\,\text{HD}_i.
$$

This is the same model, but now the coefficient of $\text{HD}_i$ is allowed to
depend on age.

Define

$$
  G_i = \beta_2 + \beta_3 \,\text{age}_i.
$$

Then

$$
  \mu_i = \alpha + \beta_1 \,\text{age}_i + G_i\,\text{HD}_i.
$$

The quantity $G_i$ plays the same role for the heart-disease effect that
$\gamma_i$ played for the age effect.

### Key takeaway

The interaction coefficient $\beta_3$ simultaneously describes

- how the effect of age changes across HD/non-HD, and
- how the effect of HD changes across levels of age.

In a linear model, interactions are symmetric. The choice of which variable is
called the "predictor" and which is called the "moderator" is often a matter of
interpretation rather than mathematics.

## Continuous interactions

Until now, we've use one continuous and one dichotomous variable.

> What if we have two continuous variables that interaction?

```{r}
#| label: cont-interaction

pain_data <- expand_grid(
  dose = 0:10,
  pain = c(0,1,2)
) |>
  mutate(
    relief = 1.6 + 0.4*dose + 0.4*pain + 0.9*dose*pain + rnorm(n(), 0,1)
    )

m <- lm(relief ~ dose * pain, data = pain_data)

summary(m)
```

```{r}
#| label: triptic-plot

pain_data |>
  mutate(pain_group = factor(pain)) |>
  ggplot(aes(dose, relief)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(
    ~ pain_group,
    labeller = as_labeller(c(
      `0` = "No pain",
      `1` = "Mild pain",
      `2` = "Severe pain"
    ))
  ) +
    labs(
        x = "Dose (mg)",
        y = "Subjective relief (0 - 25)"
    )
```

## Polynomials

Recall the linearity assumption [@james2021introduction].

Everything are not linear, is it? Consider for example the height of people:

```{r}
#| label: height
#| fig-cap: "Demographic data from Kalahari !Kung San people collected by Nancy Howell "
#| echo: false

height_data <- medisr::height_data

height_data |>
    ggplot(aes(age,height)) +
    geom_point()
```

As a human, you increase you height quite dramatically in childhood --- but this
inevitably comes to an end later in life. So how can we model this non-linear
development?

The answer is surprisingly simple --- we can compute a new variable ϕ (phi) that
is the original variable squared $(i.e. ϕ = X_1^2)$. If we do this for `age`, we
can get $ϕ = age^2$. But nothing stops us from taking the cube or above! This is
called a polynomial and looks like this:

<!-- panache-ignore-start-->
| Polynomial degree | Regression model |
|-:|---|
| 1 | $y = \beta_0 + \beta_1 x$ |
| 2 | $y = \beta_0 + \beta_1 x + \beta_2 x^2$ |
| 3 | $y = \beta_0 + \beta_1 x + \beta_2 x^2 + \beta_3 x^3$ |
| 4 | $y = \beta_0 + \beta_1 x + \beta_2 x^2 + \beta_3 x^3 + \beta_4 x^4$ |
<!-- panache-ignore-end-->

```{r}
#| label: lin-quar-fit
#| warning: false
#| fig-cap: "Demographic data from Kalahari !Kung San people collected by Nancy Howell"
#| fig-subcap:
#|   - "Linear model"
#|   - "Quardratic model"
#|   - "Cubic model"
#|   - "Quadratic model of individuals < 25 years old"
#|   #| echo: false

height_data |>
    ggplot(aes(age,height)) +
    geom_point() +
    geom_smooth(method = "lm")

height_data |>
    ggplot(aes(age,height)) +
    geom_point() +
    geom_smooth(method = "lm", formula = y ~ x + I(x^2)) # I is function for "as-is" to make math operation as is.

height_data |>
    ggplot(aes(age,height)) +
    geom_point() +
    geom_smooth(method = "lm", formula = y ~ x + I(x^2) + I(x^3)) # I is function for "as-is" to make math operation as is.

height_data |>
    dplyr::filter(age<25) |>
    ggplot(aes(age,height)) +
    geom_point() +
    geom_smooth(method = "lm", formula = y ~ x + I(x^2)) # I is function for "as-is" to make math operation as is.
```

As you can see, the models definitely fit better than the

::: {.hidden}
## The danger of

```{r}
library(sjPlot)
data(efc, "sjPlot")

m <- lm(
  neg_c_7 ~ c12hour + e16sex + c12hour:e16sex,
  data = efc
)

# fit model with 3-way-interaction
fit <- lm(neg_c_7 ~ c12hour * barthtot * e16sex, data = efc)

# select only levels 30, 50 and 70 from continuous variable Barthel-Index
sjPlot::plot_model(fit, type = "pred", terms = c("c12hour", "barthtot [30,50,70]", "e16sex"))
```

where:

neg_c_7 = caregiver burden (outcome) c12hour = hours of care provided sex =
caregiver sex

Then ask the two interaction questions:

How much does the association between caregiving hours and burden depend on
caregiver sex?

and

How much does the difference between male and female caregivers depend on
caregiving hours?

Students usually perceive these as different questions, but the interaction
coefficient is the same parameter in both cases.

```{r}
#| label: continuous

library(marginaleffects)

marginaleffects::slopes(
  fit,
  variables = "c12hour",
  by = "e16sex"
)

#marginaleffects::avg_slopes(
#  fit,
#  variables = "c12hour",
#  by = list("c12hour", "barthtot" = c(30,50,70))
#)

marginaleffects::plot_slopes(
  fit,
  variables = "c12hour",
  condition = list("c12hour", "barthtot" = c(30,50,70))
)

marginaleffects::plot_predictions(
  fit,
  variables = "c12hour",
  condition = list("c12hour", "barthtot" = c(30,50,70))
)
```

```{r}
marginaleffects::comparisons(
  fit,
  variables = "e16sex",
  by = "c12hour"
)
```

```{r}
#| label: hidden-stuff

set.seed(1)

drug_data <- expand_grid(
  dose = 0:10,
  receptor = seq(1, 10, by = 1)
) |>
  mutate(effect = .2*dose + .1*receptor +
         .10*dose*receptor +
         rnorm(n(), 0, 1))

m <- lm(effect ~ dose * receptor, data = drug_data)

summary(m)

drug_data |>
  mutate(receptor_group = cut(receptor, 3)) |>
  ggplot(aes(dose, effect)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ receptor_group)

#########


set.seed(123)

dat <- expand_grid(
  frailty = c(-1, 0, 1),
  dose    = c(-1, 0, 1),
  id      = 1:9
) %>%
  mutate(
    change_pf = 35 +
      25 * dose -
      18 * dose^2 -
      8  * frailty +
      5  * dose * frailty -
      7  * dose^2 * frailty +
      rnorm(n(), 0, 7)
  )

# Fitted curves at -1, 0, +1 SD frailty
pred <- expand_grid(
  dose = seq(-1.5, 1.5, length.out = 100),
  frailty = c(-1, 0, 1)
) %>%
  mutate(
    change_pf = 35 +
      25 * dose -
      18 * dose^2 -
      8  * frailty +
      5  * dose * frailty -
      7  * dose^2 * frailty,
    frailty = factor(
      frailty,
      levels = c(-1, 0, 1),
      labels = c("Low frailty", "Mean frailty", "High frailty")
    )
  )

ggplot(dat, aes(dose, change_pf, colour = factor(frailty))) +
  geom_point(size = 2.5) +
  geom_line(
    data = pred,
    aes(dose, change_pf, colour = frailty),
    linewidth = 1.2
  ) +
  labs(
    x = "Chemotherapy dose intensity (SD)",
    y = "Change in physical functioning",
    colour = "Frailty"
  ) +
  theme_minimal(base_size = 13)
```
:::
