# A Poissonous counting trick {#sec-poisson}

> "Choose to be optimistic, it feels better." --- Dalai Lama

> "Choose to model positive outcomes, it feels better for count outcomes." ---
> Daniel Skak Mazhari-Jensen

```{r}
#| label: setup
#| include: false

#library(tidyverse)
#ddagit782594dalibrary(tidyverse)
library(ggplot2)
library(MASS)
library(readr)
library(tidyr)
library(broom)
library(marginaleffects)
library(patchwork)
library(performance)
library(dplyr)

#hd_data <- readr::read_csv(
#  "https://raw.githubusercontent.com/DanMazJen/medicinsk-statistik/main/data/hd_data.csv"
#)

# https://arelbundock.com/posts/quarto_figures/
knitr::opts_chunk$set(
    out.width = "70%", # enough room to breath
    fig.width = 6,     # reasonable size
    fig.asp = 0.618,   # golden ratio
    fig.align = "center" # mostly what I want
)
```

## Pree-class Literature

- [Todays dataset (Epilepsy Clinical Trial)](../exercises/data.qmd)
- [Compendium]()
- Chapter 4.6 (pages 164----171) in [Introduction to statistical learning with R
  2nd edition](https://www.statlearning.com/) (click ["Download ISL with
  R"](https://hastie.su.domains/ISLR2/ISLRv2_corrected_June_2023.pdf.download.html))

# Slides

> [**Session 12 slides**](../slides/day12.html)

<div>

```{=html}
<iframe class="slide-deck" src="../slides/day12.html"></iframe>
```

</div>

::: {.callout-note collapse="true"}
## Teacher note

The slides contain speaking notes that you can view by pressing 'S' on the
keyboard.
:::

## Case session 12: ""

At PhaRmacy Inc., excitement is growing over a new anti-seizure medication,
progabide. A randomized clinical trial has just finished, and the lead
investigator walks into the weekly research meeting with a smile.

"The treatment group had fewer seizures."

The room goes quiet.

A statistician leans forward.

"Fewer... according to what model?"

Looking at the raw data raises more questions than answers. Some patients
experienced only a handful of seizures during the study, while others had
dozens. A few seem very different from the rest. The distribution certainly
doesn't look like the outcomes you're used to analyzing.

Your task is to help the research team answer a deceptively simple question:

How should seizure counts be analyzed, and what can we conclude about the effect
of progabide after accounting for differences in patients' baseline seizure
frequency and age?

As you work through the analysis, consider whether the data behave the way your
chosen model expects---and what you would do if they don't.

<!--
At PhaRmacy Inc., a new anti-seizure medication, *progabide*, has been
introduced as a potential treatment for patients with refractory epilepsy. Early
results from the clinical trial appear promising, and some neurologists have
suggested that the drug should become part of routine care because patients
receiving progabide seem to experience fewer seizures than those receiving
placebo.

However, during a recent clinical meeting, an epileptologist raises a concern:

> "I'm not convinced that looking at the average reduction tells the whole
> story. Our patients begin treatment with very different seizure burdens.
> Reducing three seizures could be life-changing for one patient but barely
> noticeable for another. Before recommending this drug, I want to know which
> patients are actually likely to benefit."

The clinicians agree that a reduction of at least three seizures during the
four-week follow-up represents the minimum clinically meaningful improvement. At
the same time, they recognize that seizure counts have unusual statistical
properties:

1. First of all, patients cannot experience "*negative seizures*", so negative
   values are not meaningful,
2. Many patients have relatively few seizures, while others experience dozens
   during the same period,
3. There is no obvious upper limit for seizure counts, so it doesn't make sense
   to talk about probabilities of seizures for a day or month.

You are asked to analyze the trial data to investigate how seizure frequency is
associated with treatment (progabide vs. placebo) while accounting for each
patient's baseline seizure frequency and age. Your goal is to determine whether
progabide provides a clinically meaningful reduction in seizures and whether
some patients are expected to benefit more than others.

You are tasked with the decision: Should progabide be recommended broadly for
patients with epilepsy, or should treatment decisions depend on each patient's
baseline seizure burden and expected reduction in seizure frequency?
-->

## Review of Logistic regression

See session 10 on logistic regression by Carsten.

Logistic regression uses the binomial distribution:

$$
  outcome \sim Binomial(N,p)
$$
where *N* is the trial count and *p* is the probability, which we are interested
in estimating with out model.

- We use the binomial distribution with *logit link* to model the outcome as the
  probability of event A.
- These are discrete counts, but we are interested in the overall probability of
  getting outcome A.
- Only two outcomes: 0 or 1 (heads or tails, dead or alive, left or right)
- The logit link transform our outcome variable to a linear space called
  log-odds that goes from $-\infty$ to $\infty$.
- The logistic function transforms our data back to the outcome space
  (probabilities that goes from 0---1).

## Reading Task: When the odds are low, go Poisson!

While the binomail distribution is great when *N* is known (trial count), there
exists studies where you don't know the upper limit (counting fish, selling a
product, etc.) --- what is the upper bound of fish in a lake we haven't fully
accounted for?

> If we don't know *N*, we cannot use the binomial!

Luckily, there's a trick for this. When N becomes huge and *p* becomes really
small, the binomial distribution takes a special shape --- the Poisson
distribution.

Explained mathematically:

The expected value of a binomial distribution is just *Np* (probability scale),
and its variance $Np(1-p)$. But when *N* is very large and *p* is very small,
the variance and expected value is the same.

E.g., you have 1 in 1000 probability of getting a price.

```{r}
#| label: binom-to-poisson

y <- rbinom(1e5, 1000, 1/1000)
c( mean(y), var(y))
```

As a result, the distribution only has one parameter, λ (pronounced *lambda*).

$$
  y_i \sim Poisson(λ)
$$

This solves several issues:

- We can now model counts and probabilities (actually, *rates*, but more on this
  later) without being in control of the experiment and knowing the finite trial
  count
- We can model events with extremely low probability in counts per sample
  instead of probability per trial
- We only need one parameter for both the mean and the variance, effectively
  saving a degree of freedom.

But how can we do this in practice?

### Which link?

As a link function, we often use log.

$$
  y_i \sim Poisson(λ_i) \\
  log(λ_i) = α + β(x_i - \bar{x})
$$
Why log?

```{r}
#| label: why-log
#| echo: false
#| fig-subcap:
#|   - "While the log-linear scale can take any real value, the ouctome scale is only positive real numbers."

# fig-cap: "Charts"
#   - "Second"


# Poisson regression:
# log(mu) = beta0 + beta1*x

dat <- data.frame(
  x = seq(-5, 5, length.out = 200)
) %>%
  mutate(
    log_mu = -1 + 0.5*x,       # can be negative
    mu = exp(log_mu),           # always positive
  ) %>%
  pivot_longer(
    cols = c(log_mu, mu),
    names_to = "scale",
    values_to = "value"
  ) %>%
  mutate(
    scale = recode(
      scale,
      log_mu = "Log-linear (Poisson) scale:\nlog(E(Y))",
      mu = "Outcome scale:\nE(Y)"
    )
  )

ggplot(dat, aes(x = x, y = value)) +
  geom_line(
    linewidth = 1.3,
    color = "#2166AC"
  ) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "grey40"
  ) +
  facet_wrap(~scale, scales = "free_y") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Why Use a Log Link in Poisson Regression?",
    subtitle = "Negative values are allowed on the log scale, but exponentiation guarantees positive counts",
    x = "Predictor (x)",
    y = "Outcome (y)"
  ) #+
 # annotate(
 #   "text",
 #   x = -2.5,
 #   y = -2.8,
 #   label = "log(μ) can be negative\n(no restriction)",
 #   size = 5,
 #   color = "black"
 # ) +
 # annotate(
 #   "text",
 #   x = 2,
 #   y = 0.25,
 #   label = "μ = exp(log(μ))\n> 0 always",
 #   size = 5,
 #   color = "black"
 # ) +
 # annotate(
 #   "segment",
 #   x = -1,
 #   xend = 1,
 #   y = -1.5,
 #   yend = 0.2,
 #   arrow = arrow(length = unit(0.2, "cm")),
 #   color = "firebrick"
 # ) +
 # annotate(
 #   "segment",
 #   x = 0,
 #   xend = 1,
 #   y = 0.4,
 #   yend = 1.5,
 #   arrow = arrow(length = unit(0.2, "cm")),
 #   color = "firebrick"
 # )
```

Note that the log-link has the same flaws and benefits from the binomial
distribution.

Pros:

- ensures positive values (for prediction)

Cons:

- implies exponential relationship with predictors
  - Grow very rapidly --- therefore, very sensitive to range
  - it cannot keep being exponential!

## Data Example

<!-- panache-ignore-format-start -->
<!-- Heres a DAG:
          Age
           │
           ▼
Baseline seizures ─────► Outcome (Seizures)
        │                  ▲
        │                  │
        ▼                  │
   Treatment ──────────────┘
        ▲
        │
 Unmeasured severity
-->
<!-- panache-ignore-format-end -->

Before doing *anything* with statistics and code, we should try imagining how
the data was generated.

### Exercise: Discuss the causal mechanisms of progabide

Discuss how progabide might work and how `baseline_seizures` might tie into
this.

<details>
<summary>**Click for a potential solution**</summary>

```{r}
#| label: dag
#| echo: false

dag <- dagitty::dagitty("
dag {

severity -> gaba
severity -> baseline

gaba -> seizures
gaba -> baseline

baseline -> seizures
drug -> gaba

}
")

dag |>
    ggdag::tidy_dagitty() |>
    ggdag::ggdag(text = FALSE) +
    ggdag::geom_dag_text(size = 2.5) +
    theme_void()

dagitty::adjustmentSets(dag,
    exposure = "drug",
    outcome = "seizures",
    effect="total"
)
```

While baseline is not a coufounding, we'll condition on it anyway to:

- increases precision (reduces residual variance)
- account for chance imbalance at baseline (since this is a small sample size)
- improve power

*This is called precision adjustment or covariate adjustment, not confounding
adjustment.*

</details>

Let's explore the dataset a bit --- but before we do, I'll do some tidying of
the dataset using `{dplyr}` for wrangling. I'm not gonna explain this bit, as
you can go back to session 8 by Rasmus for working with `{dplyr}` for wrangling.

```{r}
#| label: pre-wrangle

data(epil)

epil <- epil |>
    rename(
        seizure_count = y,
        drug = trt,
        baseline_seizures = base,
        trial_end = V4,
        log_base = lbase,
        log_age = lage
  ) |>
    #filter(period < 2 | period > 3) |>
    filter(period > 3) |>
    mutate(
        seizure_count = as.integer(seizure_count),
        period = factor(period,
            levels = c(4),
            labels = c("week_4")
            )
    )
```

And let's looks at what we have:

```{r}
#| label: just-looking

epil |>
    tibble::as_tibble()

epil |>
    count(drug)
```

So 59 patients in total with pretty balanced groups.

Now, let's look at a histogram of the `baseline_seizures` and `seizure_count`:

```{r}
#| label: explore1

seizure_lims <- range(c(epil$baseline_seizures, epil$seizure_count))

baseline_data_plot <- epil |>
    ggplot() +
    geom_histogram(
        aes(x=baseline_seizures),
        bins=30
    ) +
    coord_cartesian(xlim = seizure_lims)

posttrial_data_plot <- epil |>
    ggplot() +
    geom_histogram(
        aes(x=seizure_count),
        bins=30
    ) +
    coord_cartesian(xlim = seizure_lims)

baseline_data_plot + posttrial_data_plot
```

By looking at the data, it's deficult to really see whether the average
`seizure_count` decreases, but some of the extreme cases do seem to shrink.

We should also consider the relationship between `baseline_seizures` and
`seizure_count`:

```{r}
#| label: exploration

epil |>
    ggplot(aes(baseline_seizures,seizure_count,color=drug)) +
    geom_point(alpha=0.6) +
    geom_smooth(method="loess",se=FALSE) +
    theme_bw()
```

It definitely looks like there's an association between `baseline_seizures` and
`seizure_count`, so in accordance with our *medical scientific model*, we better
include this in the *statistical model* to account for the variance.

## Discussion task: Gaussian vs Poisson for seizure counts:

Discussion the following facts with your neighbor:

If we model the mean and variance of the seizure count, we would expect that
taking the mean and variance of the dataset and feeding it into a distribution
would closely resemble the data. Let's try that with Gaussian distribution and a
Poisson distribution:

```{r}
#| label: gauss-vs-poisson

mu <- mean(epil$seizure_count) # 7.305085
sigma <- sd(epil$seizure_count) # 9.649467

set.seed(363461174)
dist_draws <- data.frame(
    gaussian = rnorm(1000,mu,sigma),
    poisson = rpois(1000,mu),
    negbin   = rnbinom(1000, mu = mu, size = mu^2 / (sigma^2 - mu)) # Variance = mu + mu^2/size - solve for size!
    ) |>
    tidyr::pivot_longer(
        cols = c("gaussian","poisson","negbin"),
        names_to = "dist",
        values_to = "value"
    )

dist_draws |>
    ggplot() +
    geom_histogram(
        aes(x=value),
        bins=30
    ) +
    facet_wrap(~dist) +
    labs(x="seizures")
```

What do you notice? Compare the plots to the observed data we plotted just
above.

- Which best captures the data?
- Which makes the worst fit?
- What else do you notice?

<!--
```{r}
#| label: test
#| eval: false

fit <- glm(y ~ trt * base, V4, # + lage
    family = poisson,
    data = epil
)

predict(fit) |>
    data.frame(
        mu_hat = _,
        mu_obs = epil$y
    ) |>
    ggplot(aes(x=exp(mu_hat))) +
    geom_histogram(bins = 30)

predict(fit, type = "response") == exp(predict(fit))

epil |>
  mutate(
    predicted = predict(fit, type = "response")
  ) |>
  ggplot(aes(predicted, y)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, linetype = 2, colour = "red") +
  labs(
    x = "Predicted count",
    y = "Observed count"
  ) +
  theme_bw()
```
-->

## Proposing a model

First, we should propose a model. Looking at out

$$
  seizure\_count_i \sim Poisson(λ_i) \\
  log(λ_i) = α + β_1 baseline\_seizures + β_2 drug + β_3 baseline\_seizures*drug \\
$$

1. The model says that we believe `seizure_count` is Poisson distributed
2. We use a log link function
3. We expect an intercept, α
4. A covariate `baseline_seizures`
5. An effect of `drug`
6. An interaction between `baseline_seizures` and `baseline_seizures`

## Fitting a model

Let's try fitting a simple model first: an intercept-only model.

```{r}
#| label: intercept-only-fit

m1_intercept <- glm(
    seizure_count ~ drug,
    family = poisson(link="log"),
    data = epil
)

```

Great! Now, let's explore the content of it.

```{r}
#| label: intercept-only-fit-interp

summary(m1_intercept)
```

Yeah, we did it! But... What does the coefficients actually mean? Did we come
any closer to answering our medical question: did the drug reduce epilepsy?

## GLMs are notoriously difficult to interpret

Let's have a look at the coefficients:

```{r}
#| label: IO-coef

coef(m1_intercept)
```

What does it mean?

Unless you're a chatbot, I'm not sure you can read anything major from these
coefficients. It's because they're still in log-link space!

Let's convert them back from log-ratios to a simple odds-ratio. We do this by
exponentiating the value:

$$
  e^{cefficients}
$$

```{r}
#| label: inference

broom::tidy(m1_intercept,
     exponentiate = TRUE,
     conf.int = TRUE)
```

Now, it's more like it.

The intercept reads `r exp(m1_intercept$coefficients[1])`, which is the expected
`seizure_count` for the placebo group (and the average as we would expect!)
Confirm this for yourself using the following code:
`mean(subset(epil,drug=="placebo")$seizure_count)`.

Next, it's the `β1drug` `r exp(m1_intercept$coefficients[2])`, which is the
expected rate ratio in `seizure_count` when being on progabide for 8 weeks,
compared to placebo. Since it's less than 1, the expected seizure count is
lower.

In fact, `r round((1 - exp(m1_intercept$coefficients[2])) * 100, 1)`% lower on
average. That's much easier to interpret than a coefficient on the log scale.

Now, it's more like it.

The intercept reads `r m1_intercept$coefficients[1]` which is the expected
`seizure_count` (and close to the average!)

Next, it's the `β1drug` `r m1_intercept$coefficients[2]` which is the expected
odds-ratio in `seizure_count` when being on progabide for 8 weeks. Since it's
less than 1, the odds are lower.

## Reporting odds-ratios or counts

From an absolute perspective, the musquito is the most dangerous animal in the
world! It causes most deaths a year. However, if I were to battle a musquito or
a shark, I would bet my life on the musquito and not the shark! It is relatively
much more dangerous! However, conditional on being in the real-life situation of
facing a musquito or a shark, the former is much more probable.

```{r}
#| label: predictions

epil$pred <- predict(
    m1,
    type="response"
)

ggplot(epil,
       aes(pred,seizure_count))+
    geom_point()+
    geom_abline(slope=1,
                intercept=0,
                color="red")+
    theme_bw()
```

## Fitting a multiple Poisson regression

But wait - we also wanted to condition on `baseline_seizures`. Let's to that:

```{r}
#| label: model-fit

m2_baseline <- glm(
    seizure_count ~ drug * baseline_seizures,
    family = poisson(link="log"),
    data = epil
)

m3_full <- glm(
    seizure_count ~ drug * baseline_seizures * age,
    family = poisson(link="log"),
    data = epil
)

m1 <- m2_baseline

broom::tidy(m1)
```

## Marginal effects and counterfactuals:

```{r}
#| label: counterfactuals

# library(marginaleffects)

m1 |>
    plot_predictions(
    condition=c("baseline_seizures","drug"),
    ) +
    labs(
    y="Predicted seizures"
) + coord_cartesian(c(-5,150),c(-5,150))

cf <- bind_rows(
#
    predictions(
    m1,
    newdata=transform(epil,
    drug="placebo")) |>
    mutate(group="Placebo"),
#
    predictions(
    m1,
    newdata=transform(epil,
    drug="progabide")) %>%
    mutate(group="Treatment")
)

cf |>
    ggplot(aes(estimate,fill=group))+
    geom_density(alpha=0.4)+
    theme_bw()

cmp <- comparisons(
    m1,
    variables="drug"
)

# Each bar is the estimated reduction in seizures if that patient switched treatments.
cmp |>
    ggplot(aes(estimate))+
    geom_histogram(
        bins=30,
        fill="steelblue"
    )+
    theme_bw()
```

```{r}
#| label: cut-off-of-when-to-use

placebo <- predictions(
    m1,
    newdata = transform(epil, drug = "placebo")
)

treated <- predictions(
    m1,
    newdata = transform(epil, drug = "progabide")
)

epil$placebo <- placebo$estimate
epil$treated  <- treated$estimate

epil$benefit <- epil$placebo - epil$treated

epil |>
    filter(benefit >= 3)

grid <- datagrid(
    model=m1,
    baseline_seizures = seq(0, 100, by = 1),
    age = mean(epil$age), #28 yo
    #V4 = 1, # follow-up
    drug = c("placebo", "progabide")
)

pred <- marginaleffects::predictions(m1, newdata = grid)

cutoff <- pred |>
    dplyr::select(baseline_seizures, drug, estimate) |>
    tidyr::pivot_wider(
        names_from = drug,
        values_from = estimate) |>
    mutate(
        reduction = placebo - progabide
    )

cutoff_pt <- cutoff |>
  slice_min(abs(reduction - (3)), n = 1)

worsening <- cutoff |>
  slice_min(abs(reduction - (0)), n = 1)

ggplot(cutoff,
       aes(baseline_seizures, reduction)) +
    geom_line(size = 1.2) +
    geom_hline(yintercept = 3,
               linetype = 2,
               colour = "red") +
    labs(
        x = "Baseline seizures",
        y = "Expected reduction",
        title = "Expected treatment benefit"
    ) +
    theme_bw() +
    geom_label(
        data = cutoff_pt,
        aes(
            label = paste0(
            "Baseline = ", baseline_seizures,
            "\nReduction = ", round(reduction, 2))
        ),
        nudge_y = -1.5
    ) +
    geom_point(aes(baseline_seizures,reduction),
        cutoff_pt,
        colour = "steelblue",
        size = 3)
```

manual predictions:

```{r}
#| label: manual-predict

x_seq <- seq(min(epil$base), max(epil$base), length.out = 100)

b <- coef(m2_baseline)

mu_hat <- exp(
  b[1] +
  b["baseline_seizures"] * x_seq +
  b["drugprogabide"] * 0       # reference treatment
  #b["base"] * mean(epil$lbase)
)



plot(x_seq, mu_hat, type = "l")

# placebo
newdata <- data.frame(
  baseline_seizures = x_seq,
  drug = "placebo"
)
newdata$mu_hat <- predict(
  m2_baseline,
  newdata = newdata,
  type = "response"
)

# progabide
newdata2 <- data.frame(
    baseline_seizures = x_seq,
    drug = "progabide"
)
predict(m2_baseline, newdata2, type = "response")

mu_prog <- exp(
    b["(Intercept)"] +
    b["drugprogabide"] +
    (b["baseline_seizures"] + b["drugprogabide:baseline_seizures"]) * x_seq
)
data.frame(x=x_seq,y=mu_prog) |>
    ggplot(aes(x=x,y=y)) +
    geom_line()

#X <- model.matrix(~ trt + base + `trtprogabide:base`, newdata)
#X <- model.matrix(~ trt * base, newdata)
tt <- delete.response(terms(m2_baseline))

X <- model.matrix(
  tt,
  newdata,
  xlev = m2_baseline$xlevels
)

newdata$mu_hat <- exp(X %*% coef(m2_baseline))
```

This last expression,

`exp(X %*% coef(fit))`

is essentially what predict.glm(type = "response") computes internally. It's the
most general way to think about GLM predictions: construct the design matrix X,
multiply by the coefficient vector β to get the linear predictor η=Xβ, then
apply the inverse link function. For a Poisson model with the log link, the
inverse link is simply exp().

The predicted values in a generalized linear model are obtained using **matrix
multiplication** of the design matrix and the estimated coefficient vector.

Let the design matrix be

$$
  X =
  \begin{bmatrix}
  1 & 0 & 6.0 & 0\\
  1 & 0 & 7.46 & 0\\
  \vdots & \vdots & \vdots & \vdots\\
  1 & 0 & 151 & 0
  \end{bmatrix},
$$

and let the estimated regression coefficients be

$$
  \hat{\beta} =
  \begin{bmatrix}
  1.29584\\
  -0.24923\\
  0.02145\\
  0.00042
  \end{bmatrix}.
$$

The linear predictor is obtained by matrix multiplication:

$$
  \eta = X\hat{\beta}.
$$

In R, this is computed as

```r
eta <- X %*% coef(fit)
```

Here,

- $X$ is a $100 \times 4$ design matrix,
- $\hat{\beta}$ is a $4 \times 1$ coefficient vector,
- $\eta$ is a $100 \times 1$ vector of linear predictors.

For the first observation, the linear predictor is computed as the product of
the first row of $X$ and the coefficient vector:

$$
  \eta_1
  =
  \begin{bmatrix}
  1 & 0 & 6 & 0
  \end{bmatrix}
  \begin{bmatrix}
  1.29584\\
  -0.24923\\
  0.02145\\
  0.00042
  \end{bmatrix}
  =
  1(1.29584)
  +0(-0.24923)
  +6(0.02145)
  +0(0.00042)
  =
  1.42453.
$$

This row-by-column multiplication is an **inner product**, also called a **dot
product**, between the first row of the design matrix and the coefficient
vector.

Thus,

- The overall computation,
  $$
  \eta = X\hat{\beta},
  $$
  is **matrix multiplication**.

- Each individual fitted value,
  $$
  \eta_i = x_i^\top \hat{\beta},
  $$
  is an **inner product (dot product)**.

Since this is a Poisson generalized linear model with a log link, the predicted
mean response is obtained by applying the inverse link function:

$$
  \hat{\mu} = \exp(\eta).
$$

In R:

```r
eta <- X %*% coef(fit)   # matrix multiplication
mu  <- exp(eta)          # inverse log link
```

A **cross product** is **not** being computed here. In vector algebra, the cross
product is defined only for three-dimensional vectors and produces another
vector. Although R has a function named `crossprod()`, it computes

$$
  X^\top Y,
$$

which is a matrix product, not the geometric cross product.

### Exposure and the offset

The parameter λ is the expected value of a Poisson model --- but also commonly
interpreted as a *rate*.

The rate interpretation is nice for allowing varying expoure time.

Implizitly, λ is equal to an expected number of events, μ, per unit time or
distance, τ. This implies that λ = μ/τ, which lets us redefine the link:

$$
  y_i \sim Poisson(λ_i) \\
  log λ_i = log \frac{μ_i}{τ_i} = α + βx_i
$$

Since the logarithm of a ratio is the same as the difference of logarithms:

$$
  log λ_i = log μ_i - log τ_i = α + βx_i
$$

These τ are the "exposures". If different observations *i* have different
exposures, the expected value on row *i* is given by

$$
  log μ_i = log τ_i + α + βx_i
$$

When τ_i = 1, then logτ_i = 0 and we're back to standard Poisson!

But when the exposure is not 1, τ does effectively *scales* the expected number
of events for each case *i* in the unit x = 1 (1 day, year, whatever...).
Therefore, we write:

$$
  y_i \sim Poisson(μ_i) \\
  log μ_i = log τ_i + α + βx_i
$$

Where τ is a column in the data. It's just like adding a predictor (taking its
log) without adding a parameter for it!

If you put a parameter in front of τ, you're able to model the hypothesis that
the rate is not constant with time.

*Remember:* When the length of observations, area of sampling, or intensity of
sampling varies, the counts we observe also naturally vary! Poisson assumes
stationarity (time/space) --- here, we add an offset! That is --- the logarithm
of the exposure unit.

Think of days of monitoring a patient...

```{r}
#| label: offset-example

set.seed(636961739)
num_days <- 30 # days of the experiment

hospital1 <- rpois (num_days, 1.5) # 1.5 events per day

num_weeks <- 4 # weeks of the experiment

hospital2 <- rpois(num_weeks, 0.5*7) # events * lowst unit (i.e., days)

events <- c(hospital1, hospital2)
exposure <- c(rep(1,30), rep(7,4))
hospital_id <- c(rep(0,30), rep(1,4))
hospital_data <- data.frame(
  events=events,
  days=exposure,
  hospital_id=hospital_id
)

hospital_data$log_days = log(hospital_data$days)

sim_m1 <- glm( events ~ 1 + hospital_id,
    family = poisson,
    offset = log_days,
    data = hospital_data
)

broom::tidy(sim_m1,exponentiate = TRUE)
# We simulated 1.5 for hospital1 (intercept) and 0.5 (a third or 33%) for hospital2. So OR of .31
```

### Negative binomial (gamma-Poisson) models

Poisson models often have a lot of unexplained variance, mainly from ignoring
*rate heterogeneity*.

The gamma-Poisson is a mixture of different Poisson distributions (similar to
how Student-t is a mixture of Gaussians).

```{r}
#| label: assumptions

performance::check_overdispersion(m2_baseline)

performance::check_model(m2_baseline)

m3_nb <- MASS::glm.nb(seizure_count ~ drug * baseline_seizures,
    data = epil
)

performance::check_model(m3_nb)
```

# Other dataset

https://search.r-project.org/CRAN/refmans/dobson/html/doctors.html?utm_source=chatgpt.com
https://search.r-project.org/CRAN/refmans/ACSWR/html/bs1.html?utm_source=chatgpt.com
https://github.com/agbarnett/dobson/tree/main/data

```{r}
#| eval: false
#| include: false
#| echo: false

url <- "https://raw.githubusercontent.com/agbarnett/dobson/main/data/doctors.rda"
load(url(url))
doctors

glm(
  deaths ~ age + smoking,
  family = poisson(),
  offset = log(personyears),
  data = doctors
)

doctors2 <- doctors |>
  mutate(
    age_group = factor(
      age,
      levels = 1:5,
      labels = c("35–44", "45–54", "55–64", "65–74", "75–84")
    ),
    rate = deaths / personyears,
    rate_1000 = 1000 * rate,
    offsets = log(personyears)
  ) |>
  select(age_group, smoking, deaths, personyears, rate, rate_1000, offsets)

doctors2
```

```{r}
#| eval: false
#| include: false
#| echo: false

ggplot(doctors2,
       aes(age_group, rate_1000,
           colour = smoking,
           group = smoking)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  labs(
    x = "Age group",
    y = "Deaths per 1,000 person-years",
    colour = "Smoking"
  ) +
  theme_bw()

ggplot(doctors2,
       aes(personyears, deaths,
           colour = smoking)) +
  geom_point(size = 3) +
  scale_x_log10() +
  theme_bw()

ggplot(doctors2,
       aes(age_group,
           log(rate),
           colour = smoking,
           group = smoking)) +
  geom_point(size = 3) +
  geom_line() +
  labs(y = "Log mortality rate")
```

```{r}
#| eval: false
#| include: false
#| echo: false

m1 <- glm(
  deaths ~ offset(offsets) + smoking + age_group,
  #offset = offsets,
  family = poisson(),
  data = doctors2
)

summary(m1)
```

```{r}
#| eval: false
#| include: false
#| echo: false

avg_predictions(
    m1,
    variables = "smoking",
    type = "response"
)

cf <- bind_rows(

    mutate(doctors2,
           smoking = "smoker",
           scenario = "Everyone smokes"),

    mutate(doctors2,
           smoking = "non-smoker",
           scenario = "Nobody smokes")

)

pred <- predictions(
    m1,
    newdata = datagrid(
        age_group = levels(doctors2$age_group),
        smoking = levels(doctors2$smoking),
        personyears = 1000
    )
)

ggplot(pred,
       aes(age_group,
           estimate,
           colour = smoking,
           group = smoking)) +
    geom_line() +
    geom_point(size = 3) +
    ylab("Predicted deaths per 1,000 person-years")

avg_comparisons(
    m1,
    variables = "smoking",
    comparison = "ratio"
)
```

## Exploratory analysis and modelling strategy

The outcome is the number of graduates who survived, while the total number of
graduates represents the population at risk. The analysis therefore focuses on
modelling survival *rates* rather than raw counts by including the log of the
total number of graduates as an offset in Poisson regression. This estimates
multiplicative effects on the survival rate while accounting for differences in
cohort size.

The exploratory analysis first examines temporal patterns, differences between
faculties, and differences by sex to assess whether systematic trends or effect
modification are plausible. Two Poisson models are then compared: a main-effects
model and a model allowing the temporal trend to differ by faculty. Because
Poisson models assume equality of the mean and variance, overdispersion is
assessed before interpreting results. Finally, marginal effects summarize
adjusted contrasts, while counterfactual predictions visualize model-implied
survival rates under hypothetical scenarios across years and demographic groups.

```{r}
#| eval: false
#| include: false
#| echo: false

# https://github.com/agbarnett/dobson/blob/main/data/chronic.rda
url <- "https://raw.githubusercontent.com/agbarnett/dobson/main/data/graduates.rda"
load(url(url))
doctors

#load("/Users/ze18rp/Downloads/chronic.rda")
#chronic

load("/Users/ze18rp/Downloads/graduates.rda")
graduates

#library(tidyverse)
#library(performance)
#library(marginaleffects)

graduates2 <- graduates |>
  mutate(
    survive = as.integer(survive),
    total = as.integer(total),
    year_c = year - min(year),
    faculty = as.factor(faculty)
  )

## -------------------------------------------------
## EDA
## -------------------------------------------------

# 1. Survival rate over time
ggplot(graduates2,
       aes(year, survive / total,
           colour = faculty,
           group = interaction(faculty, sex),
           shape = sex)
        ) +
  geom_line(alpha = 0.7) +
  geom_point() +
  labs(
    y = "Survival rate",
    x = "Year"
  ) +
  theme_bw()

# 2. Distribution of survival rates by faculty and sex
ggplot(graduates2,
       aes(faculty, survive / total, fill = sex)) +
  geom_boxplot() +
  labs(y = "Survival rate") +
  theme_bw()

# 3. Cohort sizes over time
ggplot(graduates2,
       aes(year, total,
           colour = faculty,
           shape = sex)) +
  geom_point(size = 2) +
  geom_line(aes(group = interaction(faculty, sex))) +
  labs(y = "Total graduates") +
  theme_bw()

## -------------------------------------------------
## Poisson GLMs
## -------------------------------------------------

m1 <- glm(
  survive ~ year_c + faculty + sex + offset(log(total)),
  family = poisson,
  data = graduates2
)

m2 <- glm(
  survive ~ year_c * faculty + sex + offset(log(total)),
  family = poisson,
  data = graduates2
)

summary(m1)
summary(m2)

# Check Poisson assumption
performance::check_overdispersion(m1)
performance::check_overdispersion(m2)

## -------------------------------------------------
## Marginal effects
## -------------------------------------------------

# Average adjusted comparison between faculties
avg_comparisons(m2, variables = "faculty")
avg_comparisons(
  m2,
  variables = list(faculty = "pairwise")
)
nd <- datagrid(
  year_c = mean(graduates2$year_c),
  sex = levels(factor(graduates2$sex)),
  faculty = levels(graduates2$faculty),
  total = 100
)
avg_comparisons(
  m2,
  variables = list(faculty = "pairwise"),
  newdata = datagrid(
    year_c = mean(graduates2$year_c),
    sex = unique,
    faculty = unique,
    total = 100
  )
)

# Average slope for year
avg_slopes(m2, variables = "year_c")

## -------------------------------------------------
## Counterfactual predictions
## -------------------------------------------------

# Predicted survival rate over time by faculty
plot_predictions(
  m2,
  condition = c("year_c", "faculty"),
  newdata = datagrid(
    year_c = seq(min(graduates2$year_c),
                 max(graduates2$year_c),
                 length.out = 100),
    faculty = unique(graduates2$faculty),
    sex = "men",
    total = 100
  ),
  type = "response"
)

# Predicted survival rate by year and sex
plot_predictions(
  m2,
  condition = c("year_c", "sex"),
  newdata = datagrid(
    year_c = seq(min(graduates2$year_c),
                 max(graduates2$year_c),
                 length.out = 100),
    faculty = "medicine",
    sex = unique(graduates2$sex),
    total = 100
  ),
  type = "response"
)
```
