# Exercises: Hallo Medical Statistics {#sec-ex-intro}

::: {.callout-tip collapse="false"}
## Hallo Medical Statistics

Pre-requicite for these exercises was covered within the
[session](/sessions/session01.qmd) and is therefore not repeated here. The
exercises build on this material.
:::

## Learning objectives

The learning objectives for this session are:

{{< include /includes/objectives/_session1.qmd >}}

## Exercises: "Just send a quick summary"

You have been asked to prepare a quick summary of a dataset of 303 chest pain
patients to support a potential clinical trial collaboration. "Nothing
complicated---just mean and standard deviation for a few variables." You open
RStudio...

### A: Let's start orienting ourselves in the dataset.

**Exercise 1A**: Import the Cleveland Heart Disease dataset into R.

- What is the name of your dataset?
- How many rows and columns does it have?
- How many variables does it have?
- How many participants does it have?
- Is there any pattern here?

View the dataset `sleep_data` from the website. Import the data in your RStudio
session. How many variables does it have?

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are struggling or are out of time."

View(sleep_data)
```

**Exercise 2A**: View the dataset.

- List 3 variables you think are numeric
- List 2 variables you think are categorical

Why does this distinction matter when calculating a mean?

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are struggling or are out of time."

numeric_vars <- c(1,2,3)
categorical_vars <- c(4,5,6)
```

**Exercise 3A**:

What happens when you take the mean of a character string? e.g., `mean("hi")`?
Why do you think this happens?

**Exercise 4A**:

Extract one variable (e.g., Age or Cholesterol) and:

- Print it to see alt it's values
- Calculate its mean

### B: Now you start doing the actual request from the email.

**Exercise 1B**:

Calculate the mean and standard deviation for:

- Age
- Chol
- RestBP
- MaxHR

**Exercise 2B**:

Check if there are missing values in the dataset.

- How many are there and for which variables (tip: view it or use `is.nan()`)?
- What happens if you try to calculate the mean when missing values are present?

**Exercise 3B**:

List all the variable types the `mean()` function can take the mean of. There
should be 5. Seek help inside RStudio.

### C: You reflect on your day at work during your evening at home...

**Exercise 1C**:

Consider now what you took the mean and standard deviation of in exercise 3B.
How can you interpret this? What can we say from this?

**Exercise 2C**:

A colleague suggested calculating the mean of Sex.

- What would that number represent?
- Is it meaningful? Why or why not?

**Exercise 3C**:

Why is it important to report both mean and standard deviation, and not just the
mean? Think in a clinical context.

## Survey

{{< include /includes/_survey.qmd >}}
