# Exercises: Hallo Medical Statistics {#sec-ex-intro}

::: {.callout-tip collapse="false"}
## Hallo Medical Statistics

Pre-requicite for these exercises was covered within the
[session](/sessions/session01.qmd) and is therefore not repeated here. The
exercises build on this material.
:::

## Learning objectives

The learning objectives for this session are:

{{< include /includes/objectives/_session1.qmd >}}

## Exercises: "Just send a quick summary"

You have been asked to prepare a quick summary of a dataset of 303 chest pain
patients to support a potential clinical trial collaboration. "Nothing
complicated---just mean and standard deviation for a few variables." You open
RStudio...

### A: Let's start orienting ourselves in the dataset.

**Exercise 1A**: Import the Cleveland Heart Disease dataset into R.

- What is the name of your dataset?
- How many rows and columns does it have?
- How many variables does it have?
- How many participants does it have?
- Is there any pattern here?

View the `Cleveland Heart Disease` dataset from the website. Import the data in
your RStudio session. How many variables does it have?

<details>
<summary>**Click for the solution**</summary>

 1. Depends on your naming choice --- could be `hd_data` as on the website.
 2. 303 by 15
 3. 15
 4. 303
 5. yes... 👀

 </details>

```{r}
#| eval: false
#| code-fold: true
#| code-summary: >-
#|   **Click for a tip**.

View(hd_data)
```

**Exercise 2A**: View the dataset.

- List 3 variables you think are numeric
- List 2 variables you think are categorical

Why does this distinction matter when calculating a mean?

<details>
<summary>**Click for the solution**</summary>

1. Age, RestBP, Chol, MaxHR
2. Sex, ChestPain, Fbs, RestECG, ExAng, Slope, Ca, Thal, HD
3. 👀

</details>

**Exercise 3A**:

What happens when you take the mean of a character string? e.g., `mean("hi")`?
Why do you think this happens?

<details>
<summary>**Click for the solution**</summary>

1. error to protect you --- don't get scared from the [red text warning]{style="color: red;"} 😄


</details>

**Exercise 4A**:

Extract one variable (e.g., Age or Cholesterol) and:

- Print it to see all it's values
- Calculate its mean

<details>
<summary>**Click for the solution**</summary>

```{r A4}

hd_data$Age # 1
mean(hd_data$Age) # 2

library(dplyr)

# 1
hd_data |>
  pull(Age)

# 2
hd_data |>
  pull(Age) |>
  mean()
```

</details>

### B: Now you start doing the actual request from the email.

**Exercise 1B**:

Calculate the mean and standard deviation for:

- Age
- Chol
- RestBP
- MaxHR

```{r}
hd_data |>
  pull(Age) |>
  mean()

hd_data |>
  pull(Age) |>
  sd()

# etc....
```

**Exercise 2B**:

Check if there are missing values in the dataset.

- How many are there and for which variables (tip: view it or use `is.nan()`)?
- What happens if you try to calculate the mean when missing values are present?
- Can you find out how to fix the above issue by reading the documentation?

<details>
<summary>**Click for the solution**</summary>

1. Ca and Thal has.
2. You get `NA`--- see
3. `na.rm` --- see `?mean` argument 3

</details>

**Exercise 3B**:

List all the variable types the `mean()` function can take the mean of. There
should be 5. Seek help inside RStudio.

<details>
<summary>**Click for the solution**</summary>

1. type `?mean` It says that "**Currently there are methods for 1. numeric, 2. logical 2. date, 4. date-time and 5. time interval**".

</details>

### C: You reflect on your day at work during your evening at home...

**Exercise 1C**:

Consider the age mean and standard deviation you previously calculated (in
exercise 1B). How can you interpret this? What can we say from this?

<details>
<summary>**Click for the solution**</summary>

1. mean 54.43894 sd 9.038662 - the group was between 72.5 and 36.5 years old so a good spread of adults.

</details>

**Exercise 2C**:

A colleague suggested calculating the mean of Sex.

- What would that number represent?
- Is it meaningful? Why or why not?

<details>
<summary>**Click for the solution**</summary>

1. It is meaningful --- it's the percent of males.

</details>

**Exercise 3C**:

Why is it important to report both mean and standard deviation, and not just the
mean? Think in a clinical context.

<details>
<summary>**Click for the solution**</summary>

1. standard deviation tells about variability in data. There could be outliers to the mean.

</details>

## Survey

{{< include /includes/_survey.qmd >}}
