# Exercises: When in doubt, plot it out {#sec-ex-ggplot2}

::: {.callout-tip collapse="false"}
## Hallo Medical Statistics

Pre-requicite for these exercises was covered within the
[session](/sessions/session01.qmd) and is therefore not repeated here. The
exercises build on this material.
:::

## Exercises: "Let's first see what the data are telling us"

You are given a dataset containing demographic measurements from !Kung San
individuals, including age, height, weight, and sex.

The research team wants to explore a simple question:

> How do height and weight vary with age, and do these patterns differ between
> males and females?

Before drawing conclusions, the team wants to first understand the data
visually. In particular, they want to examine the distributions of height and
weight, compare males and females, and explore relationships between the
continuous variables.

The Professor asks you to start by asking:

> Which visualization is most appropriate for each question?

```{r}
#| label: setup
#| echo: false

library(ggplot2)
```

### A: Let's start orienting ourselves in the dataset.

**Exercise 1A**:

Import the Height dataset into R --- see @sec-height-data for more info.

<details>
<summary>**Click for the solution**</summary>

```{r}
#| label: import-data

height_data <- medisr::height_data
```
</details>

**Exercise 2A**:

View the dataset.

1. How many observations do we have?
2. How many varialbes?
3. What type are they?

<details>
<summary>**Click for the solution**</summary>

```{r}
#| label: view-data
#| eval: false

View(height_data) <- medisr::height_data
```

1. 544?
2. 4?
3. all numeric (`dbl` and `int`)

this is easier to see if you make the dataset a `tibble`
```{r as-tibble}
height_data |>
    tibble::as_tibble()
```

</details>

**Exercise 3A**:

Are we happy with going strait to plotting or do we need to wrangle the dataset
a bit with `{dplyr}` first?

<details>
<summary>**Click for the solution**</summary>
male is obviously not continuous but rather a yes/no factor. So we should consider this later before comparing males and females.
</details>

### B: Now you start exploring the dataset

**Exercise 1B**:

The variable `male` is coded as `0` and `1`. Before plotting, convert it into a
factor with meaningful labels:

- `0` should represent `"Female"`
- `1` should represent `"Male"`
- Print the resulting variable to check that the transformation worked.

<details>
<summary>**Click for the solution**</summary>

```{r A4}

height_data <- height_data |>
  dplyr::mutate(
    male = factor(
      male,
      levels = c(0, 1),
      labels = c("Female", "Male")
    )
  )

height_data  |>
  dplyr::pull(male)
```

</details>

**Exercise 2B**:

Create a plot that proofes to yourself that the age of males and females are
similar within the dataset.

- What will you put on the x-axis?
- What will you put on the y-axis?
- Any color coding?
- Which `geom_`?

**First** draw it so that the two groups have similar age in the population.
Consider the fact that this is a demographic study with the entire population of
the Kalahari !Kung San people (so we have a wide range of ages).

**Second** implement the plot in R.

<details>
<summary>**Click for the solution**</summary>

```{r A4B}

height_data |>
  ggplot(
    aes(
      x = male,
      y = age,
      color = male
    )
  ) +
  geom_boxplot()
```
</details>

**Exercise 3B**:

Compare the `geom_smooth()` curves for males and females.

Look at the graph and answer:

- At approximately what age does height stop increasing rapidly?
- Does height appear to reach a plateau?
- Do males and females reach this plateau at approximately the same age?
- Which sex appears to reach a greater adult height?
- Are there ages where the difference between males and females appears
  particularly large?

<details>
<summary>**Click for the solution**</summary>

There is no single numerical answer to these questions. Use the smoothed curves to describe the pattern in the data.

In particular, compare where each curve changes from a steep increase to a relatively flat section. Also compare the approximate height of the two curves during adulthood.

The exact age at which growth "stops" cannot be determined simply by looking at the graph, but the `geom_smooth()` curves can help identify an approximate age range where height becomes relatively stable.

</details>

**Exercise 4B**:

Investigate whether the relationship between **height and weight** appears to be
linear.

Create a scatter plot with:

- `height` on the x-axis
- `weight` on the y-axis
- Points colored by sex
- A separate `geom_smooth()` curve for each sex

Then answer:

- Does weight increase approximately linearly with height?
- At what range of heights does the relationship appear to become less linear?
- Does the shape of the relationship look similar for males and females?
- Is there evidence that one sex tends to weigh more at the same height?

<details>
<summary>**Click for the solution**</summary>

```{r A4D}

height_data |>
  ggplot(
    aes(
      x = height,
      y = weight,
      color = male
    )
  ) +
  geom_point() +
  geom_smooth()
```
</details>

### C: You reflect on your day at work during your evening at home...

**Exercise 1C**:

A researcher wants to investigate the following hypothesis:

> **Hypothesis:** The distribution of height differs between males and females.

Which visualization would be most appropriate?

Explain **why** you chose this visualization and what features of the graph you
would look at to evaluate the hypothesis.

Now implement it using `{ggplot2}`.

<details>
<summary>**Click for the solution**</summary>

Code depends on your plot of choice 👀

</details>

**Exercise 2C**:

Consider the hypothesis:

> **Hypothesis:** Males and females have different median body weights.

Which visualization would be most appropriate?

Explain your reasoning and implement the plot using `{ggplot2}`.

<details>
<summary>**Click for the solution**</summary>

Did you choose box plot? 👀

</details>

**Exercise 3C**:

Consider the hypothesis:

> **Hypothesis:** The variability in body weight is greater among males than
> among females.

Which visualization would be most appropriate?

<details>
<summary>**Click for the solution**</summary>

**Best choice: Box plot**

Weight is continuous and sex provides two groups. A box plot makes it easy to compare the spread of weight between the groups.

The size of the box and the whiskers provide information about variability.

</details>

**Exercise 4C**:

::: {.hidden}
## Survey

{{< include /includes/_survey.qmd >}}
:::
