# Exercises: When in doubt, plot it out {#sec-ex-ggplot2}

::: {.callout-tip collapse="false"}
## Hallo Medical Statistics

Pre-requicite for these exercises was covered within the
[session](/sessions/session01.qmd) and is therefore not repeated here. The
exercises build on this material.
:::

## Exercises: "Let's first see what the data are telling us"

You are given a dataset containing demographic measurements from !Kung San
individuals, including age, height, weight, and sex.

The research team wants to explore a simple question:

> How do height and weight vary with age, and do these patterns differ between
> males and females?

Before drawing conclusions, the team wants to first understand the data
visually. In particular, they want to examine the distributions of height and
weight, compare males and females, and explore relationships between the
continuous variables.

The Professor asks you to start by asking:

> Which visualization is most appropriate for each question?

```{r}
#| label: setup
#| echo: false

library(ggplot2)
```

### A: Preparing to start plotting

Before we can plot anything, we need our data and we need to know which state
and format it's in. This sections shows a common workflow to do this.

**Exercise 1A**:

Import the Height dataset into R --- see @sec-height-data for more info.

<details>
<summary>**Click for the solution**</summary>

```{r}
#| label: import-data

height_data <- medisr::height_data
```
</details>

**Exercise 2A**:

View the dataset.

1. How many observations do we have?
2. How many variables?
3. What type are they?

<details>
<summary>**Click for the solution**</summary>

```{r}
#| label: view-data
#| eval: false

View(height_data) <- medisr::height_data
```

1. 544
2. 4
3. all numeric (`dbl` and `int`)

this is easier to see if you make the dataset a `tibble`
```{r as-tibble}
height_data |>
    tibble::as_tibble()
```

</details>

**Exercise 3A**:

Are we happy with going strait to plotting or do we need to wrangle the dataset
a bit with `{dplyr}` first?

<details>
<summary>**Click for the solution**</summary>
male is obviously not continuous but rather a yes/no factor. So we should consider this later before comparing males and females.
</details>

### B: Exploratory data analysis with plotting

It is extremely useful (and often necessary) to plot your data before you start
doing any statistics or conclusions. This section covers a typical workflow.

**Exercise 1B**:

The variable `male` is coded as `0` and `1`. Before plotting, convert it into a
factor with meaningful labels:

- `0` should represent `"Female"`
- `1` should represent `"Male"`
- Print the resulting variable to check that the transformation worked.

<details>
<summary>**Click for the solution**</summary>

```{r A4}

height_data <- height_data |>
  dplyr::mutate(
    male = factor(
      male,
      levels = c(0, 1),
      labels = c("Female", "Male")
    )
  )

height_data  |>
  dplyr::pull(male)
```

</details>

**Exercise 2B**:

Create a plot that proofes to yourself that the median age of males and females
are similar within the dataset.

- What will you put on the x-axis?
- What will you put on the y-axis?
- Any color coding?
- Which `geom_`?

**First** draw it so that the two groups have similar age in the population.
Consider the fact that this is a demographic study with the entire population of
the Kalahari !Kung San people (so we have a wide range of ages).

**Second** implement the plot in R.

<details>
<summary>**Click for the solution**</summary>

```{r A4B}

height_data |>
  ggplot(
    aes(
      x = male,
      y = age,
      color = male
    )
  ) +
  geom_boxplot()
```
</details>

**Exercise 3B**:

Investigate whether the relationship between **height and weight** appears to be
linear.

Create a scatter plot with:

- What will you put on the x-axis?
- What will you put on the y-axis?
- Any color coding?
- Which `geom_`?

Then answer:

- Does weight increase approximately linearly with height?
- Does the shape of the relationship look similar for males and females?
- Is there evidence that one sex tends to weigh more at the same height?

<details>
<summary>**Click for the solution**</summary>

```{r A4D}

height_data |>
  ggplot(
    aes(
      x = height,
      y = weight,
      color = male
    )
  ) +
  geom_point() +
  geom_smooth()
```
</details>

### C: Fitting a hypothesis with a plot type

Finally, plots often accompany hypothesis. We want to show the reader of our
plot that it supports a specific conclusion. Here, you get experience fitting
plot types with hypotheses.

**Exercise 1C**:

A researcher wants to investigate the following hypothesis:

> **Hypothesis:** There is more variability in male height than in female
> height.

Which visualization would be most appropriate?

Explain **why** you chose this visualization and what features of the graph you
would look at to evaluate the hypothesis.

Now implement it using `{ggplot2}`.

<details>
<summary>**Click for the solution**</summary>

Code depends on your plot of choice 👀

Example:

```{r ex-c1}
height_data |>
  ggplot(
    aes(
      x = height,
      color = male
    )
  ) +
  geom_density()
```

</details>

**Exercise 2C**:

Consider the hypothesis:

> **Hypothesis:** Males have higher body weight than females.

Which visualization would be most appropriate?

Explain your reasoning and implement the plot using `{ggplot2}`.

<details>
<summary>**Click for the solution**</summary>

Did you choose box plot? 👀

Example:

```{r ex-c2}
height_data |>
  ggplot(
    aes(
      x = male,
      y = weight,
      color = male
    )
  ) +
  geom_boxplot()
```

</details>

**Exercise 3C**:

Consider the hypothesis:

> **Hypothesis:** Females increase their height earlier in development than
> males.

Which visualization would be most appropriate?

Explain your reasoning and implement the plot using `{ggplot2}`.

<details>
<summary>**Click for the solution**</summary>

example:

```{r ex-c3}
height_data |>
  ggplot(
    aes(
      x = age,
      y = height,
      color = male
    )
  ) +
  geom_point() +
  geom_smooth() +
  xlim(c(0,20)) # this line helps a lot by zooming in at the age of interest!
```

</details>

::: {.hidden}
**Exercise 4C**: Compare the `geom_smooth()` curves for males and females.

Look at the graph and answer:

- At approximately what age does height stop increasing rapidly?
- Does height appear to reach a plateau?
- Do males and females reach this plateau at approximately the same age?
- Which sex appears to reach a greater adult height?
- Are there ages where the difference between males and females appears
  particularly large?

<details>
<summary>**Click for the solution**</summary>

There is no single numerical answer to these questions. Use the smoothed curves to describe the pattern in the data.

In particular, compare where each curve changes from a steep increase to a relatively flat section. Also compare the approximate height of the two curves during adulthood.

The exact age at which growth "stops" cannot be determined simply by looking at the graph, but the `geom_smooth()` curves can help identify an approximate age range where height becomes relatively stable.

</details>

## Survey

{{< include /includes/_survey.qmd >}}
:::
