# Exercises: Conditional variables {#sec-ex-interaction}

::: {.callout-tip collapse="false"}
## Conditional variables

Pre-requicite for these exercises was covered within the
[session](/sessions/session11.qmd) and is therefore not repeated here. The
exercises build on this material.
:::

## Learning objectives

The learning objectives for this session are:

{{< include /includes/objectives/_session1.qmd >}}

## Exercises: "name-of-exercises"

Does the association between uric acid and blood glucose differ across patient
groups?

A research team in Zhejiang Province is interested in whether elevated serum
uric acid (UA) is associated with fasting blood glucose (FBG). Previous studies
suggest that this relationship may not be identical for all patients. For
example, hormonal differences, obesity, and hypertension may alter the
biological pathways linking uric acid metabolism and glucose regulation.

The investigators therefore suspect that the effect of UA on FBG may depend on
other patient characteristics.

### A: Let's start practicing by finding interactions from written text.

**Exercise 1A**:

For each of the causal relationships below, name a hypothetical third variable
that would lead to an interaction effect:

1) Higher serum uric acid is associated with higher fasting blood glucose.
2) Higher BMI is associated with higher fasting blood glucose.
3) Reduced kidney function (lower GFR) is associated with higher serum uric
   acid.

Hint: Think of a variable that could make the relationship stronger in some
patients and weaker in others.

**Exercise 2A**:

Which of the following explanations invokes an interaction:

1) Men tend to have higher uric acid levels than women.
2) Older participants tend to have higher blood pressure.
3) The association between uric acid and fasting blood glucose is stronger among
   participants with obesity than among participants without obesity.
4) Participants with hypertension have higher average fasting blood glucose than
   participants without hypertension.

**Exercise 3A**:

For each of the explanations in **2A**, write a linear model that expresses the
stated relationship.

Assume:

- FBG = fasting blood glucose
- UA = uric acid
- BMI = body mass index
- Age = age
- HTN = hypertension status

1) Men tend to have higher uric acid levels than women.
2) Older participants tend to have higher blood pressure.
3) The association between uric acid and fasting blood glucose is stronger among
   participants with obesity.
4) Participants with hypertension have higher average fasting blood glucose.

Which model contains an interaction term?

### B: You start working through the data you have been assigned to analyse:

The research team asks you to investigate whether the relationship between uric
acid and fasting blood glucose differs between men and women.

Import the dataset by typing:

```{r}
SUA_data <- readr::read_csv("https://zenodo.org/records/8292712/files/SUA_CVDs_risk_factors.csv")

SUA_data <- SUA_data |>
    mutate(
        sex = factor(sex,
            levels = c(1, 2),
            labels = c("male", "female")
        ),
        hyperglycemia = factor(hyperglycemia,
            levels = c(1, 2),
            labels = c("not hypoglycemic", "hypoglycemic")
        ),
    ) |>
    dplyr::filter(
        Times == 3
    )
```

**NOTE** that we only use observation 3 (the latest follow-up all participants
completed). Had we used all observations, we would not adhered to the *iid*
assumption of linear regression, which would make overconfident variance
estimates. To correct for this, we could have used other models that would be
more appropriate for this type of data (*ask an instructor if interested*).
However, for this lecture, we're satisfied with only looking at a subset of the
dataset.

**Exercise 1B**:

Fit the following model:

```r
model1 <- lm(
  FBG ~ UA + sex + UA:sex,
  data = SUA_data
)
```

1) What is the outcome variable?
2) Write up the mathematical definition of the model.
3) Which coefficient represents the interaction?
4) In plain language, what question does the interaction term answer?
5) What is the effect of UA? (hint: `summarize()`)
6) What is the effect of sex?

(*5 & 6 can be deceiving - try as best as you can and return to this question if
you learn something new*)

### Exercise 2B

Create predicted values from the model.

```r
library(marginaleffects)

plot_predictions(
    model1,
    condition = c("UA", "sex")
)
```

1) What does each line represent?
2) If the lines are parallel, what does that suggest about the interaction?
3) If the lines diverge as UA increases, what does that suggest?
4) Which sex appears to have the stronger association between UA and FBG?

**Exercise 3B**

Suppose the fitted model produced:

  | Term             | Estimate |
  | ---------------- | -------: |
  | UA               |    0.004 |
  | Sex(Female)      |    -0.20 |
  | UA × Sex(Female) |    0.003 |

1) What is the expected effect of a 100-unit increase in UA among males?
2) What is the expected effect of a 100-unit increase in UA among females?
3) Which group shows the stronger association?

Hint: The slope for females is the sum of the UA coefficient and the interaction
coefficient.

**Exercise 4B**

A colleague says:

> "The interaction coefficient is small, so the interaction is not clinically
> important."

Why is this statement incomplete?

What additional information would you want to examine before deciding whether
the interaction is clinically meaningful?

The same colleague argues that: > "as UA is not significant, it doesn't have any
effect."

Is it valid to say there is no effect? How would you respond?

**Exercise 5B**

Interactions are symmetric.

Consider the model:

$$
FBG = β_0 + β_1UI + β_2sex + β_3(sexUI) + ε
$$

1) If the effect of UA depends on sex, what can we also say about the effect of
   sex?
2) Why does this symmetry occur mathematically?
3) How would you explain this idea to a clinical colleague?

**Exercise 6B**

Use `{marginaleffects}` to estimate how the effect of UA changes across groups.

```r
plot_slopes(
  model1,
  variables = "UA",
  condition = "sex"
) +
  geom_hline(yintercept = 0, linetype = "dotted")
```

1) What quantity is plotted on the y-axis?
2) How is this plot different from the prediction plot in **2B**?
3) Which visualization would you show a clinical audience? Why?

### C: How can we write up a clincal report for our clinical colleagues:

**Exercise 1C**:

The fitted model produced the following results:

  | Term             | Estimate | 95% CI         |
  | ---------------- | -------: | -------------- |
  | UA               |    0.004 | (0.002, 0.006) |
  | Sex(Female)      |    -0.20 | (-0.35, -0.05) |
  | UA × Sex(Female) |    0.003 | (0.001, 0.005) |

Write a short results paragraph suitable for the Results section of a clinical
paper. Avoid statistical jargon where possible.

**Exercise 2C**

The figure below shows predicted fasting blood glucose values across uric acid
levels for men and women.

*TODO: (Insert prediction plot here.)*

```{r}
#| echo: false
#| fig-cap: "TODO"

library(marginaleffects)
library(ggplot2)

plot_predictions(
    model1,
    condition = c("sex", "UA")
) +
    ylab("Predicted fasting blood glucose")

```

Write:

1) A one-sentence figure caption.
2) A two-sentence explanation suitable for clinicians.

**Exercise 3C**

Continuous × Continuous Interaction

A nephrologist suggests that the association between serum uric acid (UA) and
fasting blood glucose (FBG) may become stronger as kidney function declines.

You fit:

```r
model_cc <- lm(
  FBG ~ UA + GFR + UA:GFR,
  data = SUA_data
)
```

1) What clinical question does the interaction term address?
2) If the interaction coefficient is negative, what does that imply about the
   relationship between UA and FBG as GFR increases?
3) Which effect(s) are significant and how do we interpret the UA and GFR
   coefficients?
4) Use {marginaleffects} to create a plot that shows how the slope of UA changes
   across the observed range of GFR.
6) Is there a case in which GFR doesn't affect FBG, independent of UA?

```r
#| code-fold: true
#| code-summary: '**Click for the solution**'
plot_predictions(
    model_cc,
    condition = c("GFR", "UA")
 ) +
    ylab("Predicted fasting blood glucose")
```

**Exercise 4C**

Continuous × Continuous Interaction

Researchers suspect that BMI and hyperglycemia reinforce one another
biologically. They decide to investigate whether the association between BMI and
fasting blood glucose differs according to age.

You fit:

```r
model3 <- lm(
    FBG ~ BMI + Age + BMI:Age,
    data = SUA_data
)
```

1) What would a positive interaction coefficient imply clinically?
2) Plot the predicted FBG as a function of Age conditioned on BMI
3) If two patients differ by 5 BMI units, would you expect the difference in FBG
   to be the same at age 45 and age 80? Explain.

A colleague states:

> "BMI has an effect of 0.06 mmol/L on FBG."

Why is this statement incomplete in the presence of an interaction?

```r
#| code-fold: true
#| code-summary: '**Click for the solution**'
plot_predictions(
    model3,
    condition = list(
        "Age",
        "BMI" = c(15, 20, 25, 30, 35)
        )
 ) +
    ylab("Predicted fasting blood glucose")
```

**Exercise 5C**

Three-Way Interaction

A diabetes researcher hypothesizes that the relationship between uric acid (UA)
and fasting blood glucose (FBG) may depend on both sex and BMI.

You fit:

```r
model_3way <- lm(
  FBG ~ UA * sex * hyperglycemia,
  data = SUA_data
)
```

*(Note: Hyperglycemia is coded 0/1 after recoding.)*

1) Expand the model formula into all main effects, two-way interactions, and the
   three-way interaction.
2) What clinical question is addressed by the three-way interaction term?
3) Suppose the three-way interaction coefficient is positive. What would this
   imply?
4) Why is interpreting individual coefficients difficult in this model?
5) Use {marginaleffects} to generate predicted values of FBG across UA
   separately for: males without hyperglycemia females without hyperglycemia
   males with hyperglycemia females with hyperglycemia
6) What visual pattern would suggest evidence for a three-way interaction?
7) Explain the meaning of the three-way interaction to a clinician without using
   the word "interaction."

```r
#| code-fold: true
#| code-summary: '**Click for the solution**'
plot_predictions(
    model3,
    condition = list(
        "UA",
        "sex",
        "hyperglycemia"
        )
 ) +
    ylab("Predicted fasting blood glucose")
```

**Exercise 6C**

Logistic Regression Interaction

The outcome is now whether a participant has hyperglycemia.

You fit a logistic regression model:

```r
model_logit <- glm(
  hyperglycemia ~ UA * sex,
  family = binomial,
  data = SUA_data
)
```

1) What quantity is being modeled by this regression?
2) What does the interaction term test?
3) What is the interpretation of the interaction coefficient? (hint: *the next
   plot will help*)
4) Use {marginaleffects} to plot predicted probabilities of hyperglycemia across
   UA separately for men and women.

```r
plot_predictions(
    model_logit,
    condition = c("UA", "sex"),
    type = "response" # response scale (i.e., predicted probabilities between 0 and 1)
)
```

Suppose the predicted probabilities are:

  | UA  | Male | Female |
  | --- | ---- | ------ |
  | 200 | 0.10 | 0.08   |
  | 600 | 0.03 | 0.25   |

What evidence do these values provide regarding interaction?

Explain why interaction effects on the log-odds scale and interaction effects on
the probability scale are not the same thing.
