# Exercises: A fishy counting trick {#sec-ex-poisson}

::: {.callout-tip collapse="false"}
## Conditional variables

Prerequisite for these exercises was covered within the
[session](/sessions/session12.qmd) and is therefore not repeated here. The
exercises build on this material.
:::

## Learning objectives

The learning objectives for this session are:

{{< include /includes/objectives/_session1.qmd >}}

## Exercises: Revisiting smoking regulations

You are working in the Medical Regulations Department, where a historical review
of evidence used to inform tobacco regulation is underway.

As part of the review, your team has been given mortality data from the British
Doctors Study, the landmark study initiated by Richard Doll and Austin Bradford
Hill in the 1950s. The study followed British doctors over many years and
recorded their smoking habits and subsequent mortality.

The dataset available to you is an aggregated version of the historical study.
It contains the number of deaths and the total person-years of observation for
doctors grouped by age and smoking status.

At an initial briefing, one analyst presents a simple summary of the data:

> "The mortality rate appears to be higher among smokers. The difference is
> quite striking."

This catches the attention of the regulatory team. However, another
epidemiologist raises an important concern:

> "We should be careful about interpreting that comparison. Mortality changes
> substantially with age, and the smoking groups don't necessarily contribute
> the same amount of observation at every age."

The team therefore decides that simply comparing the number of deaths between
smokers and non-smokers is not sufficient. Doctors contribute different amounts
of observation time, and age may need to be accounted for when comparing
mortality rates.

You are asked to investigate the historical data more systematically.

Your task is to examine mortality rates in relation to smoking status while
accounting for differences in person-time and age. The department is
particularly interested in whether a statistical model can describe the observed
mortality patterns and quantify the association between smoking and mortality.

The data are aggregated, so the analysis will require you to think carefully
about what the response variable represents and how the amount of observation
differs between groups.

### A: Let's start off by considering when in general to use Poisson, logistic or normal regression.

**Exercise 1A**: Poisson vs. Binomial vs. Gaussian

You are designing a study to analyze the number of hospital readmissions for
heart failure patients within 30 days of discharge.

- 1: Should you use a Poisson, Binomial, or Gaussian model? Justify your choice.
- 2: What are the key assumptions of the distribution you chose? How would you
  check them?
- 3: If the outcome were the probability of readmission (yes/no), which model
  would you use? Why?
- 4: If the outcome were the length of stay (in days), which model would you
  use? Why?

<details>
<summary>**Click for the solution**</summary>

1. Poisson.
2. Positive real numbers. Exponential. Mean and variance is the same.
3. Logistic regression.
4. Poisson.

</details>

**Exercise 1B**: Reviewing a colleagues paper.

You're assigned as an internal peer reviewer for a study titled:

> Association Between Corticosteroid Therapy and Hospital-Acquired Infections in
> an Intensive Care Unit

Here is the paper extract:

**Background**

Researchers conducted a retrospective cohort study of 612 patients admitted to
an intensive care unit (ICU). The outcome was the number of hospital-acquired
infections each patient experienced during their ICU stay. Patients contributed
different lengths of follow-up because some were discharged early while others
remained hospitalized for several weeks.

The investigators collected the following variables:

- Number of hospital-acquired infections (instances from electronic journal)
- Corticosteroid treatment (Yes/No)
- Age (years)
- Diabetes (Yes/No)
- Length of ICU stay (days)

A Poisson regression model was fitted.

**Statistical Methods**

Poisson regression was used to estimate the association between corticosteroid
therapy and infection occurrence. Age was included as the offset variable to
account for differences in patient risk over time. The model included
corticosteroid treatment as main treatment effect and adjusting for diabetes.

The fitted model was

$$
  \text{Infections} = Poisson(λ) \\
  log(λ_i​) = β0​ + β1​(Steroid) + β2​(Diabetes) + \text{offset}(Age)
$$

**results**

  | Variable               | Coefficient | Standard Error | p-value |
  | ---------------------- | ----------: | -------------: | ------: |
  | Intercept              |       -2.31 |           0.28 |  <0.001 |
  | Corticosteroid therapy |        0.47 |           0.11 |  <0.001 |
  | Diabetes               |        0.29 |           0.10 |   0.004 |

Raw coefficients can be seen from the table above.

Patients receiving corticosteroids had 0.47 times higher odds of developing
hospital-acquired infections compared with patients who did not receive
corticosteroids (p < 0.001).

These findings indicate that corticosteroid therapy substantially increases the
odds of hospital-acquired infection after adjustment for age and diabetes.

**Discussion**

After adjusting for age using an offset variable, corticosteroid therapy
remained an independent predictor of hospital-acquired infection. The regression
coefficient of 0.47 suggests a clinically meaningful increase in infection risk.

These results support careful monitoring of patients receiving corticosteroids
in the ICU.

**Your task:**

- Identify all errors you can --- you may ask for tips.

<details>
<summary>**Click for tips**</summary>
There are several mistakes:

- What was age doing in the model?
- What is the table outputting and how did they make inference in the result and discussion section about that?
- Is this Poisson or Logistic regression?

</details>

<details>
<summary>**Click for the solution**</summary>

1. Wrong offset
  - The model uses Age as the offset.
  - In Poisson regression, the offset should represent the logarithm of exposure or person-time (e.g., log(length of ICU stay)), not a patient characteristic.
  - Age should be included as a regular covariate if adjustment is desired.
2. Wrong interpretation
  - The text interprets the results as odds ("0.47 times higher odds").
  - Standard Poisson regression estimates log-ratios, not (log) odds ratios. Saying odds ratios or incidence rate ratio (IRRs) is double wrong (see next point).
3. Coefficients not exponentiated
  - The table reports coefficients (0.47 and 0.29), but the narrative treats them as effect sizes.
  - The correct interpretation requires exponentiation:
    - exp(0.47) = 1.60, meaning a 60% higher infection rate.
    - exp(0.29) = 1.34, meaning a 34% higher infection rate.
</details>

### B: Analyzing the historic "Doctors" dataset

Let's start by loading the dataset:

```{r}
#| label: load-data

doctors <- medisr::doctors
```

You should also read the documentation `?medisr::doctors`.

**Exercise 1B**

Approach the dataset naïvely by simply counting up the number of deaths for
smokers and non-smokers.

- How many died?

;<details>
<summary>**Click for the solution**</summary>

```{r}
#| label: 1b

library(dplyr)

doctors |>
    group_by(smoking) |>
    summarise(n_deaths = sum(deaths))
```

- non-smoker: 101
- smoker: 630

</details>

**Exercise 2B**

Check the balance in the groups.

- How many years in total did the researchers follow the smoking group and how
  many the non-smoking group?

;<details>
<summary>**Click for the solution**</summary>

```{r}
#| label: 2b

doctors |>
    group_by(smoking) |>
    summarise(n_years = sum(personyears))
```

- non-smoker: 39220
- smoker: 142247

</details>

**Exercise 3B**

Finally, check whether the age groups are composed evenly and whether they are
comparable between the smoking groups.

1. Which group is nearly half of the participants between 35 to 44 years old?
2. In the other group, how many of the age categories has higher percent
   distribution?

*Tip*: For each group, compute the percent of all participants belonging to each
of the five age groups and visually compare them one by one.

;<details>
<summary>**Click for the solution**</summary>

```{r}
#| label: 3b

doctors |>
  group_by(smoking) |>
  mutate(percent = 100 * personyears / sum(personyears)) |>
  ungroup() |>
  select(age, smoking, percent)
```

<!-- panache-ignore-start-->
    age smoker `non-smoker`
  <dbl>  <dbl>        <dbl>
1     1  36.8         47.9
2     2  30.4         27.2
3     3  20.1         14.6
4     4   8.90         6.59
5     5   3.74         3.73
<!-- panache-ignore-end-->

1. non-smokers
2. 2---5, so the 4 older groups.

</details>

**Exercise 4B**

Now, run the model naïvely, without accounting for these differences.

The model will be:

$$
  deaths = Poisson(λ) \\
  log(λ) = β0 + β1_{smoking}
$$
Report the (incorrect) estimand for incidence rate ratio for smoking vs.
non-smoking.

;<details>
<summary>**Click for the solution**</summary>

```{r}
#| label: 4B

m4b <- glm(deaths ~ smoking,
                     family = poisson(link = "log"),
                     data = doctors)

m4b |> parameters::parameters(exponentiate=TRUE)
```

<!-- panache-ignore-start-->
Parameter        |   IRR |   SE |         95% CI |     z |      p
-----------------------------------------------------------------
(Intercept)      | 20.20 | 2.01 | [16.51, 24.40] | 30.21 | < .001
smoking [smoker] |  6.24 | 0.67 | [ 5.08,  7.74] | 17.08 | < .001
<!-- panache-ignore-end-->

</details>

**Exercise 5B**

Let's try to correct for these issues.

Provide the scientific regression equation we will run

<details>
<summary>**Click for the solution**</summary>

The model will be:

$$
  deaths = Poisson(λ) \\
  log(λ) = β0 + β1_{smoking} + β2_{age} + \text{offset}(personyears)
$$

*Note that the offset does not have a β coefficient associated with it!*

</details>

**Exercise 6B**

The following R code contains two mistakes. Identify and fix them:

```r
m5b <- glm(deaths ~ smoking + age,
                     family = poisson(link = "logit"),
                     offset = personyears,
                     data = doctors)
```

<details>
<summary>**Click for the solution**</summary>

```{r}
#| label: 6b

# Mistake 1: the offset should be log-personyears!
# Mistake 2: the link should be log (logit is for logistic regression)!
m6b <- glm(deaths ~ smoking + age,
                     family = poisson(link = "log"),
                     offset = log(personyears),
                     data = doctors)
```

</details>

### C: Writing a Clinical Report

**Exercise 3A:**

The final model is provided in **Exercise 6B**. If you didn't make that
exercise, go ahead and see the solution and copy the code. You will not make
inference based on the model.

You will write a clincal report based on the inference we got from the model.

> Structuring the Report

- 1: How would you explain the incidence rate ratio (IRR) to a clinician who is
  not familiar with statistics?
- 2: Write a 1-paragraph summary of the findings from your Poisson model for a
  clinical audience.
- 3: Compare these results to the naïve model from **Exercise 4B**
- 4: How would you visually present the effect of smoking on mortality in a
  figure for the report? Describe the plot.
- 5: Now implement the plot using {ggplot2}.
- 6: Make a legend to the plot so that others can read it with the intention you
  had.

No solution is given here, as you can be creative 😄

Please discuss your ideas with the instructors!
